---
title: "Retinal Microvasculature, HIV, and Cardiovascular Risk: Analysis Code"
format: 
  html: 
    embed-resources: true
    page-layout: full
    toc: true 
    toc-depth: 2
    theme: journal
    fontsize: 1.1em
    linestretch: 1.7
editor: source 
editor_options:
  chunk_output_type: console
execute:
  echo: true
  warning: false
  error: false
knitr:
  opts_chunk:
    decimal.mark: "."
    digits: 3
---

## About this document

Analysis code for my BSc Honours project in Bioinformatics and Computational
Biology (Stellenbosch University, 2024). It examines retinal microvascular
measures in relation to HIV, antiretroviral therapy (ART) and cardiovascular
disease (CVD) risk factors in adults from the EndoAfrica study.

**The data are not included.** They are confidential participant data,
collected under ethics approval and managed by the EndoAfrica study team. To
render this document you need two files in a `data/` folder:

- `data/endoafrica_retinal_data.xlsx`: the study spreadsheet
- `data/variable_names_edited.txt`: a tab-delimited mapping from raw to clean
  column names. It is a hand-edited copy of `data/variable_names_raw.csv`
  (which this document writes), with the `new` column filled in.

Rendering also writes figure files (`.svg` and `.png`) to the project folder.
The random forest and decision tree sections at the end are exploratory and
were not part of the final report.

## Overview

Packages and the shared plot theme.

::: panel-tabset

### Packages

```{r}
#| label: packages
library(broom)
library(caret)
library(circlize)
library(cowplot)
library(ComplexHeatmap)
library(dplyr)
library(DT)
library(ggplot2)
library(janitor)
library(lubridate)
library(PCAtools)
library(purrr)
library(randomForest)
library(rpart)
library(rpart.plot)
library(stringi)
library(stringr)
library(tibble)
library(tidyr)
library(kableExtra)

# Also required, called with :: below: readr, readxl
```

### Plot theme and colours

```{r}
#| label: plot-theme
color_map_list <- c(
  "Female" = "#C52233",
  "Male" = "#2A7221",
  "No" = "#C52233",
  "Yes" = "#2A7221",
  "NA" = "#D0EAF3",
  ">5 years ago" = "#B1D4E7",
  ">15 years ago" = "#8FB0CE",
  "2 - 5 years ago" = "#6B8CC9",
  "In last year" = "#2F5ABF",
  "2nd" = "#D0EAF3",
  "1st" = "#2F5ABF",
  "Unknown" = "#D0EAF3",
  "1-5 years" = "#B1D4E7",
  "< 12 months" = "#8FB0CE",
  "6-15 years" = "#6B8CC9",
  "Yes currently" = "#2F5ABF",
  "In the past" = "#8FB0CE",
  "Never smoked" = "#D0EAF3",
  "Type II Diabetes (also known as Non-insulin Dependent Diabetes)" = "#D0EAF3",
  "Don't know" = "#2F5ABF",
  "Viral load > 1000" = "#2A7221",
  "Viral load < 1000" = "#C52233",
  "Less than a year" = "#D0EAF3",
  "1-5 years" = "#8FB0CE",
  "More than 5 years" = "#2F5ABF"
)

# Light theme: white background, black text and lines, light grey grid
custom_theme <- theme_light() +
  theme(
    panel.background = element_rect(fill = "white"),
    axis.text = element_text(color = "black"),
    axis.title = element_text(color = "black"),
    axis.line = element_line(color = "black"),
    legend.background = element_rect(fill = "white"),
    legend.text = element_text(color = "black"),
    legend.title = element_text(color = "black"),
    plot.title = element_text(size = 14, face = "bold", color = "black"),
    plot.subtitle = element_text(size = 12, color = "black"),
    plot.caption = element_text(size = 10, color = "black"),
    panel.grid.major = element_line(color = "#F0EDEE", size = 0.5),
    panel.grid.minor = element_line(color = "#F0EDEE", size = 0.25),
    strip.text = element_text(size = 12, color = "black"),
    strip.background = element_rect(fill = "white")
  )

theme_set(custom_theme)
```

:::

## Data preparation

Reads the EndoAfrica spreadsheet, cleans variable names and types, and builds the grouping variables and variable lists used in later sections.

::: panel-tabset

### Load and clean the EndoAfrica data

```{r}
#| label: load-clean-data
path_dta <- "data/endoafrica_retinal_data.xlsx"

dta <- readxl::read_xlsx(path_dta)

# The sheet repeats its header row partway down (cases and controls are
# stacked), so drop that row and any rows without a study ID
dta <- dta |>
  filter(!str_detect(`Study ID`, "Study ID") &
    !is.na(`Study ID`))

# Clean variable names ----

# Names were mapped by hand in Excel: the raw and cleaned names are written
# out, and the hand-edited mapping is read back in
dta_names_clean <- tibble(
  orig = names(dta),
  new = NA
)

tmp <- clean_names(dta)

dta <- clean_names(dta)

tmp <- names(dta)

dta_names_clean$interim <- tmp

readr::write_csv(dta_names_clean, "data/variable_names_raw.csv")

# Hand-edited mapping of raw to clean names (not included in the repository)
tmp <- read.delim("data/variable_names_edited.txt")

# Where no new name was given, keep the janitor-cleaned name
tmp <- tmp |>
  mutate(new = if_else(is.na(new) | new == "", interim, new))

names(dta) <- tmp$new

# make NA true NA ----

for (i in names(dta)) {
  dta[[i]][is.na(dta[[i]])] <- NA
}

# remove new crve/crae/ratio
dta <- dta |>
  select(-c(new_crae, new_crve, new_crae_crve_ratio))

# Interview dates: convert Excel serial numbers to date-times
dta <- dta |>
  mutate(interview_start_time = as.numeric(interview_start_time))

dta <- dta |>
  mutate(interview_start_time = as.POSIXct(
    (interview_start_time - 25569) * 86400,
    origin = "1970-01-01",
    tz = "UTC"
  ))

# Group recruitment sites (used in the bimodal-distribution plots)
dta <- dta |>
  mutate(
    site_group = case_when(
      recruitment_site %in% c("Uitsig", "Elsies", "Adriaanse", "Ravensmead") ~
        "Cape Town Clinics",
      recruitment_site == "Worcester" ~ "Worcester",
      recruitment_site %in% c("Fisantekraal", "Durbanville") ~
        "Northern Cape Town"
    )
  )

# Interview month and season
dta <- dta |>
  mutate(
    interview_start_time = as.POSIXct(
      interview_start_time,
      origin = "1970-01-01",
      tz = "UTC"
    ),
    month = month(interview_start_time),
    season = case_when(
      month %in% c(5, 6, 7, 8) ~ "Warm",
      month %in% c(9, 10, 11, 12, 1, 2, 3, 4) ~ "Cold",
      TRUE ~ "Unknown" # Just in case there's an error
    )
  )

# create cvd risk factor category
dta <- dta |>
  rowwise() |>
  mutate(cvd_risk_factor = if_else(
    any(
      c(hypertension, heart_disease, high_cholesterol, tb, diabetes) == "Yes"
    ),
    "Yes",
    "No"
  )) |>
  ungroup()

# category for smoker
dta <- dta |>
  mutate(smoker_new = case_when(
    smoker %in% c("Never smoked") ~ "No",
    smoker %in% c("Yes currently", "In the past") ~ "Yes",
    TRUE ~ smoker
  ))

# Find numeric and categorical variables ----

# which vars contain numbers?
x <- map_dbl(dta, ~ sum(str_detect(.x, "[0-9]+"), na.rm = TRUE))

# of those, which ones also contain letters?
xx <- which(x != 0)
xxx <- map_dbl(dta[, xx], ~ sum(str_detect(.x, "[a-zA-Z]"), na.rm = TRUE))

# put together those with no numbers and those with letters and numbers
cats <- c(names(x)[which(x == 0)], names(xxx)[which(xxx != 0)])

map(dta[, cats], ~ unique(.x))

# acr_ratio and ldl should be numeric
cats <- cats[!cats %in% c("acr_ratio", "ldl")]

map(dta[, cats], ~ summary(.x[!is.na(.x)]))

# numeric

# remove x1 column, irrelevant
dta <- dta |>
  select(-x1)

nums <- names(dta)[which(!names(dta) %in% cats)]

map(dta[, nums], ~ summary(as.numeric(.x[!is.na(.x)])))

# tortuosity_index is not rounded: at one decimal every value becomes 0.9

# Clean numeric variables ----

# check for letters or symbols in vars
x <- dta |>
  select(all_of(nums)) |>
  map_dbl(~ sum(str_detect(.x, "^[^0-9]+$"), na.rm = TRUE))

x <- x[x != 0]

dta |>
  select(all_of(names(x))) |>
  distinct() |>
  print(n = Inf)

# fix - make all numeric vars numeric

dta <- dta |>
  mutate(across(all_of(nums), ~ as.numeric(.x)))

dta |>
  select(all_of(nums)) |>
  summary()

# Implausible values: ART durations over 4000 weeks become NA, and heights
# under 3 (recorded in metres) are converted to cm
thresholds <- list(
  art_duration_weeks = 4000,
  height_threshold = 3
)

dta <- dta |>
  mutate(
    art_duration_weeks = if_else(
      art_duration_weeks > thresholds$art_duration_weeks,
      NA_real_,
      art_duration_weeks
    ),
    height = if_else(height < thresholds$height_threshold, height * 100, height)
  )

# Print summary of numeric columns
dta |>
  select(where(is.numeric)) |>
  summary()

# Clean character variables ----

dta |>
  select(all_of(cats)) |>
  map(~ unique(.x))

# make all long answers shorter in family history
columns_to_modify <- c(
  "father_diabetes", "father_heart_disease", "father_high_cholesterol",
  "father_hypertension", "father_stroke", "mother_diabetes",
  "mother_heart_disease", "mother_high_cholesterol", "mother_hypertension",
  "mother_stroke", "alcohol_frequency", "physically_active"
)

modifications <- list(
  "Yes, but I don't know the age" = "Yes",
  "Don't know" = "Unknown",
  "Yes, over the age of 60" = "Yes",
  "Yes, under the age of 60" = "Yes",
  "Yes, but I don't know type" = "Yes",
  "8 or more days a month" = "Yes",
  "Less than 8 days a month" = "Yes",
  "Daily" = "Yes"
)

for (pattern in names(modifications)) {
  dta <- dta |>
    mutate(
      across(
        all_of(columns_to_modify[which(columns_to_modify %in% colnames(dta))]),
        ~ case_when(
          . == pattern ~ modifications[[pattern]],
          TRUE ~ .
        )
      )
    )
}

# Check the recoded answers
dta |>
  select(all_of(cats)) |>
  map(~ unique(.x))

# Log10-transform skewed variables
dta <- dta |>
  mutate(
    log_crp = log10(crp),
    log_ggt = log10(ggt),
    log_viral_load = log10(viral_load),
    log_acr_ratio = log10(acr_ratio),
    log_microalbuminuria = log10(microalbuminuria),
    log_ten_year_cvd_risk = log10(ten_year_cvd_risk)
  )

dta <- dta |>
  select(
    everything(),
    crp, log_crp,
    ggt, log_ggt,
    viral_load, log_viral_load,
    acr_ratio, log_acr_ratio,
    microalbuminuria, log_microalbuminuria,
    ten_year_cvd_risk, log_ten_year_cvd_risk,
  )

nums <- names(dta)[which(!names(dta) %in% cats)]

nums <- names(dta)[!names(dta) %in% c(cats, "interview_start_time")]

# Grouping variables for age, ART duration, viral load and CD4 count ----

dta <- dta |>
  mutate(age_group = cut(age,
    breaks = c(18, 29, 39, 49, 59, 69, 79),
    labels = c("18-29", "30-39", "40-49", "50-59", "60-69", "70-79"),
    right = TRUE
  ))

dta <- dta |>
  mutate(art_duration_group = cut(art_duration_weeks,
    breaks = c(0, 52, 260, 1000),
    labels = c("Less than a year", "1-5 years", "More than 5 years"),
    right = TRUE
  ))

dta <- dta |>
  mutate(viral_load_group = cut(viral_load,
    breaks = c(0, 1000, 1000000),
    labels = c("Viral load < 1000", "Viral load > 1000"),
    right = TRUE
  ))

dta <- dta |>
  mutate(cd_count_clinic_group = cut(cd4_count_clinic,
    breaks = c(0, 200, 500, 100000),
    labels = c("0-199", "200-499", "500+"),
    right = TRUE
  ))

dta <- dta |>
  mutate(cd_count_clinic_group_2 = cut(cd4_count_clinic,
    breaks = c(0, 500, 100000),
    labels = c("0-499", "500+"),
    right = TRUE
  ))

dta <- dta |>
  mutate(cd4_count_endo_group = cut(endo_result,
    breaks = c(0, 200, 500, 100000),
    labels = c("0-199", "200-499", "500+"),
    right = TRUE
  ))

dta <- dta |>
  mutate(cd4_count_endo_group_2 = cut(endo_result,
    breaks = c(0, 500, 100000),
    labels = c("0-499", "500+"),
    right = TRUE
  ))

# ART regimen names (free text) mapped to medication class
medication_classes <- c(
  "NA" = "HIV negative",
  "Aluvia Lamivudine Tenofovir" = "FDC",
  "Odimune" = "FDC",
  "Aluvia Tenofovir Lamuvidine Lopinavir ritonavir ciproflatoca" = "FDC",
  "Odimune 1 daily (EFV, 3tc and Tenofovir)" = "FDC",
  "Trivenz (ENT200+TEN300+EFA600) One daily" = "FDC",
  "Trivenz (efavirenz, 3tc and TDF)" = "FDC",
  "Aspen Nevirapine and Adco Lamivudine" = "FDC",
  "Trivenz( EFV, Tenofovir, 3TC)" = "FDC",
  "Tribuss( EFV, 3TC and TDF )" = "FDC",
  "Odimune( efavirenz, 3tc and tenofovir )" = "FDC",
  "Tribuss (EMT 200 +TEN 300 +EFA 600)" = "FDC",
  "Tribuss ( EMT 200 + Ten 300 + EFA 600 )" = "FDC",
  "Tribuss(EMT 200,TEN 300, EFR600)" = "FDC",
  "Tribuss ( ENT 200, TEN 300, EFA 600)" = "FDC",
  "Aluvia 200/50mg; Lamivudine/Zidovudine 150/300" = "FDC",
  "Tribuss (Tenofovir/Emtricitabine /EFAV)" = "FDC",
  "Tribuss / Odiomune" = "FDC",
  "Tribuss" = "FDC",
  "Tenovir/Emitricitabine/Efavirenz" = "FDC",
  "TDF,FTC,EFV" = "FDC",
  "Odimune (Tenovir300/Emtrcitabine200/EFAV600" = "FDC",
  "Atroiza" = "NNRTI",
  "Lamivudine 300mg / Tenofovir 300mg / Nevirapine 200mg" = "FDC",
  "Atenef" = "NNRTI",
  "Odimune  ( EMT  200, TEN300, EFA600)" = "FDC",
  "Tenovir300 / Emitricitabine200 / Efavirenz600" = "FDC",
  "TDC" = "HIV negative",
  "odimune" = "FDC",
  "Odimune + Aluvia" = "FDC",
  "TDF / 3TC/NVP" = "FDC",
  "TDF/3TC/Aluvia" = "FDC",
  "Tenofovir / EMT / EFAV" = "FDC",
  "Tenofovir 300/ Emtricitabine 200/ EFAV 600" = "FDC",
  "Lamzid/Aluvia" = "FDC",
  "FDC" = "FDC",
  "TDF/FTC/EFU" = "FDC",
  "TDF/FTC/EFV" = "FDC",
  "TDF/3TC/NVP" = "FDC",
  "Odimune 1 at night" = "FDC",
  "ABC/3TC/Efv" = "FDC",
  "Lamid/Aluvia" = "FDC",
  "FDC Lamid/Efv" = "FDC",
  "Aluvia, Tenemine" = "FDC",
  "TDF/FTC/EFV (Odimune)" = "FDC",
  "TDF/FTC,NVP" = "FDC",
  "Tenofovir 300mg/ Emtriatabine 200mg/ Efavirenz 600mg" = "FDC",
  "FDC, Atroiza" = "FDC",
  "TDF / FTC / EFV" = "FDC",
  "FDC (TDF/FTC/EFv)" = "FDC",
  "Aluvia, Lomzid, Lamivudine/zidovudine, lopinavir/ritonavir" = "FDC",
  "TDF300mg/FTC200mg/EFV600mg" = "FDC",
  "FDC (TDF / FTC) NVP" = "FDC",
  "FDC (TDF / FTC / EFV)" = "FDC",
  "TDF / 3TC and Aluvia" = "FDC",
  "FDC (TDF FTC EFV)" = "FDC",
  "FDC (TDF/FTC/EFV)" = "FDC",
  "FDC (TDF/EMT/EFV)" = "FDC",
  "FDC (Ten/EMT/Efv)" = "FDC",
  "FDC (Tdf/ Emt/ Efv)" = "FDC",
  "TDF (3TC/EFV/FTC)" = "FDC",
  "TDF (FTC/EFV/TDF)" = "FDC",
  "FDC/FTC/TDF aluvia" = "FDC",
  "FDC (Ten/Emt/Efv)" = "FDC",
  "FDC (TDF/TC/EFV)" = "FDC",
  "TDF/FTC/Aluvia" = "FDC",
  "FDC (TDF/TTC/EFV)" = "FDC",
  "TDP/3TC/NVP" = "FDC",
  "TDF/FTC/Efv" = "FDC",
  "Lamzia and ALluvia" = "FDC",
  "FDC (Tnf/Efv/Emt)" = "FDC",
  "FDC (Tdf/Ftc/Efv)" = "FDC",
  "FDC (Tnf/Efv/Emt" = "FDC",
  "Fen/Emtr/Evf" = "FDC"
)

dta <- dta |>
  mutate(
    medication_class = case_when(
      is.na(art_name) & hiv_pos == "No" ~ "HIV negative",
      is.na(art_name) ~ NA_character_,
      art_name %in% names(medication_classes) ~ medication_classes[art_name],
      TRUE ~ "Unknown"
    )
  )

new_vars <- c(
  "medication_class", "art_duration_group", "age_group", "art_duration_group",
  "hiv_duration", "cd_count_clinic_group_2", "cd_count_clinic_group",
  "cd4_count_endo_group", "cd4_count_endo_group_2", "viral_load_group"
)

cats <- c(cats, new_vars)

dta <- dta |>
  mutate(art_duration_group_2 = cut(art_duration_weeks,
    breaks = c(0, 260, 1000),
    labels = c("0-5 years", "More than 5 years"),
    right = TRUE
  ))

new_vars <- c("viral_load", "art_duration_weeks", "endo_result")

nums <- c(nums, new_vars)

dta <- dta |>
  mutate(across(all_of(cats), as.character))

dta <- dta |>
  mutate(across(all_of(nums), as.numeric))
```

### Variable groups used in later sections

```{r}
#| label: variable-groups
# Clinical and retinal variables combined
all_var <- c(
  "total_cholesterol", "hdl", "ldl", "triglycerides", "fasting_glucose",
  "hba1c", "crp", "ggt", "haemoglobin", "creatinine", "mdrd_egfr",
  "urine_creatinine", "acr_ratio", "microalbuminuria", "cd4_count_clinic",
  "endo_result", "ckd_epi_egfr", "ten_year_cvd_risk", "vascular_age",
  "diameter_baseline", "fmd_percentage", "average_carotid_imt", "age", "bmi",
  "waist_hip_ratio", "mean_systolic_pressure", "mean_diastolic_pressure",
  "mean_heart_rate", "crae", "crve", "crae_crve_ratio", "segment_tree_area",
  "total_length_skeletonised_tree", "segmented_fractal_analysis",
  "skeleton_fractal_analysis", "fourier_fractal_enhanced",
  "lacunarity_segmented", "lacunarity_skeleton", "branch_points", "endpoints",
  "average_arterial_tree_diameter", "retinal_area", "arterial_tree_segments",
  "tortuosity_index", "branch_segment_tortuosity_average",
  "branch_segment_tortuosity_skewness", "branch_segment_tortuosity_kurtosis",
  "branch_segment_length_average", "branch_segment_length_skewness",
  "branch_segment_length_kurtosis", "arteriole_branch_parent_width",
  "arteriole_branch_smaller_branch", "arteriole_branch_larger_branch",
  "arteriole_branch_angle", "arteriole_branch_optimality",
  "arteriole_mean_branching_coefficient", "arteriole_mean_asymmetry_ratio",
  "arteriole_mean_angular_asymmetry", "venule_branch_parent_width",
  "venule_branch_smaller_branch", "venule_branch_larger_branch",
  "venule_branch_angle", "venule_branch_optimality",
  "venule_mean_branching_coefficient", "venule_mean_asymmetry_ratio",
  "venule_mean_angular_asymmetry"
)

# Drop the retinal measures with bimodal distributions (see bimodal_var)
x <- all_var

xx <- which(x %in% c(
  "fourier_fractal_enhanced", "branch_segment_length_kurtosis",
  "segment_tree_area", "total_length_skeletonised_tree", "branch_points",
  "endpoints", "fourier_fractal_analysis", "lacunarity_segmented",
  "lacunarity_skeleton", "retinal_area", "branch_segment_length_skewness",
  "branch_segment_tortuosity_average", "tortuosity_index",
  "arteriole_mean_branching_coefficient"
))

all_var <- x[-xx]

# Clinical variables
clinical_var <- c(
  "total_cholesterol", "hdl", "ldl", "triglycerides", "fasting_glucose",
  "hba1c", "crp", "ggt", "haemoglobin", "cd4_count_clinic", "endo_result",
  "viral_load", "creatinine", "mdrd_egfr", "urine_creatinine", "acr_ratio",
  "microalbuminuria", "ckd_epi_egfr", "ten_year_cvd_risk", "vascular_age",
  "diameter_baseline", "fmd_percentage", "average_carotid_imt", "age", "bmi",
  "waist_hip_ratio", "mean_systolic_pressure", "mean_diastolic_pressure",
  "mean_heart_rate", "log_crp", "log_ggt", "log_viral_load", "log_acr_ratio",
  "log_microalbuminuria", "log_ten_year_cvd_risk"
)

clinical_var_pca <- c(
  "total_cholesterol", "hdl", "ldl", "triglycerides", "fasting_glucose",
  "hba1c", "crp", "ggt", "haemoglobin", "creatinine", "mdrd_egfr",
  "urine_creatinine", "acr_ratio", "microalbuminuria", "ckd_epi_egfr",
  "ten_year_cvd_risk", "vascular_age", "diameter_baseline", "fmd_percentage",
  "average_carotid_imt", "age", "bmi", "waist_hip_ratio",
  "mean_systolic_pressure", "mean_diastolic_pressure", "mean_heart_rate"
)

clinical_var_art <- c(
  "total_cholesterol", "hdl", "ldl", "triglycerides", "fasting_glucose",
  "hba1c", "crp", "ggt", "haemoglobin", "cd4_count_clinic", "endo_result",
  "viral_load", "creatinine", "mdrd_egfr", "urine_creatinine", "acr_ratio",
  "microalbuminuria", "ckd_epi_egfr", "ten_year_cvd_risk", "vascular_age",
  "diameter_baseline", "fmd_percentage", "average_carotid_imt", "age", "bmi",
  "waist_hip_ratio", "mean_systolic_pressure", "mean_diastolic_pressure",
  "mean_heart_rate"
)

# Retinal measures
retinal_var <- c(
  "crae", "crve", "crae_crve_ratio", "segment_tree_area",
  "total_length_skeletonised_tree", "segmented_fractal_analysis",
  "skeleton_fractal_analysis", "fourier_fractal_enhanced",
  "lacunarity_segmented", "lacunarity_skeleton", "branch_points", "endpoints",
  "average_arterial_tree_diameter", "retinal_area", "arterial_tree_segments",
  "tortuosity_index", "branch_segment_tortuosity_average",
  "branch_segment_tortuosity_skewness", "branch_segment_tortuosity_kurtosis",
  "branch_segment_length_average", "branch_segment_length_skewness",
  "branch_segment_length_kurtosis", "arteriole_branch_parent_width",
  "arteriole_branch_smaller_branch", "arteriole_branch_larger_branch",
  "arteriole_branch_angle", "arteriole_branch_optimality",
  "arteriole_mean_branching_coefficient", "arteriole_mean_asymmetry_ratio",
  "arteriole_mean_angular_asymmetry", "venule_branch_parent_width",
  "venule_branch_smaller_branch", "venule_branch_larger_branch",
  "venule_branch_angle", "venule_branch_optimality",
  "venule_mean_branching_coefficient", "venule_mean_asymmetry_ratio",
  "venule_mean_angular_asymmetry"
)

# Drop the retinal measures with bimodal distributions (see bimodal_var)
x <- retinal_var

xx <- which(x %in% c(
  "branch_segment_length_kurtosis", "segment_tree_area",
  "total_length_skeletonised_tree", "branch_points", "endpoints",
  "fourier_fractal_analysis", "lacunarity_segmented", "lacunarity_skeleton",
  "retinal_area", "branch_segment_length_skewness",
  "branch_segment_tortuosity_average", "tortuosity_index",
  "arteriole_mean_branching_coefficient"
))

retinal_var <- x[-xx]

# Retinal measures in small groups, for plotting
retinal_var_1 <- c("crae", "crve", "crae_crve_ratio")

retinal_var_2 <- c(
  "segmented_fractal_analysis", "skeleton_fractal_analysis",
  "average_arterial_tree_diameter"
)

retinal_var_3 <- c(
  "arterial_tree_segments", "branch_segment_tortuosity_skewness",
  "branch_segment_tortuosity_kurtosis"
)

retinal_var_4 <- c(
  "branch_segment_length_average", "arteriole_branch_parent_width",
  "arteriole_branch_smaller_branch"
)

retinal_var_5 <- c(
  "arteriole_branch_larger_branch", "arteriole_branch_angle",
  "arteriole_branch_optimality"
)

retinal_var_6 <- c(
  "arteriole_mean_asymmetry_ratio", "arteriole_mean_angular_asymmetry",
  "venule_branch_parent_width"
)

retinal_var_7 <- c(
  "venule_branch_smaller_branch", "venule_branch_larger_branch",
  "venule_branch_angle"
)

retinal_var_8 <- c(
  "venule_branch_optimality", "venule_mean_branching_coefficient"
)

retinal_var_9 <- c(
  "venule_mean_asymmetry_ratio", "venule_mean_angular_asymmetry"
)

# Retinal measures with bimodal distributions
bimodal_var_1 <- c(
  "fourier_fractal_enhanced", "branch_segment_length_kurtosis",
  "segment_tree_area"
)

bimodal_var_2 <- c(
  "total_length_skeletonised_tree", "branch_points", "endpoints"
)

bimodal_var_3 <- c(
  "lacunarity_segmented", "lacunarity_skeleton", "retinal_area"
)

bimodal_var_4 <- c(
  "branch_segment_length_skewness", "branch_segment_tortuosity_average"
)

bimodal_var_5 <- c("tortuosity_index", "arteriole_mean_branching_coefficient")

bimodal_var <- c(
  "fourier_fractal_enhanced", "branch_segment_length_kurtosis",
  "segment_tree_area", "total_length_skeletonised_tree", "branch_points",
  "endpoints", "lacunarity_segmented", "lacunarity_skeleton", "retinal_area",
  "branch_segment_length_skewness", "branch_segment_tortuosity_average",
  "tortuosity_index", "arteriole_mean_branching_coefficient"
)

# Variable groups for tables and plots
character_values <- c(
  "age_group", "cvd_risk_factor", "on_art", "smoker", "clarity",
  "recruitment_site", "past_heart_attack", "past_cancer", "past_tb",
  "past_stroke", "hypertension", "heart_disease", "high_cholesterol", "tb",
  "diabetes", "diabetes_type", "diabetes_duration", "father_diabetes",
  "father_heart_disease", "father_high_cholesterol", "father_stroke",
  "father_hypertension", "mother_diabetes", "mother_heart_disease",
  "mother_high_cholesterol", "mother_hypertension", "mother_stroke", "hiv_pos",
  "gender", "hiv_related", "art_line", "medication_class", "art_duration_group",
  "age_group", "art_duration_group", "hiv_duration", "cd_count_clinic_group_2",
  "cd_count_clinic_group", "cd4_count_endo_group", "cd4_count_endo_group_2",
  "viral_load_group"
)

log_values <- c(
  "log_crp", "log_ggt", "log_acr_ratio", "log_microalbuminuria",
  "log_ten_year_cvd_risk"
)

log_values_hiv <- c(
  "log_crp", "log_ggt", "log_viral_load", "log_acr_ratio",
  "log_microalbuminuria", "log_ten_year_cvd_risk"
)

past_medical_history <- c(
  "past_heart_attack", "past_cancer", "past_tb", "past_stroke"
)

medical_condition <- c(
  "hypertension", "heart_disease", "high_cholesterol", "tb"
)

diabetes_lev <- c("diabetes", "diabetes_type", "diabetes_duration")

family_history_1 <- c(
  "father_diabetes", "father_heart_disease", "father_high_cholesterol",
  "father_stroke", "father_hypertension"
)

family_history_2 <- c(
  "mother_diabetes", "mother_heart_disease", "mother_high_cholesterol",
  "mother_hypertension", "mother_stroke"
)

clarity <- c("clarity")

gender <- c("gender")

hiv_pos <- c("hiv_pos")

hiv_related <- c("hiv_duration", "on_art", "art_line")

art_name <- c("art_name")

art_duration <- ("art_duration_weeks")

health_dta <- c("age", "bmi", "waist_hip_ratio")

mean_heart_measurements <- c(
  "mean_systolic_pressure", "mean_diastolic_pressure", "mean_heart_rate"
)

lifestyle_factors <- c("smoker", "alcohol_frequency", "physically_active")

clinical_dta_1 <- c(
  "total_cholesterol", "hdl", "ldl", "triglycerides", "fasting_glucose",
  "hba1c", "crp", "ggt", "haemoglobin"
)

clinical_dta_2 <- c("cd4_count_clinic", "cd4_count_endo_africa ", "viral_load")

clinical_dta_3 <- c(
  "creatinine", "mdrd_egfr", "urine_creatinine", "acr_ratio",
  "microalbuminuria", "ckd_epi_egfr"
)

clinical_dta_4 <- c(
  "10_year_cvd_risk", "vascular_age", "diameter_baseline", "fmd_percentage",
  "average_carotid_imt"
)

clinical_dta_10 <- c("total_cholesterol", "hdl", "ldl")

clinical_dta_11 <- c("triglycerides", "fasting_glucose", "hba1c")

clinical_dta_12 <- c("crp", "ggt", "haemoglobin")

clinical_dta_13 <- c("creatinine", "mdrd_egfr", "urine_creatinine")

clinical_dta_14 <- c("acr_ratio", "microalbuminuria", "ckd_epi_egfr")

clinical_dta_15 <- c("10_year_cvd_risk", "vascular_age", "diameter_baseline")

clinical_dta_16 <- c("fmd_percentage", "average_carotid_imt")

clinical_dta_17 <- c("cd4_count_clinic", "cd4_count_endo_africa ", "viral_load")

parameters_1 <- c(
  "crae", "crve", "crae_crve_ratio", "new_crae", "new_crve",
  "new_crae_crve_ratio", "segment_tree_area", "total_length_skeletonised_tree",
  "segmented_fractal_analysis"
)

parameters_2 <- c(
  "skeleton_fractal_analysis", "fourier_fractal_enhanced",
  "lacunarity_segmented", "lacunarity_skeleton", "branch_points", "endpoints",
  "average_arterial_tree_diameter", "retinal_area", "arterial_tree_segments"
)

parameters_3 <- c(
  "tortuosity_index", "branch_segment_tortuosity_average",
  "branch_segment_tortuosity_skewness", "branch_segment_tortuosity_kurtosis",
  "branch_segment_length_average", "branch_segment_length_skewness",
  "branch_segment_length_kurtosis", "arteriole_branch_parent_width",
  "arteriole_branch_smaller_branch"
)

parameters_4 <- c(
  "arteriole_branch_larger_branch", "arteriole_branch_angle",
  "arteriole_branch_optimality", "arteriole_mean_branching_coefficient",
  "arteriole_mean_asymmetry_ratio", "arteriole_mean_angular_asymmetry",
  "venule_branch_parent_width", "venule_branch_smaller_branch",
  "venule_branch_larger_branch"
)

parameters_5 <- c(
  "venule_branch_larger_branch", "venule_branch_angle",
  "venule_branch_optimality", "venule_mean_branching_coefficient",
  "venule_mean_asymmetry_ratio", "venule_mean_angular_asymmetry"
)

cvd_risk_factor <- c("cvd_risk_factor")

parameters_10 <- c("crae", "crve", "crae_crve_ratio")

parameters_11 <- c(
  "segment_tree_area", "total_length_skeletonised_tree",
  "segmented_fractal_analysis"
)

parameters_12 <- c(
  "skeleton_fractal_analysis", "fourier_fractal_enhanced",
  "lacunarity_segmented"
)

parameters_13 <- c("lacunarity_skeleton", "branch_points", "endpoints")

parameters_14 <- c(
  "average_arterial_tree_diameter", "retinal_area", "arterial_tree_segments"
)

parameters_15 <- c(
  "tortuosity_index", "branch_segment_tortuosity_average",
  "branch_segment_tortuosity_skewness"
)

parameters_16 <- c(
  "branch_segment_tortuosity_kurtosis", "branch_segment_length_average",
  "branch_segment_length_skewness"
)

parameters_17 <- c(
  "branch_segment_length_kurtosis", "arteriole_branch_parent_width",
  "arteriole_branch_smaller_branch"
)

parameters_18 <- c(
  "arteriole_branch_larger_branch", "arteriole_branch_angle",
  "arteriole_branch_optimality"
)

parameters_19 <- c(
  "arteriole_mean_branching_coefficient", "arteriole_mean_asymmetry_ratio",
  "arteriole_mean_angular_asymmetry"
)

parameters_20 <- c(
  "venule_branch_parent_width", "venule_branch_smaller_branch",
  "venule_branch_larger_branch"
)

parameters_21 <- c(
  "venule_branch_larger_branch", "venule_branch_angle",
  "venule_branch_optimality"
)

parameters_22 <- c(
  "venule_mean_branching_coefficient", "venule_mean_asymmetry_ratio",
  "venule_mean_angular_asymmetry"
)
```

:::

## Summary statistics

Descriptive statistics for the whole sample and by HIV and ART status.

::: panel-tabset

### Whole sample: numeric

```{r}
#| label: summary-all-numeric
# Summary statistics for each clinical variable
dta_nums <- dta |>
  select(study_id, event, hiv_pos, all_of(nums))

dta_nums <- dta_nums |>
  pivot_longer(
    cols = -c(event, study_id, hiv_pos),
    names_to = "key",
    values_to = "value"
  )

summary_stats_clinical <- list()

for (var in clinical_var) {
  summary_statistics <- dta_nums |>
    filter(key == var) |>
    summarise(
      mean = mean(value, na.rm = TRUE),
      std_dev = sd(value, na.rm = TRUE),
      `1st_quartile` = quantile(value, 0.25, na.rm = TRUE),
      median = median(value, na.rm = TRUE),
      `3rd_quartile` = quantile(value, 0.75, na.rm = TRUE),
      min = min(value, na.rm = TRUE),
      max = max(value, na.rm = TRUE),
      .groups = "drop"
    )
  summary_stats_clinical[[var]] <- summary_statistics
}

summary_stats_df_numeric <- bind_rows(summary_stats_clinical, .id = "variable")

datatable(summary_stats_df_numeric) |>
  formatStyle(
    columns = c(
      "mean", "variable", "std_dev", "1st_quartile", "median", "3rd_quartile",
      "min", "max"
    ),
    fontSize = "11px",
    color = "black",
    backgroundColor = "white"
  ) |>
  formatRound(
    columns = c(
      "mean", "variable", "std_dev", "1st_quartile", "median", "3rd_quartile",
      "min", "max"
    ),
    digits = 4
  )
```

### Whole sample: categorical

```{r}
#| label: summary-all-categorical
# Percentage of participants in each category
dta_cats <- dta |>
  select(study_id, event, all_of(cats)) |>
  pivot_longer(
    cols = -c(event, study_id),
    names_to = "key",
    values_to = "value"
  )

summary_stats_categorical <- list()

for (var in character_values) {
  summary_statistics <- dta_cats |>
    filter(key == var) |>
    group_by(key, value) |>
    summarise(count = n(), .groups = "drop") |>
    group_by(key) |>
    mutate(percent = (count / sum(count)) * 100) |>
    select(key, outcome = value, percent)

  summary_stats_categorical[[var]] <- summary_statistics
}

summary_stats_df_categorical <- bind_rows(
  summary_stats_categorical,
  .id = "key"
)

datatable(summary_stats_df_categorical) |>
  formatStyle(
    columns = c("key", "outcome", "percent"),
    fontSize = "11px",
    color = "black",
    backgroundColor = "white"
  )
```

### Whole sample: retinal

```{r}
#| label: summary-all-retinal
# Summary statistics for each retinal measure
dta_nums <- dta |>
  select(study_id, event, hiv_pos, all_of(retinal_var))

dta_nums <- dta_nums |>
  pivot_longer(
    cols = -c(event, study_id, hiv_pos),
    names_to = "key",
    values_to = "value"
  )

summary_stats_clinical <- list()

for (var in retinal_var) {
  summary_statistics <- dta_nums |>
    filter(key == var) |>
    summarise(
      mean = mean(value, na.rm = TRUE),
      std_dev = sd(value, na.rm = TRUE),
      `1st_quartile` = quantile(value, 0.25, na.rm = TRUE),
      median = median(value, na.rm = TRUE),
      `3rd_quartile` = quantile(value, 0.75, na.rm = TRUE),
      min = min(value, na.rm = TRUE),
      max = max(value, na.rm = TRUE),
      .groups = "drop"
    )
  summary_stats_clinical[[var]] <- summary_statistics
}

summary_stats_df_numeric <- bind_rows(summary_stats_clinical, .id = "variable")

datatable(summary_stats_df_numeric) |>
  formatStyle(
    columns = c(
      "mean", "variable", "std_dev", "1st_quartile", "median", "3rd_quartile",
      "min", "max"
    ),
    fontSize = "11px",
    color = "black",
    backgroundColor = "white"
  ) |>
  formatRound(
    columns = c(
      "mean", "std_dev", "1st_quartile", "median", "3rd_quartile", "min", "max"
    ),
    digits = 2
  )
```

### HIV-positive vs HIV-negative: numeric

```{r}
#| label: summary-hiv-numeric
# Clinical variables by HIV status
dta_nums <- dta |>
  select(study_id, event, hiv_pos, all_of(nums))

dta_nums <- dta_nums |>
  pivot_longer(
    cols = -c(event, study_id, hiv_pos),
    names_to = "key",
    values_to = "value"
  )

hiv_summary_stats_clinical <- list()

for (var in clinical_var) {
  hiv_summary_statistics <- dta_nums |>
    filter(key == var) |>
    group_by(hiv_pos) |>
    summarise(
      mean = mean(value, na.rm = TRUE),
      std_dev = sd(value, na.rm = TRUE),
      `1st_quartile` = quantile(value, 0.25, na.rm = TRUE),
      median = median(value, na.rm = TRUE),
      `3rd_quartile` = quantile(value, 0.75, na.rm = TRUE),
      min = min(value, na.rm = TRUE),
      max = max(value, na.rm = TRUE),
      .groups = "drop"
    )
  hiv_summary_stats_clinical[[var]] <- hiv_summary_statistics
}

hiv_summary_stats_df_numeric <- bind_rows(
  hiv_summary_stats_clinical,
  .id = "variable"
)

datatable(hiv_summary_stats_df_numeric) |>
  formatStyle(
    columns = c(
      "mean", "variable", "std_dev", "1st_quartile", "median", "3rd_quartile",
      "min", "max"
    ),
    fontSize = "11px",
    color = "black",
    backgroundColor = "white"
  ) |>
  formatRound(
    columns = c(
      "mean", "std_dev", "1st_quartile", "median", "3rd_quartile", "min", "max"
    ),
    digits = 2
  )
```

### HIV-positive vs HIV-negative: categorical

```{r}
#| label: summary-hiv-categorical
# Categorical variables by HIV status
dta_cats <- dta |>
  select(study_id, event, hiv_pos, all_of(cats)) |>
  pivot_longer(
    cols = -c(event, study_id, hiv_pos),
    names_to = "key",
    values_to = "value"
  )

hiv_summary_stats_categorical <- list()

for (var in character_values) {
  hiv_summary_statistics <- dta_cats |>
    filter(key == var) |>
    group_by(hiv_pos, key, value) |>
    summarise(count = n(), .groups = "drop") |>
    group_by(hiv_pos) |>
    mutate(percent = (count / sum(count)) * 100) |>
    select(category = value, percent)

  hiv_summary_stats_categorical[[var]] <- hiv_summary_statistics
}

hiv_summary_stats_df_categorical <- bind_rows(
  hiv_summary_stats_categorical,
  .id = "variable"
)

datatable(hiv_summary_stats_df_categorical) |>
  formatStyle(
    columns = c("variable", "hiv_pos", "category", "percent"),
    fontSize = "11px",
    color = "black",
    backgroundColor = "white"
  ) |>
  formatRound(
    columns = c("percent"),
    digits = 2
  )
```

### HIV-positive vs HIV-negative: retinal

```{r}
#| label: summary-hiv-retinal
# Retinal measures by HIV status
dta_nums <- dta |>
  select(study_id, event, hiv_pos, all_of(retinal_var))

dta_nums <- dta_nums |>
  pivot_longer(
    cols = -c(event, study_id, hiv_pos),
    names_to = "key",
    values_to = "value"
  )

hiv_summary_stats_clinical <- list()

for (var in retinal_var) {
  hiv_summary_statistics <- dta_nums |>
    filter(key == var) |>
    group_by(hiv_pos) |>
    summarise(
      mean = mean(value, na.rm = TRUE),
      std_dev = sd(value, na.rm = TRUE),
      `1st_quartile` = quantile(value, 0.25, na.rm = TRUE),
      median = median(value, na.rm = TRUE),
      `3rd_quartile` = quantile(value, 0.75, na.rm = TRUE),
      min = min(value, na.rm = TRUE),
      max = max(value, na.rm = TRUE),
      .groups = "drop"
    )
  hiv_summary_stats_clinical[[var]] <- hiv_summary_statistics
}

hiv_summary_stats_df_numeric <- bind_rows(
  hiv_summary_stats_clinical,
  .id = "variable"
)

datatable(hiv_summary_stats_df_numeric) |>
  formatStyle(
    columns = c(
      "mean", "variable", "std_dev", "1st_quartile", "median", "3rd_quartile",
      "min", "max"
    ),
    fontSize = "11px",
    color = "black",
    backgroundColor = "white"
  ) |>
  formatRound(
    columns = c(
      "mean", "std_dev", "1st_quartile", "median", "3rd_quartile", "min", "max"
    ),
    digits = 2
  )
```

### ART-naive vs on ART: numeric

```{r}
#| label: summary-art-numeric
# Clinical variables by ART status
dta_nums <- dta |>
  select(study_id, event, on_art, all_of(nums))

dta_nums <- dta_nums |>
  pivot_longer(
    cols = -c(event, study_id, on_art),
    names_to = "key",
    values_to = "value"
  )
art_summary_stats_clinical <- list()

for (var in clinical_var) {
  art_summary_statistics <- dta_nums |>
    filter(key == var) |>
    group_by(on_art) |>
    summarise(
      mean = mean(value, na.rm = TRUE),
      std_dev = sd(value, na.rm = TRUE),
      `1st_quartile` = quantile(value, 0.25, na.rm = TRUE),
      median = median(value, na.rm = TRUE),
      `3rd_quartile` = quantile(value, 0.75, na.rm = TRUE),
      min = min(value, na.rm = TRUE),
      max = max(value, na.rm = TRUE),
      .groups = "drop"
    )
  art_summary_stats_clinical[[var]] <- art_summary_statistics
}

art_summary_stats_df_numeric <- bind_rows(
  art_summary_stats_clinical,
  .id = "variable"
)

datatable(art_summary_stats_df_numeric) |>
  formatStyle(
    columns = c(
      "mean", "variable", "std_dev", "1st_quartile", "median", "3rd_quartile",
      "min", "max"
    ),
    fontSize = "11px",
    color = "black",
    backgroundColor = "white"
  ) |>
  formatRound(
    columns = c(
      "mean", "std_dev", "1st_quartile", "median", "3rd_quartile", "min", "max"
    ),
    digits = 2
  )
```

### ART-naive vs on ART: categorical

```{r}
#| label: summary-art-categorical
# Categorical variables by ART status
dta_cats <- dta |>
  select(study_id, event, on_art, all_of(cats)) |>
  pivot_longer(
    cols = -c(event, study_id, on_art),
    names_to = "key",
    values_to = "value"
  )

art_summary_stats_categorical <- list()

for (var in character_values) {
  art_summary_statistics <- dta_cats |>
    filter(key == var) |>
    group_by(on_art, key, value) |>
    summarise(count = n(), .groups = "drop") |>
    group_by(on_art) |>
    mutate(percent = (count / sum(count)) * 100) |>
    select(category = value, percent)

  art_summary_stats_categorical[[var]] <- art_summary_statistics
}

art_summary_stats_df_categorical <- bind_rows(
  art_summary_stats_categorical,
  .id = "variable"
)

datatable(art_summary_stats_df_categorical) |>
  formatStyle(
    columns = c("variable", "on_art", "category", "percent"),
    fontSize = "11px",
    color = "black",
    backgroundColor = "white"
  )
```

### ART-naive vs on ART: retinal

```{r}
#| label: summary-art-retinal
# Retinal measures by ART status
dta_nums <- dta |>
  select(study_id, event, on_art, all_of(retinal_var))

dta_nums <- dta_nums |>
  pivot_longer(
    cols = -c(event, study_id, on_art),
    names_to = "key",
    values_to = "value"
  )

hiv_summary_stats_clinical <- list()

for (var in retinal_var) {
  hiv_summary_statistics <- dta_nums |>
    filter(key == var) |>
    group_by(on_art) |>
    summarise(
      mean = mean(value, na.rm = TRUE),
      std_dev = sd(value, na.rm = TRUE),
      `1st_quartile` = quantile(value, 0.25, na.rm = TRUE),
      median = median(value, na.rm = TRUE),
      `3rd_quartile` = quantile(value, 0.75, na.rm = TRUE),
      min = min(value, na.rm = TRUE),
      max = max(value, na.rm = TRUE),
      .groups = "drop"
    )
  hiv_summary_stats_clinical[[var]] <- hiv_summary_statistics
}

hiv_summary_stats_df_numeric <- bind_rows(
  hiv_summary_stats_clinical,
  .id = "variable"
)

datatable(hiv_summary_stats_df_numeric) |>
  formatStyle(
    columns = c(
      "mean", "variable", "std_dev", "1st_quartile", "median", "3rd_quartile",
      "min", "max"
    ),
    fontSize = "11px",
    color = "black",
    backgroundColor = "white"
  ) |>
  formatRound(
    columns = c(
      "mean", "std_dev", "1st_quartile", "median", "3rd_quartile", "min", "max"
    ),
    digits = 2
  )
```

:::

## Exploration: HIV-positive vs HIV-negative

Bar graphs, box plots and density plots comparing HIV-positive and HIV-negative participants.

::: panel-tabset

### Bar graphs

```{r}
#| label: explore-hiv-bar
# Bar graphs of categorical variables by HIV status

# select character data
dta_cats <- dta |>
  select(study_id, event, hiv_pos, all_of(cats))

# pivot character data
dta_cats <- dta_cats |>
  pivot_longer(
    cols = -c(event, study_id, hiv_pos),
    names_to = "key",
    values_to = "value"
  )

plot_bar <- function(data, vars, var_fill, angl, ncols) {
  tmp <- data |>
    distinct(study_id, hiv_pos) |>
    group_by(hiv_pos) |>
    count(name = "total") |>
    inner_join(data |>
      group_by(hiv_pos, key, value) |>
      count()) |>
    mutate(perc = n / total * 100) |>
    arrange(key)

  unique_values <- unique(tmp$value)
  color_map_filtered <- color_map_list[names(color_map_list) %in% unique_values]

  p <- tmp |>
    filter(key %in% vars) |>
    ggplot(aes(x = hiv_pos, y = perc, fill = value)) +
    geom_col(position = "dodge") +
    facet_wrap(~key, scales = "free", ncol = ncols) +
    labs(y = "Percentage") +
    scale_fill_manual(values = color_map_list) +
    theme(
      axis.text.x = element_text(angle = angl, hjust = 1),
      legend.position = "bottom"
    )

  return(p)
}

for (group in list(
  past_medical_history, medical_condition, family_history_1, family_history_2,
  gender
)) {
  plot_bar(
    data = dta_cats,
    vars = group,
    var_fill = "value",
    angl = 45,
    ncols = 2
  ) |>
    print()
}

for (group in list(diabetes_lev, lifestyle_factors)) {
  plot_bar(
    data = dta_cats,
    vars = group,
    var_fill = "value",
    angl = 45,
    ncols = 3
  ) |>
    print()
}
```

### Box plots

```{r}
#| label: explore-hiv-box
# Box plots of numeric variables by HIV status

# select numeric data
dta_nums <- dta |>
  select(study_id, event, hiv_pos, all_of(nums))

# pivot numeric data
dta_nums <- dta_nums |>
  pivot_longer(
    cols = -c(event, study_id, hiv_pos),
    names_to = "key",
    values_to = "value"
  )

# make function
plot_box <- function(data, vars, var_fill, angl, ncols) {
  p <- data |>
    filter(key %in% vars) |>
    ggplot(aes(x = hiv_pos, y = value, color = hiv_pos)) +
    geom_boxplot(fill = NA) +
    scale_color_manual(values = var_fill) +
    facet_wrap(~key, scales = "free", ncol = ncols) +
    theme(
      axis.text.x = element_text(angle = angl, hjust = 1),
      legend.position = "bottom"
    )

  return(p)
}

# for loop to make graphs
for (group in list(
  health_dta, mean_heart_measurements, clinical_dta_10, clinical_dta_11,
  clinical_dta_12, clinical_dta_13, clinical_dta_14, clinical_dta_15,
  clinical_dta_16, retinal_var_1, retinal_var_2, retinal_var_3, retinal_var_4,
  retinal_var_5, retinal_var_6, retinal_var_7, retinal_var_8, retinal_var_9,
  log_values
)) {
  plot_box(
    data = dta_nums,
    vars = group,
    var_fill = color_map_list,
    angl = 45,
    ncols = 3
  ) |>
    print()
}
```

### Density plots

```{r}
#| label: explore-hiv-density
# Density plots of numeric variables by HIV status

# Select relevant columns from the original data
dta_nums <- dta |>
  select(study_id, event, hiv_pos, all_of(nums))

# Pivot numeric data
dta_nums <- dta_nums |>
  pivot_longer(
    cols = -c(event, study_id, hiv_pos),
    names_to = "key",
    values_to = "value"
  )

# Function to create density plots
plot_density <- function(data, vars, var_fill, angl, ncols) {
  p <- data |>
    filter(key %in% vars) |>
    na.omit() |>
    ggplot(aes(x = value, color = .data[[var_fill]])) +
    geom_density(fill = NA, linewidth = 0.5) +
    facet_wrap(~key, scales = "free", ncol = ncols) +
    scale_color_manual(values = c("No" = "#C52233", "Yes" = "#2A7221")) +
    theme(
      axis.text.x = element_text(size = 18), # Increase x-axis number size
      axis.title.x = element_text(size = 20), # Increase x-axis title size
      axis.text.y = element_text(size = 16), # Adjust y-axis number size
      axis.title.y = element_text(size = 20),
      strip.text = element_text(size = 16),
      legend.text = element_text(size = 16), # Size of legend text
      legend.title = element_text(size = 20),
      legend.position = "bottom"
    )

  return(p)
}

combined_plot <- plot_density(
  data = dta_nums,
  vars = retinal_var_2,
  var_fill = "hiv_pos",
  angl = 45,
  ncols = 3
)
print(combined_plot)
ggsave(
  "hiv_ret_var_combined.svg",
  plot = combined_plot,
  width = 12,
  height = 10
)

for (group in list(
  health_dta, mean_heart_measurements, clinical_dta_10, clinical_dta_11,
  clinical_dta_12, clinical_dta_13, clinical_dta_14, clinical_dta_15,
  clinical_dta_16, clinical_dta_17, retinal_var_1, retinal_var_2, retinal_var_3,
  retinal_var_4, retinal_var_5, retinal_var_6, retinal_var_7, retinal_var_8,
  retinal_var_9
)) {
  plot_density(
    data = dta_nums,
    vars = group,
    var_fill = "hiv_pos",
    angl = 45,
    ncols = 3
  ) |>
    print()
}

for (group in list(bimodal_var_1, bimodal_var_2, bimodal_var_3)) {
  plot_density(
    data = dta_nums,
    vars = group,
    var_fill = "hiv_pos",
    angl = 45,
    ncols = 3
  ) |>
    print()
}

for (group in list(bimodal_var_4, bimodal_var_5)) {
  plot_density(
    data = dta_nums,
    vars = group,
    var_fill = "hiv_pos",
    angl = 45,
    ncols = 2
  ) |>
    print()
}
```

:::

## Exploration: on ART vs ART-naive

The same plots for participants on ART and ART-naive participants, plus retinal measures by ART duration and viral load group.

::: panel-tabset

### Bar graphs

```{r}
#| label: explore-art-bar
# Bar graphs of categorical variables by ART status

# select character data
dta_cats <- dta |>
  select(study_id, event, hiv_pos, on_art, all_of(cats))

# pivot character data
dta_cats <- dta_cats |>
  pivot_longer(
    cols = -c(event, study_id, hiv_pos, on_art),
    names_to = "key",
    values_to = "value"
  )

# make function
plot_bar <- function(data, vars, var_fill, angl, ncols) {
  data <- data |>
    filter(!is.na(on_art))

  tmp <- data |>
    distinct(study_id, on_art) |>
    group_by(on_art) |>
    count(name = "total") |>
    inner_join(data |>
      group_by(on_art, key, value) |> count()) |>
    mutate(perc = n / total * 100) |>
    arrange(key)

  p <- tmp |>
    filter(key %in% vars) |>
    ggplot(aes(x = hiv_pos, y = perc, fill = value)) +
    geom_col(position = "dodge") +
    facet_wrap(~key, scales = "free", ncol = ncols) +
    labs(y = "Percentage") +
    scale_fill_manual(values = color_map_list) +
    theme(
      axis.text.x = element_text(angle = angl, hjust = 1),
      legend.position = "bottom"
    )

  return(p)
}

for (group in list(
  gender, past_medical_history, medical_condition, family_history_1,
  family_history_2, hiv_related
)) {
  plot_bar(
    data = dta_cats,
    vars = group,
    var_fill = "value",
    angl = 45,
    ncols = 2
  ) |>
    print()
}

for (group in list(diabetes_lev, lifestyle_factors)) {
  plot_bar(
    data = dta_cats,
    vars = group,
    var_fill = "value",
    angl = 45,
    ncols = 3
  ) |>
    print()
}
```

### Box plots

```{r}
#| label: explore-art-box
# Box plots of numeric variables by ART status

# select numeric data
dta_nums <- dta |>
  select(study_id, event, hiv_pos, on_art, all_of(nums))

# pivot numeric data
dta_nums <- dta_nums |>
  pivot_longer(
    cols = -c(event, study_id, hiv_pos, on_art),
    names_to = "key",
    values_to = "value"
  )

# make function
plot_box <- function(data, vars, var_fill, angl, ncols) {
  data <- data |>
    filter(!is.na(on_art))

  p <- data |>
    filter(key %in% vars) |>
    ggplot(aes(x = on_art, y = value, color = on_art)) +
    geom_boxplot(fill = NA) +
    scale_color_manual(values = var_fill) +
    facet_wrap(~key, scales = "free", ncol = ncols) +
    theme(
      axis.text.x = element_text(angle = angl, hjust = 1),
      legend.position = "bottom"
    )

  return(p)
}

for (group in list(
  health_dta, mean_heart_measurements, clinical_dta_10, clinical_dta_11,
  clinical_dta_12, clinical_dta_13, clinical_dta_14, clinical_dta_15,
  clinical_dta_16, clinical_dta_17, retinal_var_1, retinal_var_2, retinal_var_3,
  retinal_var_4, retinal_var_5, retinal_var_6, retinal_var_7, retinal_var_8,
  retinal_var_9, log_values
)) {
  plot_box(
    data = dta_nums,
    vars = group,
    var_fill = color_map_list,
    angl = 45,
    ncols = 3
  ) |>
    print()
}
```

### Box plots: ART duration

```{r}
#| label: explore-art-box-duration
# Box plots of retinal measures by ART duration group
dta_nums <- dta |>
  select(study_id, on_art, art_duration_group, all_of(retinal_var))

# pivot numeric data
dta_nums <- dta_nums |>
  pivot_longer(
    cols = -c(study_id, on_art, art_duration_group),
    names_to = "key",
    values_to = "value"
  )

# make function
plot_box <- function(data, vars, var_fill, angl, ncols) {
  data <- data |>
    filter(!is.na(on_art))

  p <- data |>
    filter(key %in% vars) |>
    ggplot(aes(x = art_duration_group, y = value, color = art_duration_group)) +
    geom_boxplot(fill = NA) +
    scale_color_manual(values = var_fill) +
    facet_wrap(~key, scales = "free", ncol = ncols) +
    theme(
      axis.text.x = element_text(angle = angl, hjust = 1),
      legend.position = "bottom"
    )

  return(p)
}

for (group in list(
  retinal_var_1, retinal_var_2, retinal_var_3, retinal_var_4, retinal_var_5,
  retinal_var_6, retinal_var_7, retinal_var_8, retinal_var_9
)) {
  plot_box(
    data = dta_nums,
    vars = group,
    var_fill = color_map_list,
    angl = 45,
    ncols = 3
  ) |>
    print()
}
```

### Box plots: viral load

```{r}
#| label: explore-art-box-viral-load
# Box plots of retinal measures by viral load group
dta_nums <- dta |>
  select(study_id, on_art, viral_load_group, all_of(retinal_var))

# pivot numeric data
dta_nums <- dta_nums |>
  pivot_longer(
    cols = -c(study_id, on_art, viral_load_group),
    names_to = "key",
    values_to = "value"
  )

# make function
plot_box <- function(data, vars, var_fill, angl, ncols) {
  data <- data |>
    filter(!is.na(on_art))

  p <- data |>
    filter(key %in% vars) |>
    ggplot(aes(x = viral_load_group, y = value, color = viral_load_group)) +
    geom_boxplot(fill = NA) +
    scale_color_manual(values = var_fill) +
    facet_wrap(~key, scales = "free", ncol = ncols) +
    theme(
      axis.text.x = element_text(angle = angl, hjust = 1),
      legend.position = "bottom"
    )

  return(p)
}

for (group in list(
  retinal_var_1, retinal_var_2, retinal_var_3, retinal_var_4, retinal_var_5,
  retinal_var_6, retinal_var_7, retinal_var_8, retinal_var_9
)) {
  plot_box(
    data = dta_nums,
    vars = group,
    var_fill = color_map_list,
    angl = 45,
    ncols = 3
  ) |>
    print()
}
```

### Density plots

```{r}
#| label: explore-art-density
# Density plots of numeric variables by ART status

# select numeric data
dta_nums <- dta |>
  select(study_id, event, on_art, all_of(nums))

# pivot numeric data
dta_nums <- dta_nums |>
  pivot_longer(
    cols = -c(event, study_id, on_art),
    names_to = "key",
    values_to = "value"
  )

plot_density <- function(data, vars, var_fill, angl, ncols) {
  data <- data |>
    filter(!is.na(on_art))

  p <- data |>
    filter(key %in% vars) |>
    na.omit() |>
    ggplot(aes(x = value, color = .data[[var_fill]])) +
    geom_density(fill = NA, linewidth = 0.5) +
    facet_wrap(~key, scales = "free", ncol = ncols) +
    scale_color_manual(values = c("No" = "#C52233", "Yes" = "#2A7221")) +
    theme(
      axis.text.x = element_text(size = 18), # Increase x-axis number size
      axis.title.x = element_text(size = 20), # Increase x-axis title size
      axis.text.y = element_text(size = 16), # Adjust y-axis number size
      axis.title.y = element_text(size = 20),
      strip.text = element_text(size = 16),
      legend.text = element_text(size = 16), # Size of legend text
      legend.title = element_text(size = 20),
      legend.position = "bottom"
    )

  return(p)
}

combined_plot <- plot_density(
  data = dta_nums,
  vars = retinal_var_2,
  var_fill = "on_art",
  angl = 45,
  ncols = 3
)
print(combined_plot)
ggsave(
  "art_ret_var_combined.svg",
  plot = combined_plot,
  width = 12,
  height = 10
)

for (group in list(
  health_dta, mean_heart_measurements, clinical_dta_10, clinical_dta_11,
  clinical_dta_12, clinical_dta_13, clinical_dta_14, clinical_dta_15,
  clinical_dta_16, clinical_dta_17, retinal_var_1, retinal_var_2, retinal_var_3,
  retinal_var_4, retinal_var_5, retinal_var_6, retinal_var_7, retinal_var_8,
  retinal_var_9
)) {
  plot_density(
    data = dta_nums,
    vars = group,
    var_fill = "on_art",
    angl = 45,
    ncols = 3
  ) |>
    print()
}

for (group in list(bimodal_var_1, bimodal_var_2, bimodal_var_3)) {
  plot_density(
    data = dta_nums,
    vars = group,
    var_fill = "on_art",
    angl = 45,
    ncols = 3
  ) |>
    print()
}

for (group in list(bimodal_var_4, bimodal_var_5)) {
  plot_density(
    data = dta_nums,
    vars = group,
    var_fill = "on_art",
    angl = 45,
    ncols = 2
  ) |>
    print()
}
```

:::

## Exploration: reasons for bimodal distributions

Density plots checking whether retinal measures with two peaks split by gender, diabetes, smoking, recruitment site, hypertension, high cholesterol, season or image quality.

::: panel-tabset

### Gender

```{r}
#| label: bimodal-gender
# Density plots by gender

# select numeric data
dta_nums <- dta |>
  select(study_id, event, gender, all_of(nums))

# pivot numeric data
dta_nums <- dta_nums |>
  pivot_longer(
    cols = -c(event, study_id, gender),
    names_to = "key",
    values_to = "value"
  )

plot_density <- function(data, vars, var_fill, angl, ncols) {
  p <- data |>
    filter(key %in% vars) |>
    na.omit() |>
    ggplot(aes(x = value, fill = .data[[var_fill]])) +
    geom_density(alpha = 0.5) +
    facet_wrap(~key, scales = "free", ncol = ncols) +
    theme(
      axis.text.x = element_text(angle = angl, hjust = 1),
      legend.position = "bottom"
    )

  return(p)
}

for (group in list(
  parameters_1, parameters_2, parameters_3, parameters_4, parameters_5
)) {
  plot_density(
    data = dta_nums,
    vars = group,
    var_fill = "gender",
    angl = 45,
    ncols = 3
  ) |>
    print()
}
```

### Diabetes

```{r}
#| label: bimodal-diabetes
# Density plots by diabetes

# select numeric data
dta_nums <- dta |>
  select(study_id, event, diabetes, all_of(nums))

# pivot numeric data
dta_nums <- dta_nums |>
  pivot_longer(
    cols = -c(event, study_id, diabetes),
    names_to = "key",
    values_to = "value"
  )

plot_density <- function(data, vars, var_fill, angl, ncols) {
  p <- data |>
    filter(key %in% vars) |>
    na.omit() |>
    ggplot(aes(x = value, fill = .data[[var_fill]])) +
    geom_density(alpha = 0.5) +
    facet_wrap(~key, scales = "free", ncol = ncols) +
    theme(
      axis.text.x = element_text(angle = angl, hjust = 1),
      legend.position = "bottom"
    )

  return(p)
}

for (group in list(
  parameters_1, parameters_2, parameters_3, parameters_4, parameters_5
)) {
  plot_density(
    data = dta_nums,
    vars = group,
    var_fill = "diabetes",
    angl = 45,
    ncols = 3
  ) |>
    print()
}
```

### Smoking

```{r}
#| label: bimodal-smoking
# Density plots by smoking status

# select numeric data
dta_nums <- dta |>
  select(study_id, event, smoker, all_of(nums))

# pivot numeric data
dta_nums <- dta_nums |>
  pivot_longer(
    cols = -c(event, study_id, smoker),
    names_to = "key",
    values_to = "value"
  )

plot_density <- function(data, vars, var_fill, angl, ncols) {
  p <- data |>
    filter(key %in% vars) |>
    na.omit() |>
    ggplot(aes(x = value, fill = .data[[var_fill]])) +
    geom_density(alpha = 0.5) +
    facet_wrap(~key, scales = "free", ncol = ncols) +
    theme(
      axis.text.x = element_text(angle = angl, hjust = 1),
      legend.position = "bottom"
    )

  return(p)
}

for (group in list(
  parameters_1, parameters_2, parameters_3, parameters_4, parameters_5
)) {
  plot_density(
    data = dta_nums,
    vars = group,
    var_fill = "smoker",
    angl = 45,
    ncols = 3
  ) |>
    print()
}
```

### Recruitment site

```{r}
#| label: bimodal-site
# Density plots by recruitment site

# select numeric data
dta_nums <- dta |>
  select(study_id, event, recruitment_site, all_of(nums))

# pivot numeric data
dta_nums <- dta_nums |>
  pivot_longer(
    cols = -c(event, study_id, recruitment_site),
    names_to = "key",
    values_to = "value"
  )

plot_density <- function(data, vars, var_fill, angl, ncols) {
  p <- data |>
    filter(key %in% vars) |>
    na.omit() |>
    ggplot(aes(x = value, fill = .data[[var_fill]])) +
    geom_density(alpha = 0.5) +
    facet_wrap(~key, scales = "free", ncol = ncols) +
    theme(
      axis.text.x = element_text(angle = angl, hjust = 1),
      legend.position = "bottom"
    )

  return(p)
}

for (group in list(
  parameters_1, parameters_2, parameters_3, parameters_4, parameters_5
)) {
  plot_density(
    data = dta_nums,
    vars = group,
    var_fill = "recruitment_site",
    angl = 45,
    ncols = 3
  ) |>
    print()
}
```

### Hypertension

```{r}
#| label: bimodal-hypertension
# Density plots by hypertension

# select numeric data
dta_nums <- dta |>
  select(study_id, event, hypertension, all_of(nums))

# pivot numeric data
dta_nums <- dta_nums |>
  pivot_longer(
    cols = -c(event, study_id, hypertension),
    names_to = "key",
    values_to = "value"
  )

plot_density <- function(data, vars, var_fill, angl, ncols) {
  p <- data |>
    filter(key %in% vars) |>
    na.omit() |>
    ggplot(aes(x = value, fill = .data[[var_fill]])) +
    geom_density(alpha = 0.5) +
    facet_wrap(~key, scales = "free", ncol = ncols) +
    theme(
      axis.text.x = element_text(angle = angl, hjust = 1),
      legend.position = "bottom"
    )

  return(p)
}

for (group in list(
  parameters_1, parameters_2, parameters_3, parameters_4, parameters_5
)) {
  plot_density(
    data = dta_nums,
    vars = group,
    var_fill = "hypertension",
    angl = 45,
    ncols = 3
  ) |>
    print()
}
```

### High cholesterol

```{r}
#| label: bimodal-cholesterol
# Density plots by high cholesterol

# select numeric data
dta_nums <- dta |>
  select(study_id, event, high_cholesterol, all_of(nums))

# pivot numeric data
dta_nums <- dta_nums |>
  pivot_longer(
    cols = -c(event, study_id, high_cholesterol),
    names_to = "key",
    values_to = "value"
  )

plot_density <- function(data, vars, var_fill, angl, ncols) {
  p <- data |>
    filter(key %in% vars) |>
    na.omit() |>
    ggplot(aes(x = value, fill = .data[[var_fill]])) +
    geom_density(alpha = 0.5) +
    facet_wrap(~key, scales = "free", ncol = ncols) +
    theme(
      axis.text.x = element_text(angle = angl, hjust = 1),
      legend.position = "bottom"
    )

  return(p)
}

for (group in list(
  parameters_1, parameters_2, parameters_3, parameters_4, parameters_5
)) {
  plot_density(
    data = dta_nums,
    vars = group,
    var_fill = "high_cholesterol",
    angl = 45,
    ncols = 3
  ) |>
    print()
}
```

### Warm vs cold months

```{r}
#| label: bimodal-season
# Density plots by season
dta_nums <- dta |>
  select(study_id, event, season, all_of(retinal_var))

dta_nums <- dta_nums |>
  pivot_longer(
    cols = -c(event, study_id, season),
    names_to = "key",
    values_to = "value"
  )

plot_density <- function(data, vars, var_fill, angl, ncols) {
  p <- data |>
    filter(key %in% vars) |>
    na.omit() |>
    ggplot(aes(x = value, fill = .data[[var_fill]])) +
    geom_density(alpha = 0.5) +
    facet_wrap(~key, scales = "free", ncol = ncols) +
    theme(
      axis.text.x = element_text(angle = angl, hjust = 1),
      legend.position = "bottom"
    )

  return(p)
}

for (group in list(
  parameters_1, parameters_2, parameters_3, parameters_4, parameters_5
)) {
  plot_density(
    data = dta_nums,
    vars = group,
    var_fill = "season",
    angl = 45,
    ncols = 3
  ) |>
    print()
}
```

### Image quality

```{r}
#| label: bimodal-image-quality
# Density plots by image quality (clarity)
dta_nums <- dta |>
  select(study_id, event, clarity, all_of(retinal_var))

dta_nums <- dta_nums |>
  pivot_longer(
    cols = -c(event, study_id, clarity),
    names_to = "key",
    values_to = "value"
  )

plot_density <- function(data, vars, var_fill, angl, ncols) {
  p <- data |>
    filter(key %in% vars) |>
    na.omit() |>
    ggplot(aes(x = value, fill = .data[[var_fill]])) +
    geom_density(alpha = 0.5) +
    facet_wrap(~key, scales = "free", ncol = ncols) +
    theme(
      axis.text.x = element_text(angle = angl, hjust = 1),
      legend.position = "bottom"
    )

  return(p)
}

for (group in list(
  parameters_1, parameters_2, parameters_3, parameters_4, parameters_5
)) {
  plot_density(
    data = dta_nums,
    vars = group,
    var_fill = "clarity",
    angl = 45,
    ncols = 3
  ) |>
    print()
}
```

### Recruitment site groups

```{r}
#| label: bimodal-site-group
# Density plots by recruitment site group
dta_nums <- dta |>
  select(study_id, event, site_group, all_of(retinal_var))

dta_nums <- dta_nums |>
  pivot_longer(
    cols = -c(event, study_id, site_group),
    names_to = "key",
    values_to = "value"
  )

plot_density <- function(data, vars, var_fill, angl, ncols) {
  p <- data |>
    filter(key %in% vars) |>
    na.omit() |>
    ggplot(aes(x = value, fill = .data[[var_fill]])) +
    geom_density(alpha = 0.5) +
    facet_wrap(~key, scales = "free", ncol = ncols) +
    theme(
      axis.text.x = element_text(angle = angl, hjust = 1),
      legend.position = "bottom"
    )

  return(p)
}

for (group in list(
  parameters_1, parameters_2, parameters_3, parameters_4, parameters_5
)) {
  plot_density(
    data = dta_nums,
    vars = group,
    var_fill = "site_group",
    angl = 45,
    ncols = 3
  ) |>
    print()
}
```

### Bimodal retinal variables: HIV status

```{r}
#| label: bimodal-vars-hiv
# Density plots of the bimodal retinal measures by HIV status
dta_nums <- dta |>
  select(study_id, event, hiv_pos, all_of(bimodal_var))

dta_nums <- dta_nums |>
  pivot_longer(
    cols = -c(event, study_id, hiv_pos),
    names_to = "key",
    values_to = "value"
  )

plot_density <- function(data, vars, var_fill, angl, ncols) {
  p <- data |>
    filter(key %in% vars) |>
    na.omit() |>
    ggplot(aes(x = value, fill = .data[[var_fill]])) +
    geom_density(alpha = 0.5) +
    facet_wrap(~key, scales = "free", ncol = ncols) +
    theme(
      axis.text.x = element_text(angle = angl, hjust = 1),
      legend.position = "bottom"
    )

  return(p)
}

for (group in list(bimodal_var_1, bimodal_var_2)) {
  plot_density(
    data = dta_nums,
    vars = group,
    var_fill = "hiv_pos",
    angl = 45,
    ncols = 3
  ) |>
    print()
}
```

### Bimodal retinal variables: ART status

```{r}
#| label: bimodal-vars-art
# Density plots of the bimodal retinal measures by ART status
# Note: this overwrites bimodal_var for the rest of the document
x <- bimodal_var

xx <- which(x %in% c(
  "fourier_fractal_enhanced", "arteriole_mean_branching_coefficient",
  "branch_segment_length_skewness", "lacunarity_skeleton", "tortuosity_index"
))

bimodal_var <- x[-xx]

dta_nums <- dta |>
  select(study_id, event, on_art, all_of(bimodal_var))

dta_nums <- dta_nums |>
  pivot_longer(
    cols = -c(event, study_id, on_art),
    names_to = "key",
    values_to = "value"
  )

plot_density <- function(data, vars, var_fill, angl, ncols) {
  data <- data |>
    filter(!is.na(on_art))

  p <- data |>
    filter(key %in% vars) |>
    na.omit() |>
    ggplot(aes(x = value, fill = .data[[var_fill]])) +
    geom_density(alpha = 0.5) +
    facet_wrap(~key, scales = "free", ncol = ncols) +
    theme(
      axis.text.x = element_text(angle = angl, hjust = 1),
      legend.position = "bottom"
    )

  return(p)
}

for (group in list(bimodal_var)) {
  plot_density(
    data = dta_nums,
    vars = group,
    var_fill = "on_art",
    angl = 45,
    ncols = 3
  ) |>
    print()
}

# Note: this redefines nums for all later sections. It drops most bimodal
# retinal measures plus the raw body measurements and the individual
# blood pressure and heart rate readings.
x <- nums
xx <- which(x %in% c(
  "segment_tree_area", "total_length_skeletonised_tree", "branch_points",
  "endpoints", "fourier_fractal_analysis", "lacunarity_segmented",
  "lacunarity_skeleton", "retinal_area", "branch_segment_length_skewness",
  "branch_segment_tortuosity_average", "tortuosity_index",
  "arteriole_mean_branching_coefficient", "height", "hip_circumference",
  "waist_circumference", "total_body_mass", "diastolic_bp", "systolic_bp",
  "heart_rate", "diastolic_bp_2", "diastolic_bp_3", "systolic_bp_2",
  "systolic_bp_3", "heart_rate_2", "heart_rate_3"
))
nums <- x[-xx]
```

:::

## Difference tests (t-tests)

Welch two-sample t-tests of each numeric variable between two groups, sorted by p-value.

::: panel-tabset

### HIV status

```{r}
#| label: ttest-hiv
# Welch t-tests of numeric variables by HIV status
dta_nums <- dta |>
  select(study_id, event, hiv_pos, all_of(nums)) |>
  pivot_longer(
    cols = -c(event, study_id, hiv_pos),
    names_to = "key",
    values_to = "value"
  )

x <- unique(dta_nums$key)
xx <- which(x %in% c(
  "last_menstrual_period_days", "aspirin_dosage", "consent_date",
  "interview_start_time", "art_duration_weeks", "age_began_regular_drinking",
  "red_wine_units", "white_wine_units", "cd4_count_clinic", "endo_result",
  "viral_load", "log_crp", "log_ggt", "log_viral_load", "log_acr_ratio",
  "log_microalbuminuria", "log_ten_year_cvd_risk"
))
x <- x[-xx] # drop variables not tested here

t_test_results <- vector("list", length(x))

for (i in 1:length(x)) {
  tmp <- dta_nums |>
    filter(key == x[i])
  t_test <- t.test(value ~ hiv_pos, data = tmp)
  results <- tidy(t_test)
  results$variable <- x[i]
  t_test_results[[i]] <- results
}

selected_columns <- c(
  "variable", "estimate", "p.value", "conf.low", "conf.high"
)

t_results_df <- bind_rows(t_test_results) |>
  select(all_of(selected_columns)) |>
  arrange(p.value)

datatable(t_results_df) |>
  formatStyle(
    columns = c("variable", "estimate", "p.value", "conf.low", "conf.high"),
    fontSize = "12px",
    color = "black",
    backgroundColor = "white"
  ) |>
  formatRound(
    columns = c("estimate", "p.value", "conf.low", "conf.high"),
    digits = 3
  )
```

### ART status

```{r}
#| label: ttest-art
# Welch t-tests of numeric variables by ART status
dta_nums <- dta |>
  select(study_id, event, on_art, all_of(nums)) |>
  pivot_longer(
    cols = -c(event, study_id, on_art),
    names_to = "key",
    values_to = "value"
  )

x <- unique(dta_nums$key)
xx <- which(x %in% c(
  "last_menstrual_period_days", "aspirin_dosage", "consent_date",
  "interview_start_time", "art_duration_weeks", "age_began_regular_drinking",
  "red_wine_units", "white_wine_units", "log_crp", "log_ggt", "log_viral_load",
  "log_acr_ratio", "log_microalbuminuria", "log_ten_year_cvd_risk"
))
x <- x[-xx] # drop variables not tested here

t_test_results <- vector("list", length(x))

for (i in 1:length(x)) {
  tmp <- dta_nums |>
    filter(key == x[i])
  t_test <- t.test(value ~ on_art, data = tmp)
  results <- tidy(t_test)
  results$variable <- x[i]
  t_test_results[[i]] <- results
}

selected_columns <- c(
  "variable", "estimate", "p.value", "conf.low", "conf.high"
)

t_results_df <- bind_rows(t_test_results) |>
  select(all_of(selected_columns)) |>
  arrange(p.value)

datatable(t_results_df) |>
  formatStyle(
    columns = c("variable", "estimate", "p.value", "conf.low", "conf.high"),
    fontSize = "12px",
    color = "black",
    backgroundColor = "white"
  ) |>
  formatRound(
    columns = c("estimate", "p.value", "conf.low", "conf.high"),
    digits = 3
  )
```

### ART duration

```{r}
#| label: ttest-art-duration
# Welch t-tests of numeric variables by ART duration (0-5 vs >5 years)
dta_nums <- dta |>
  select(study_id, event, art_duration_group_2, all_of(nums)) |>
  pivot_longer(
    cols = -c(event, study_id, art_duration_group_2),
    names_to = "key",
    values_to = "value"
  )

dta_nums$art_duration_group_2 <- factor(dta_nums$art_duration_group_2)

levels(dta_nums$art_duration_group_2)

x <- unique(dta_nums$key)
xx <- which(x %in% c(
  "last_menstrual_period_days", "aspirin_dosage", "consent_date",
  "interview_start_time", "art_duration_weeks", "age_began_regular_drinking",
  "red_wine_units", "white_wine_units", "log_crp", "log_ggt", "log_viral_load",
  "log_acr_ratio", "log_microalbuminuria", "log_ten_year_cvd_risk"
))
x <- x[-xx] # drop variables not tested here

t_test_results <- vector("list", length(x))

for (i in 1:length(x)) {
  tmp <- dta_nums |>
    filter(key == x[i])
  t_test <- t.test(value ~ art_duration_group_2, data = tmp)
  results <- tidy(t_test)
  results$variable <- x[i]
  t_test_results[[i]] <- results
}

selected_columns <- c(
  "variable", "estimate", "p.value", "conf.low", "conf.high"
)

t_results_df <- bind_rows(t_test_results) |>
  select(all_of(selected_columns)) |>
  arrange(p.value)

datatable(t_results_df) |>
  formatStyle(
    columns = c("variable", "estimate", "p.value", "conf.low", "conf.high"),
    fontSize = "12px",
    color = "black",
    backgroundColor = "white"
  )

# Significant variables from the table above, entered by hand
sig_art_duration <- c(
  "venule_mean_asymmetry_ratio", "arterial_tree_segments", "fmd_percentage",
  "waist_hip_ratio", "branch_segment_tortuosity_kurtosis",
  "venule_branch_angle", "venule_branch_smaller_branch"
)

sig_art_duration_ret <- c(
  "venule_mean_asymmetry_ratio", "arterial_tree_segments",
  "branch_segment_tortuosity_kurtosis", "venule_branch_angle",
  "venule_branch_smaller_branch"
)
```

### Viral load

```{r}
#| label: ttest-viral-load
# Welch t-tests of numeric variables by viral load group
dta_nums <- dta |>
  select(study_id, event, viral_load_group, all_of(nums)) |>
  pivot_longer(
    cols = -c(event, study_id, viral_load_group),
    names_to = "key",
    values_to = "value"
  )

dta_nums$viral_load_group <- factor(dta_nums$viral_load_group)

levels(dta_nums$viral_load_group)

x <- unique(dta_nums$key)
xx <- which(x %in% c(
  "last_menstrual_period_days", "aspirin_dosage", "consent_date",
  "interview_start_time", "art_duration_weeks", "age_began_regular_drinking",
  "red_wine_units", "white_wine_units", "log_crp", "log_ggt", "log_viral_load",
  "log_acr_ratio", "log_microalbuminuria", "log_ten_year_cvd_risk"
))
x <- x[-xx] # drop variables not tested here

t_test_results <- vector("list", length(x))

for (i in 1:length(x)) {
  tmp <- dta_nums |>
    filter(key == x[i])
  t_test <- t.test(value ~ viral_load_group, data = tmp)
  results <- tidy(t_test)
  results$variable <- x[i]
  t_test_results[[i]] <- results
}

selected_columns <- c(
  "variable", "estimate", "p.value", "conf.low", "conf.high"
)

t_results_df <- bind_rows(t_test_results) |>
  select(all_of(selected_columns)) |>
  arrange(p.value)

datatable(t_results_df) |>
  formatStyle(
    columns = c("variable", "estimate", "p.value", "conf.low", "conf.high"),
    fontSize = "12px",
    color = "black",
    backgroundColor = "white"
  )

# Significant variables from the table above, entered by hand
sig_viral_load <- c(
  "crve", "crae_crve_ratio", "average_arterial_tree_diameter",
  "arteriole_branch_angle", "venule_mean_angular_asymmetry",
  "skeleton_fractal_analysis", "haemoglobin", "bmi", "hdl", "crp",
  "arteriole_branch_smaller_branch", "arteriole_branch_larger_branch",
  "branch_segment_tortuosity_kurtosis", "venule_branch_smaller_branch"
)

sig_viral_load_ret <- c(
  "crve", "crae_crve_ratio", "average_arterial_tree_diameter",
  "arteriole_branch_angle", "venule_mean_angular_asymmetry",
  "skeleton_fractal_analysis", "arteriole_branch_smaller_branch",
  "arteriole_branch_larger_branch", "branch_segment_tortuosity_kurtosis",
  "venule_branch_smaller_branch"
)
```

### CD4 count (clinic)

```{r}
#| label: ttest-cd4-clinic
# Welch t-tests of numeric variables by clinic CD4 count (<500 vs 500+)
dta_nums <- dta |>
  select(study_id, event, cd_count_clinic_group_2, all_of(nums)) |>
  pivot_longer(
    cols = -c(event, study_id, cd_count_clinic_group_2),
    names_to = "key",
    values_to = "value"
  )

dta_nums$cd_count_clinic_group_2 <- factor(dta_nums$cd_count_clinic_group_2)

levels(dta_nums$cd_count_clinic_group_2)

x <- unique(dta_nums$key)
xx <- which(x %in% c(
  "last_menstrual_period_days", "aspirin_dosage", "consent_date",
  "interview_start_time", "art_duration_weeks", "cd4_count_clinic",
  "endo_result", "age_began_regular_drinking", "red_wine_units",
  "white_wine_units", "viral_load", "log_crp", "log_ggt", "log_viral_load",
  "log_acr_ratio", "log_microalbuminuria", "log_ten_year_cvd_risk"
))
x <- x[-xx] # drop variables not tested here

t_test_results <- vector("list", length(x))

for (i in 1:length(x)) {
  tmp <- dta_nums |>
    filter(key == x[i])
  t_test <- t.test(value ~ cd_count_clinic_group_2, data = tmp)
  results <- tidy(t_test)
  results$variable <- x[i]
  t_test_results[[i]] <- results
}

selected_columns <- c(
  "variable", "estimate", "p.value", "conf.low", "conf.high"
)

t_results_df <- bind_rows(t_test_results) |>
  select(all_of(selected_columns)) |>
  arrange(p.value)

datatable(t_results_df) |>
  formatStyle(
    columns = c("variable", "estimate", "p.value", "conf.low", "conf.high"),
    fontSize = "12px",
    color = "black",
    backgroundColor = "white"
  )

# Significant variables from the table above, entered by hand
sig_clinic_cd_count <- c(
  "crve", "branch_segment_length_kurtosis", "arteriole_branch_larger_branch",
  "arteriole_branch_optimality", "skeleton_fractal_analysis",
  "arteriole_branch_smaller_branch", "branch_segment_length_average", "ldl",
  "crp"
)

sig_clinic_cd_count_ret <- c(
  "crve", "branch_segment_length_kurtosis", "arteriole_branch_larger_branch",
  "arteriole_branch_optimality", "skeleton_fractal_analysis",
  "arteriole_branch_smaller_branch", "branch_segment_length_average"
)
```

### CD4 count (EndoAfrica)

```{r}
#| label: ttest-cd4-endo
# Welch t-tests of numeric variables by EndoAfrica CD4 count (<500 vs 500+)
dta_nums <- dta |>
  select(study_id, event, cd4_count_endo_group_2, all_of(nums)) |>
  pivot_longer(
    cols = -c(event, study_id, cd4_count_endo_group_2),
    names_to = "key",
    values_to = "value"
  )

dta_nums$cd4_count_endo_group_2 <- factor(dta_nums$cd4_count_endo_group_2)

levels(dta_nums$cd4_count_endo_group_2)

x <- unique(dta_nums$key)
xx <- which(x %in% c(
  "last_menstrual_period_days", "aspirin_dosage", "consent_date",
  "interview_start_time", "art_duration_weeks", "cd4_count_clinic",
  "endo_result", "age_began_regular_drinking", "red_wine_units",
  "white_wine_units", "viral_load", "log_crp", "log_ggt", "log_viral_load",
  "log_acr_ratio", "log_microalbuminuria", "log_ten_year_cvd_risk"
))
x <- x[-xx] # drop variables not tested here

t_test_results <- vector("list", length(x))

for (i in 1:length(x)) {
  tmp <- dta_nums |>
    filter(key == x[i])
  t_test <- t.test(value ~ cd4_count_endo_group_2, data = tmp)
  results <- tidy(t_test)
  results$variable <- x[i]
  t_test_results[[i]] <- results
}

selected_columns <- c(
  "variable", "estimate", "p.value", "conf.low", "conf.high"
)

t_results_df <- bind_rows(t_test_results) |>
  select(all_of(selected_columns)) |>
  arrange(p.value)

datatable(t_results_df) |>
  formatStyle(
    columns = c("variable", "estimate", "p.value", "conf.low", "conf.high"),
    fontSize = "12px",
    color = "black",
    backgroundColor = "white"
  )

# Significant variables from the table above, entered by hand
sig_endo_result <- c(
  "arteriole_branch_larger_branch", "arteriole_branch_optimality",
  "arteriole_branch_smaller_branch", "ggt", "triglycerides", "crve", "bmi",
  "crp", "average_arterial_tree_diameter"
)

sig_endo_result_ret <- c(
  "arteriole_branch_larger_branch", "arteriole_branch_optimality",
  "arteriole_branch_smaller_branch", "crve", "average_arterial_tree_diameter"
)
```

### Alcohol in the last 12 months

```{r}
#| label: ttest-alcohol
# Welch t-tests of numeric variables by alcohol use in the last 12 months
dta_nums <- dta |>
  select(study_id, event, alcohol_last_12_months, all_of(nums)) |>
  pivot_longer(
    cols = -c(event, study_id, alcohol_last_12_months),
    names_to = "key",
    values_to = "value"
  )

dta_nums$alcohol_last_12_months <- factor(dta_nums$alcohol_last_12_months)

levels(dta_nums$alcohol_last_12_months)

x <- unique(dta_nums$key)
xx <- which(x %in% c(
  "last_menstrual_period_days", "aspirin_dosage", "consent_date",
  "interview_start_time", "art_duration_weeks", "age_began_regular_drinking",
  "red_wine_units", "white_wine_units", "cd4_count_clinic",
  "cd4_count_endo_africa ", "viral_load", "log_crp", "log_ggt",
  "log_viral_load", "log_acr_ratio", "log_microalbuminuria",
  "log_ten_year_cvd_risk"
))
x <- x[-xx] # drop variables not tested here

t_test_results <- vector("list", length(x))

for (i in 1:length(x)) {
  tmp <- dta_nums |>
    filter(key == x[i])
  t_test <- t.test(value ~ alcohol_last_12_months, data = tmp)
  results <- tidy(t_test)
  results$variable <- x[i]
  t_test_results[[i]] <- results
}

selected_columns <- c(
  "variable", "estimate", "p.value", "conf.low", "conf.high"
)

t_results_df <- bind_rows(t_test_results) |>
  select(all_of(selected_columns)) |>
  arrange(p.value)

datatable(t_results_df) |>
  formatStyle(
    columns = c("variable", "estimate", "p.value", "conf.low", "conf.high"),
    fontSize = "12px",
    color = "black",
    backgroundColor = "white"
  )

sig_alcohol <- c("crae_crve_ratio", "arteriole_branch_angle", "ggt")
sig_alcohol_ret <- c("crae_crve_ratio", "arteriole_branch_angle")
```

### Hypertension

```{r}
#| label: ttest-hypertension
# Welch t-tests of numeric variables by hypertension
dta_nums <- dta |>
  select(study_id, event, hypertension, all_of(nums)) |>
  filter(hypertension != "Unknown") |>
  pivot_longer(
    cols = -c(event, study_id, hypertension),
    names_to = "key",
    values_to = "value"
  )

dta_nums$hypertension <- factor(dta_nums$hypertension)

levels(dta_nums$hypertension)

x <- unique(dta_nums$key)
xx <- which(x %in% c(
  "last_menstrual_period_days", "aspirin_dosage", "consent_date",
  "interview_start_time", "art_duration_weeks", "age_began_regular_drinking",
  "red_wine_units", "white_wine_units", "cd4_count_clinic",
  "cd4_count_endo_africa ", "viral_load", "log_crp", "log_ggt",
  "log_viral_load", "log_acr_ratio", "log_microalbuminuria",
  "log_ten_year_cvd_risk"
))
x <- x[-xx] # drop variables not tested here

t_test_results <- vector("list", length(x))

for (i in 1:length(x)) {
  tmp <- dta_nums |>
    filter(key == x[i])
  t_test <- t.test(value ~ hypertension, data = tmp)
  results <- tidy(t_test)
  results$variable <- x[i]
  t_test_results[[i]] <- results
}

selected_columns <- c(
  "variable", "estimate", "p.value", "conf.low", "conf.high"
)

t_results_df <- bind_rows(t_test_results) |>
  select(all_of(selected_columns)) |>
  arrange(p.value)

datatable(t_results_df) |>
  formatStyle(
    columns = c("variable", "estimate", "p.value", "conf.low", "conf.high"),
    fontSize = "12px",
    color = "black",
    backgroundColor = "white"
  )

# Significant variables from the table above, entered by hand
sig_hypertension <- c(
  "vascular_age", "age", "mean_systolic_pressure", "ten_year_cvd_risk",
  "mean_diastolic_pressure", "ckd_epi_egfr", "bmi", "crae_crve_ratio", "crae",
  "mdrd_egfr", "waist_hip_ratio", "fourier_fractal_enhanced",
  "total_cholesterol", "average_carotid_imt", "hba1c", "fasting_glucose", "ldl",
  "urine_creatinine", "triglycerides"
)

sig_hypertension_ret <- c("crae_crve_ratio", "crae", "fourier_fractal_enhanced")
```

### Smoking

```{r}
#| label: ttest-smoking
# Welch t-tests of numeric variables by smoking status (yes/no)
dta_nums <- dta |>
  select(study_id, event, smoker_new, all_of(nums)) |>
  pivot_longer(
    cols = -c(event, study_id, smoker_new),
    names_to = "key",
    values_to = "value"
  )
dta_nums$smoker_new <- factor(dta_nums$smoker_new)

levels(dta_nums$smoker_new)

x <- unique(dta_nums$key)
xx <- which(x %in% c(
  "last_menstrual_period_days", "aspirin_dosage", "consent_date",
  "interview_start_time", "art_duration_weeks", "age_began_regular_drinking",
  "red_wine_units", "white_wine_units", "cd4_count_clinic",
  "cd4_count_endo_africa ", "viral_load", "log_crp", "log_ggt",
  "log_viral_load", "log_acr_ratio", "log_microalbuminuria",
  "log_ten_year_cvd_risk"
))
x <- x[-xx] # drop variables not tested here

t_test_results <- vector("list", length(x))

for (i in 1:length(x)) {
  tmp <- dta_nums |>
    filter(key == x[i])
  t_test <- t.test(value ~ smoker_new, data = tmp)
  results <- tidy(t_test)
  results$variable <- x[i]
  t_test_results[[i]] <- results
}

selected_columns <- c(
  "variable", "estimate", "p.value", "conf.low", "conf.high"
)

t_results_df <- bind_rows(t_test_results) |>
  select(all_of(selected_columns)) |>
  arrange(p.value)

datatable(t_results_df) |>
  formatStyle(
    columns = c("variable", "estimate", "p.value", "conf.low", "conf.high"),
    fontSize = "12px",
    color = "black",
    backgroundColor = "white"
  )

# Significant variables from the table above, entered by hand
sig_smoker <- c(
  "bmi", "ten_year_cvd_risk", "crve", "crae", "vascular_age",
  "skeleton_fractal_analysis", "mean_diastolic_pressure",
  "segmented_fractal_analysis"
)

sig_smoker_ret <- c(
  "crve", "crae", "skeleton_fractal_analysis", "segmented_fractal_analysis"
)
```

:::

## Pearson correlations

Pearson correlations between clinical and retinal variables, and among the retinal measures.

::: panel-tabset

### Clinical vs retinal

```{r}
#| label: pearson-clinical-retinal
# Pearson correlation for every clinical-retinal pair
# Note: this redefines clinical_var for the rest of the document
clinical_var <- c(
  "total_cholesterol", "hdl", "ldl", "triglycerides", "fasting_glucose",
  "hba1c", "crp", "ggt", "haemoglobin", "cd4_count_clinic", "viral_load",
  "creatinine", "mdrd_egfr", "urine_creatinine", "acr_ratio",
  "microalbuminuria", "ckd_epi_egfr", "diameter_baseline", "fmd_percentage",
  "average_carotid_imt", "age", "bmi", "waist_hip_ratio",
  "mean_systolic_pressure", "mean_diastolic_pressure", "mean_heart_rate"
)

combined_df <- expand.grid(x = clinical_var, y = retinal_var)

results_list <- vector("list", nrow(combined_df))

dta_nums <- dta |>
  select(study_id, event, hiv_pos, all_of(nums)) |>
  pivot_longer(
    cols = -c(event, study_id, hiv_pos),
    names_to = "key",
    values_to = "value"
  )

for (i in 1:nrow(combined_df)) {
  tmp_x <- dta_nums |>
    filter(key == combined_df$x[i]) |>
    select(study_id, x = value)
  tmp_y <- dta_nums |>
    filter(key == combined_df$y[i]) |>
    select(study_id, y = value)
  tmp <- tmp_x |>
    inner_join(tmp_y, by = "study_id")
  pearson_test <- suppressWarnings(cor.test(
    x = tmp$x,
    y = tmp$y,
    method = "pearson"
  ))
  results_list[[i]] <- tidy(pearson_test) |>
    mutate(x = combined_df$x[i], y = combined_df$y[i])
}

selected_columns <- c("x", "y", "p.value", "estimate", "statistic")

combined_corr_results_pearson <- bind_rows(results_list) |>
  select(all_of(selected_columns)) |>
  arrange(p.value, estimate)

datatable(combined_corr_results_pearson) |>
  formatStyle(
    columns = c("x", "y", "p.value", "estimate", "statistic"),
    fontSize = "12px",
    color = "black",
    backgroundColor = "white"
  )
```

### Retinal vs retinal

```{r}
#| label: pearson-retinal-retinal
# Pearson correlation for every pair of retinal measures
retinal_df <- expand.grid(x = retinal_var, y = retinal_var)

retinal_results_list <- vector("list", nrow(retinal_df))

for (i in 1:nrow(retinal_df)) {
  tmp_x <- dta_nums |>
    filter(key == retinal_df$x[i]) |>
    select(study_id, x = value)
  tmp_y <- dta_nums |>
    filter(key == retinal_df$y[i]) |>
    select(study_id, y = value)
  tmp <- tmp_x |>
    inner_join(tmp_y, by = "study_id")
  pearson_test <- suppressWarnings(cor.test(
    x = tmp$x,
    y = tmp$y,
    method = "pearson"
  ))
  retinal_results_list[[i]] <- tidy(pearson_test) |>
    mutate(x = retinal_df$x[i], y = retinal_df$y[i])
}

selected_columns <- c("x", "y", "p.value", "estimate", "statistic")

retinal_corr_results_pearson <- bind_rows(retinal_results_list) |>
  select(all_of(selected_columns)) |>
  arrange(x, estimate)

datatable(retinal_corr_results_pearson) |>
  formatStyle(
    columns = c("x", "y", "p.value", "estimate", "statistic"),
    fontSize = "12px",
    color = "black",
    backgroundColor = "white"
  )
```

:::

## Correlation heatmaps

Heatmaps of the Pearson correlation coefficients from the previous section.

::: panel-tabset

### Clinical vs retinal

```{r}
#| label: corr-heatmap-clinical-retinal
# Heatmap of the correlation coefficients
clinical_plot <- ggplot(
  combined_corr_results_pearson,
  aes(x = reorder(x, estimate), y = reorder(y, estimate), fill = estimate)
) +
  geom_tile(color = "white") +
  scale_fill_gradient2(
    low = "blue", mid = "white", high = "red", midpoint = 0,
    name = "Pearson's Correlation"
  ) +
  theme_minimal() +
  labs(
    title = "Pearson's Correlation Heatmap",
    x = "Clinical Variables",
    y = "Retinal Variables"
  ) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, size = 12),
    axis.text.y = element_text(vjust = 1, size = 12),
    axis.title.x = element_text(size = 14),
    axis.title.y = element_text(size = 14)
  ) +
  coord_fixed(ratio = 1)

clinical_plot
ggsave("correlation_heatmap_1.svg", plot = clinical_plot)
```

### Retinal vs retinal

```{r}
#| label: corr-heatmap-retinal-retinal
# Heatmap of the correlation coefficients
correlation_plot <- ggplot(
  retinal_corr_results_pearson,
  aes(x = reorder(x, estimate), y = reorder(y, estimate), fill = estimate)
) +
  geom_tile(color = "white") +
  scale_fill_gradient2(
    low = "blue", mid = "white", high = "red", midpoint = 0,
    name = "Pearson's Correlation"
  ) +
  theme_minimal() +
  labs(
    title = "Pearson's Correlation Heatmap",
    x = "Clinical Variables",
    y = "Retinal Variables"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
    axis.text.y = element_text(vjust = 1, size = 12),
    axis.title.x = element_text(size = 14),
    axis.title.y = element_text(size = 14)
  )

ggsave(
  "correlation_heatmap_2.svg",
  plot = correlation_plot,
  width = 10,
  height = 8
)
```

:::

## Standardised heatmaps

Retinal measures converted to z-scores and plotted per participant.

::: panel-tabset

### Whole sample

```{r}
#| label: zscore-heatmap-all
# Z-scores of each retinal measure, one row per participant
dta_standard <- dta |>
  select(study_id, all_of(retinal_var)) |>
  mutate(across(all_of(retinal_var), scale)) |>
  pivot_longer(
    cols = -study_id,
    names_to = "retinal",
    values_to = "value"
  )

ggplot(dta_standard, aes(x = retinal, y = study_id, fill = value[, ])) +
  geom_tile() +
  scale_fill_gradient2(
    low = "blue",
    mid = "white",
    high = "blue",
    midpoint = 0,
    limits = c(-3, 3)
  ) +
  labs(
    title = "standardised retinal parameters",
    x = "retinal parameters",
    y = "study_id",
    fill = "value"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### HIV status

```{r}
#| label: zscore-heatmap-hiv
# Z-scores of each retinal measure, one row per participant
dta_standard <- dta |>
  select(study_id, hiv_pos, all_of(retinal_var)) |>
  mutate(across(all_of(retinal_var), scale)) |>
  pivot_longer(
    cols = all_of(retinal_var),
    names_to = "retinal",
    values_to = "value"
  )

ggplot(
  dta_standard,
  aes(x = retinal, y = reorder(study_id, hiv_pos), fill = value[, ])
) +
  geom_tile() +
  scale_fill_viridis_c(limit = c(-3, 3)) +
  labs(
    title = "Standardized Retinal Parameters",
    x = "Retinal Parameters",
    y = "Study ID",
    fill = "Standardized Value"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

:::

## Clustered heatmaps

Z-scored retinal measures with hierarchical clustering (Euclidean distance, complete linkage) and group annotations, drawn with ComplexHeatmap.

::: panel-tabset

### HIV status

```{r}
#| label: cluster-hiv
# Z-scored retinal measures with hierarchical clustering
dta_clean <- dta |>
  select(study_id, hiv_pos, all_of(retinal_var)) |>
  mutate(across(all_of(retinal_var), scale))

# Study IDs become row names; hiv_pos is kept out of the matrix and used
# for the row annotation instead
retinal_matrix <- dta_clean |>
  column_to_rownames(var = "study_id") |>
  select(-hiv_pos) |>
  as.matrix()

# Row annotation showing HIV status
anno_hiv_pos <- rowAnnotation(
  hiv_pos = anno_simple(dta_clean$hiv_pos,
    col = c("Yes" = "white", "No" = "black")
  )
)

# Rows: participants clustered by their z-scored retinal profile.
# Columns: retinal measures clustered by similarity.
heatmap <- Heatmap(
  retinal_matrix,
  name = "Retinal Parameters",
  row_names_side = "left",
  column_names_side = "bottom",
  cluster_rows = TRUE,
  cluster_columns = TRUE,
  show_row_names = TRUE,
  show_column_names = TRUE,
  right_annotation = anno_hiv_pos,
  col = colorRamp2(c(-3, 0, 3), c("blue", "white", "black")),
  clustering_distance_rows = "euclidean",
  clustering_method_rows = "complete",
  clustering_distance_columns = "euclidean",
  clustering_method_columns = "complete"
)

draw(heatmap)

png("hierarchical_1.png", width = 10, height = 8, units = "in", res = 300)
draw(heatmap)
dev.off() # Close the PNG device
```

### ART status

```{r}
#| label: cluster-art
dta_clean <- dta |>
  filter(!is.na(on_art)) |>
  select(study_id, on_art, all_of(retinal_var)) |>
  mutate(across(all_of(retinal_var), scale))

retinal_matrix <- dta_clean |>
  column_to_rownames(var = "study_id") |>
  select(-on_art) |>
  as.matrix()

anno_on_art <- rowAnnotation(
  on_art = anno_simple(dta_clean$on_art,
    col = c("Yes" = "green", "No" = "purple")
  )
)

heatmap <- Heatmap(
  retinal_matrix,
  name = "Retinal Parameters",
  row_names_side = "left",
  column_names_side = "bottom",
  cluster_rows = TRUE,
  cluster_columns = TRUE,
  show_row_names = TRUE,
  show_column_names = TRUE,
  right_annotation = anno_on_art,
  col = colorRamp2(c(-3, 0, 3), c("blue", "white", "red")),
  clustering_distance_rows = "euclidean",
  clustering_method_rows = "complete",
  clustering_distance_columns = "euclidean",
  clustering_method_columns = "complete"
)

draw(heatmap)
```

### Gender

```{r}
#| label: cluster-gender
dta_standard <- dta |>
  select(study_id, gender, all_of(retinal_var)) |>
  mutate(across(all_of(retinal_var), scale))

retinal_matrix <- dta_standard |>
  column_to_rownames(var = "study_id") |>
  select(-gender) |>
  as.matrix()

anno_hiv_pos <- rowAnnotation(
  hiv_pos = anno_simple(dta_standard$gender,
    col = c("Female" = "red", "Male" = "blue")
  )
)

heatmap <- Heatmap(
  retinal_matrix,
  name = "Retinal Parameters",
  row_names_side = "left",
  column_names_side = "bottom",
  cluster_rows = TRUE,
  cluster_columns = TRUE,
  show_row_names = TRUE,
  show_column_names = TRUE,
  right_annotation = anno_hiv_pos,
  col = colorRamp2(c(-3, 0, 3), c("green", "white", "purple")),
  clustering_distance_rows = "euclidean",
  clustering_method_rows = "complete",
  clustering_distance_columns = "euclidean",
  clustering_method_columns = "complete"
)

draw(heatmap)

# Euclidean distance between participants; complete linkage merges clusters
# by their largest pairwise distance
```

### HIV status, ART status and gender

```{r}
#| label: cluster-hiv-art-gender
dta_clean <- dta |>
  select(study_id, hiv_pos, gender, on_art, all_of(retinal_var)) |>
  mutate(across(all_of(retinal_var), scale)) |>
  column_to_rownames(var = "study_id")

retinal_matrix <- dta_clean |>
  select(-hiv_pos, -gender, -on_art) |>
  as.matrix()

anno_hiv_pos <- anno_simple(dta_clean$hiv_pos,
  col = c("Yes" = "red", "No" = "blue"),
  which = "row"
)
anno_gender <- anno_simple(dta_clean$gender,
  col = c("Female" = "pink", "Male" = "green"),
  which = "row"
)
anno_on_art <- anno_simple(dta_clean$on_art,
  col = c("Yes" = "purple", "No" = "black", "NA" = "white"),
  which = "row"
)

combined_annotation <- rowAnnotation(
  hiv_pos = anno_hiv_pos,
  sex = anno_gender,
  on_art = anno_on_art
)

heatmap <- Heatmap(
  retinal_matrix,
  name = "Retinal Parameters",
  row_names_side = "left",
  column_names_side = "bottom",
  cluster_rows = TRUE,
  cluster_columns = TRUE,
  show_row_names = TRUE,
  show_column_names = TRUE,
  right_annotation = combined_annotation,
  col = colorRamp2(c(-3, 0, 3), c("blue", "white", "red")),
  clustering_distance_rows = "euclidean",
  clustering_method_rows = "complete",
  clustering_distance_columns = "euclidean",
  clustering_method_columns = "complete"
)

draw(heatmap)

# Save the heatmap as a high-resolution PNG
png("hierarchical_1.png", units = "in", res = 300)
draw(heatmap)
dev.off() # Close the PNG device

heatmap <- Heatmap(
  retinal_matrix,
  name = "Retinal Parameters",
  row_names_side = "left",
  column_names_side = "bottom",
  cluster_rows = FALSE,
  cluster_columns = TRUE,
  show_row_names = TRUE,
  show_column_names = TRUE,
  right_annotation = combined_annotation,
  col = colorRamp2(c(-3, 0, 3), c("blue", "white", "red")),
  clustering_distance_rows = "euclidean",
  clustering_method_rows = "complete",
  clustering_distance_columns = "euclidean",
  clustering_method_columns = "complete"
)

draw(heatmap)
```

### HIV status: t-test significant variables

```{r}
#| label: cluster-hiv-sig
# Variables that differed by HIV status in the t-tests, entered by hand
t_sig_var_hiv <- c(
  "mdrd_egfr", "ggt", "ckd_epi_egfr", "bmi", "crve", "urine_creatinine",
  "haemoglobin", "acr_ratio", "creatinine", "venule_branch_larger_branch",
  "venule_branch_parent_width", "skeleton_fractal_analysis", "waist_hip_ratio"
)

dta_clean <- dta |>
  select(
    study_id, hiv_pos, gender, on_art, cvd_risk_factor, all_of(t_sig_var_hiv)
  ) |>
  mutate(across(all_of(t_sig_var_hiv), scale)) |>
  column_to_rownames(var = "study_id")

dta_clean <- dta_clean |>
  arrange(hiv_pos, on_art, gender, cvd_risk_factor) |>
  as.data.frame()

retinal_matrix <- dta_clean |>
  select(-hiv_pos, -gender, -on_art, -cvd_risk_factor) |>
  as.matrix()

anno_hiv_pos <- anno_simple(dta_clean$hiv_pos,
  col = c("Yes" = "red", "No" = "blue"),
  which = "row"
)
anno_cvd <- anno_simple(dta_clean$cvd_risk_factor,
  col = c("Yes" = "pink", "No" = "black"),
  which = "row"
)
anno_gender <- anno_simple(dta_clean$gender,
  col = c("Female" = "yellow", "Male" = "green"),
  which = "row"
)
anno_on_art <- anno_simple(dta_clean$on_art,
  col = c("Yes" = "purple", "No" = "orange", "NA" = "black"),
  which = "row"
)

combined_annotation <- rowAnnotation(
  hiv_pos = anno_hiv_pos,
  cvd_risk_factor = anno_cvd,
  gender = anno_gender,
  on_art = anno_on_art
)

heatmap <- Heatmap(
  retinal_matrix,
  name = "Retinal Parameters",
  row_names_side = "left",
  column_names_side = "bottom",
  cluster_rows = TRUE,
  cluster_columns = TRUE,
  show_row_names = TRUE,
  show_column_names = TRUE,
  left_annotation = combined_annotation,
  col = colorRamp2(c(-3, 0, 3), c("blue", "white", "red")),
  clustering_distance_rows = "euclidean",
  clustering_method_rows = "complete",
  clustering_distance_columns = "euclidean",
  clustering_method_columns = "complete"
)

draw(heatmap)

# Save the heatmap as a high-resolution PNG
png("hierarchical_2.png", width = 10, height = 8, units = "in", res = 300)
draw(heatmap)
dev.off() # Close the PNG device

heatmap <- Heatmap(
  retinal_matrix,
  name = "Retinal Parameters",
  row_names_side = "left",
  column_names_side = "bottom",
  cluster_rows = FALSE,
  cluster_columns = TRUE,
  show_row_names = TRUE,
  show_column_names = TRUE,
  left_annotation = combined_annotation,
  col = colorRamp2(c(-3, 0, 3), c("blue", "white", "red")),
  clustering_distance_rows = "euclidean",
  clustering_method_rows = "complete",
  clustering_distance_columns = "euclidean",
  clustering_method_columns = "complete"
)

draw(heatmap)
```

### ART status: t-test significant variables

```{r}
#| label: cluster-art-sig
# Variables that differed by ART status in the t-tests, entered by hand
t_sig_var_art <- c(
  "fourier_fractal_enhanced", "branch_segment_length_kurtosis", "ggt",
  "total_cholesterol", "ten_year_cvd_risk", "age", "segmented_fractal_analysis",
  "hdl", "acr_ratio", "vascular_age", "ldl", "venule_mean_angular_asymmetry"
)

dta_clean <- dta |>
  select(
    study_id, hiv_pos, gender, on_art, cvd_risk_factor, all_of(t_sig_var_art)
  ) |>
  mutate(across(all_of(t_sig_var_art), scale)) |>
  column_to_rownames(var = "study_id")

dta_clean <- dta_clean |>
  arrange(hiv_pos, on_art, gender, cvd_risk_factor) |>
  as.data.frame()

retinal_matrix <- dta_clean |>
  select(-hiv_pos, -gender, -on_art, -cvd_risk_factor) |>
  as.matrix()

anno_hiv_pos <- anno_simple(dta_clean$hiv_pos,
  col = c("Yes" = "red", "No" = "blue"),
  which = "row"
)
anno_cvd <- anno_simple(dta_clean$cvd_risk_factor,
  col = c("Yes" = "pink", "No" = "black"),
  which = "row"
)
anno_gender <- anno_simple(dta_clean$gender,
  col = c("Female" = "yellow", "Male" = "green"),
  which = "row"
)
anno_on_art <- anno_simple(dta_clean$on_art,
  col = c("Yes" = "purple", "No" = "orange", "NA" = "black"),
  which = "row"
)

combined_annotation <- rowAnnotation(
  hiv_pos = anno_hiv_pos,
  cvd_risk_factor = anno_cvd,
  gender = anno_gender,
  on_art = anno_on_art
)

heatmap <- Heatmap(
  retinal_matrix,
  name = "Retinal Parameters",
  row_names_side = "left",
  column_names_side = "bottom",
  cluster_rows = FALSE,
  cluster_columns = TRUE,
  show_row_names = TRUE,
  show_column_names = TRUE,
  left_annotation = combined_annotation,
  col = colorRamp2(c(-3, 0, 3), c("blue", "white", "red")),
  clustering_distance_rows = "euclidean",
  clustering_method_rows = "complete",
  clustering_distance_columns = "euclidean",
  clustering_method_columns = "complete"
)

draw(heatmap)

heatmap <- Heatmap(
  retinal_matrix,
  name = "Retinal Parameters",
  row_names_side = "left",
  column_names_side = "bottom",
  cluster_rows = TRUE,
  cluster_columns = TRUE,
  show_row_names = TRUE,
  show_column_names = TRUE,
  left_annotation = combined_annotation,
  col = colorRamp2(c(-3, 0, 3), c("blue", "white", "red")),
  clustering_distance_rows = "euclidean",
  clustering_method_rows = "complete",
  clustering_distance_columns = "euclidean",
  clustering_method_columns = "complete"
)

draw(heatmap)
```

:::

## Principal component analysis

PCA with base R `prcomp()` (centred and scaled) on retinal, clinical and combined variable sets. Points are coloured by group, and PC1 is compared between groups with t-tests and Wilcoxon tests.

::: panel-tabset

### HIV and ART status: retinal

```{r}
#| label: pca-hiv-art-retinal
# PCA (centred and scaled) on participants with complete data, coloured by
# group. PC1 is then compared between groups.
pca_dta <- dta |>
  select(hiv_pos, on_art, all_of(retinal_var))

complete_rows <- complete.cases(pca_dta |>
  select(all_of(retinal_var)))

filtered_pca_dta <- pca_dta[complete_rows, ]

numeric_data <- filtered_pca_dta |>
  select(all_of(retinal_var))

metadata <- filtered_pca_dta |>
  select(on_art, hiv_pos)

metadata <- metadata |>
  mutate(on_art = ifelse(is.na(on_art), "HIV negative", on_art))

pca_result <- prcomp(numeric_data, center = TRUE, scale. = TRUE)
pca_scores <- as.data.frame(pca_result$x)
pca_meta <- cbind(pca_scores, metadata)

pca_plot <- ggplot(pca_meta, aes(x = PC1, y = PC2, color = on_art)) +
  geom_point(size = 4) +
  labs(
    title = "PCA: PC1 vs PC2",
    x = "Principal Component 1",
    y = "Principal Component 2"
  ) +
  scale_color_manual(values = c(
    "Yes" = "red",
    "No" = "blue",
    "HIV negative" = "black"
  )) +
  theme_minimal()

ggsave("pca_all_ret.svg", plot = pca_plot, width = 10, height = 8)

ggplot(pca_meta, aes(x = PC3, y = PC4, color = on_art, shape = hiv_pos)) +
  geom_point() +
  labs(
    title = "PCA: PC3 vs PC4 ",
    x = "Principal Component 3",
    y = "Principal Component 4"
  ) +
  scale_color_manual(values = c(
    "Yes" = "red",
    "No" = "blue",
    "HIV negative" = "black"
  )) +
  theme_minimal()

pca_result

sdev <- pca_result$sdev
variance <- sdev^2
variance_explained <- variance / sum(variance)
print(variance_explained)

screeplot_df <- data.frame(
  PC = seq_along(pca_result$sdev),
  Variance = pca_result$sdev^2,
  Proportion = pca_result$sdev^2 / sum(pca_result$sdev^2)
)

ggplot(screeplot_df, aes(x = PC, y = Proportion)) +
  geom_point() +
  geom_line() +
  labs(
    title = "Scree Plot",
    x = "Principal Component",
    y = "Proportion of Variance Explained"
  ) +
  theme_minimal()

t_test_result <- t.test(PC1 ~ hiv_pos, data = pca_meta, var.equal = TRUE)

print(t_test_result)

wilcox_test_result <- wilcox.test(PC1 ~ hiv_pos, data = pca_meta)

print(wilcox_test_result)

pca_meta <- pca_meta |>
  filter(on_art != "HIV negative")

t_test_result <- t.test(PC1 ~ on_art, data = pca_meta, var.equal = TRUE)

print(t_test_result)

wilcox_test_result <- wilcox.test(PC1 ~ on_art, data = pca_meta)

print(wilcox_test_result)
```

### ART duration: retinal

```{r}
#| label: pca-art-duration-retinal
pca_dta <- dta |>
  select(art_duration_group, all_of(retinal_var))

complete_rows <- complete.cases(pca_dta |>
  select(all_of(retinal_var)))

filtered_pca_dta <- pca_dta[complete_rows, ]

numeric_data <- filtered_pca_dta |>
  select(all_of(retinal_var))

metadata <- filtered_pca_dta |>
  select(art_duration_group)

metadata <- metadata |>
  mutate(art_duration_group = factor(
    art_duration_group,
    levels = c(
      "Less than a year", "Not on ART or HIV neg", "1-5 years",
      "More than 5 years"
    ),
    labels = c(
      "Less than a year", "Not on ART or HIV neg", "1-5 years",
      "More than 5 years"
    )
  ))

pca_result <- prcomp(numeric_data, center = TRUE, scale. = TRUE)
pca_scores <- as.data.frame(pca_result$x)
pca_meta <- cbind(pca_scores, metadata)

ggplot(
  pca_meta,
  aes(x = PC1, y = PC2, color = art_duration_group, shape = art_duration_group)
) +
  geom_point() +
  labs(
    title = "PCA: PC1 vs PC2",
    x = "Principal Component 1",
    y = "Principal Component 2"
  ) +
  scale_color_manual(values = color_map_list) +
  theme_minimal()

ggplot(
  pca_meta,
  aes(x = PC3, y = PC4, color = art_duration_group, shape = art_duration_group)
) +
  geom_point() +
  labs(
    title = "PCA: PC3 vs PC4",
    x = "Principal Component 3",
    y = "Principal Component 4"
  ) +
  scale_color_manual(values = color_map_list) +
  theme_minimal()

pca_result

sdev <- pca_result$sdev
variance <- sdev^2
variance_explained <- variance / sum(variance)
print(variance_explained)

screeplot_df <- data.frame(
  PC = seq_along(pca_result$sdev),
  Variance = pca_result$sdev^2,
  Proportion = pca_result$sdev^2 / sum(pca_result$sdev^2)
)

ggplot(screeplot_df, aes(x = PC, y = Proportion)) +
  geom_point() +
  geom_line() +
  labs(
    title = "Scree Plot",
    x = "Principal Component",
    y = "Proportion of Variance Explained"
  ) +
  theme_minimal()

# Two ART duration groups (0-5 vs more than 5 years): compare PC1
pca_dta <- dta |>
  select(art_duration_group_2, all_of(retinal_var))

complete_rows <- complete.cases(pca_dta |>
  select(all_of(retinal_var)))

filtered_pca_dta <- pca_dta[complete_rows, ]

numeric_data <- filtered_pca_dta |>
  select(all_of(retinal_var))

metadata <- filtered_pca_dta |>
  select(art_duration_group_2)

metadata <- metadata |>
  mutate(art_duration_group_2 = factor(
    art_duration_group_2,
    levels = c("0-5 years", "More than 5 years"),
    labels = c("Less than 5 years", "More than 5 years")
  ))

pca_result <- prcomp(numeric_data, center = TRUE, scale. = TRUE)
pca_scores <- as.data.frame(pca_result$x)
pca_meta <- cbind(pca_scores, metadata)

t_test_result <- t.test(
  PC1 ~ art_duration_group_2,
  data = pca_meta,
  var.equal = TRUE
)

print(t_test_result)

wilcox_test_result <- wilcox.test(PC1 ~ art_duration_group_2, data = pca_meta)

print(wilcox_test_result)
```

### Viral load: retinal

```{r}
#| label: pca-viral-load-retinal
pca_dta <- dta |>
  select(viral_load_group, all_of(retinal_var))

complete_rows <- complete.cases(pca_dta |>
  select(all_of(retinal_var)))

filtered_pca_dta <- pca_dta[complete_rows, ]

numeric_data <- filtered_pca_dta |>
  select(all_of(retinal_var))

metadata <- filtered_pca_dta |>
  select(viral_load_group)

metadata <- metadata |>
  mutate(viral_load_group = factor(
    viral_load_group,
    levels = c("Viral load < 1000", "Viral load > 1000", "NA"),
    labels = c("Viral load < 1000", "Viral load > 1000", "NA"),
  ))

pca_result <- prcomp(numeric_data, center = TRUE, scale. = TRUE)
pca_scores <- as.data.frame(pca_result$x)
pca_meta <- cbind(pca_scores, metadata)

ggplot(
  pca_meta,
  aes(x = PC1, y = PC2, color = viral_load_group, shape = viral_load_group)
) +
  geom_point() +
  labs(
    title = "PCA: PC1 vs PC2",
    x = "Principal Component 1",
    y = "Principal Component 2"
  ) +
  scale_color_manual(values = color_map_list) +
  theme_minimal()

ggplot(
  pca_meta,
  aes(x = PC3, y = PC4, color = viral_load_group, shape = viral_load_group)
) +
  geom_point() +
  labs(
    title = "PCA: PC3 vs PC4",
    x = "Principal Component 3",
    y = "Principal Component 4"
  ) +
  scale_color_manual(values = color_map_list) +
  theme_minimal()

pca_result

sdev <- pca_result$sdev
variance <- sdev^2
variance_explained <- variance / sum(variance)
print(variance_explained)

screeplot_df <- data.frame(
  PC = seq_along(pca_result$sdev),
  Variance = pca_result$sdev^2,
  Proportion = pca_result$sdev^2 / sum(pca_result$sdev^2)
)

ggplot(screeplot_df, aes(x = PC, y = Proportion)) +
  geom_point() +
  geom_line() +
  labs(
    title = "Scree Plot",
    x = "Principal Component",
    y = "Proportion of Variance Explained"
  ) +
  theme_minimal()

t_test_result <- t.test(
  PC1 ~ viral_load_group,
  data = pca_meta,
  var.equal = TRUE
)

print(t_test_result)

wilcox_test_result <- wilcox.test(PC1 ~ viral_load_group, data = pca_meta)

print(wilcox_test_result)
```

### Age group and ART status: retinal

```{r}
#| label: pca-age-art-retinal
pca_dta <- dta |>
  select(age_group, on_art, all_of(retinal_var))

complete_rows <- complete.cases(pca_dta |>
  select(all_of(retinal_var)))

filtered_pca_dta <- pca_dta[complete_rows, ]

numeric_data <- filtered_pca_dta |>
  select(all_of(retinal_var))

metadata <- filtered_pca_dta |>
  select(age_group, on_art)

metadata <- metadata |>
  mutate(on_art = ifelse(is.na(on_art), "HIV negative", on_art))

pca_result <- prcomp(numeric_data, center = TRUE, scale. = TRUE)
pca_scores <- as.data.frame(pca_result$x)
pca_meta <- cbind(pca_scores, metadata)

ggplot(pca_meta, aes(x = PC1, y = PC2, color = age_group, shape = on_art)) +
  geom_point() +
  labs(
    title = "PCA: PC1 vs PC2",
    x = "Principal Component 1",
    y = "Principal Component 2"
  ) +
  theme_minimal()

ggplot(pca_meta, aes(x = PC3, y = PC4, color = age_group, shape = on_art)) +
  geom_point() +
  labs(
    title = "PCA: PC3 vs PC4 ",
    x = "Principal Component 3",
    y = "Principal Component 4"
  ) +
  theme_minimal()

pca_result

sdev <- pca_result$sdev
variance <- sdev^2
variance_explained <- variance / sum(variance)
print(variance_explained)

screeplot_df <- data.frame(
  PC = seq_along(pca_result$sdev),
  Variance = pca_result$sdev^2,
  Proportion = pca_result$sdev^2 / sum(pca_result$sdev^2)
)

ggplot(screeplot_df, aes(x = PC, y = Proportion)) +
  geom_point() +
  geom_line() +
  labs(
    title = "Scree Plot",
    x = "Principal Component",
    y = "Proportion of Variance Explained"
  ) +
  theme_minimal()
```

### Smoking and ART status: retinal

```{r}
#| label: pca-smoking-art-retinal
pca_dta <- dta |>
  select(smoker, on_art, all_of(retinal_var))

complete_rows <- complete.cases(pca_dta |>
  select(all_of(retinal_var)))

filtered_pca_dta <- pca_dta[complete_rows, ]

numeric_data <- filtered_pca_dta |>
  select(all_of(retinal_var))

metadata <- filtered_pca_dta |>
  select(smoker, on_art)

metadata <- metadata |>
  mutate(on_art = ifelse(is.na(on_art), "HIV negative", on_art))

pca_result <- prcomp(numeric_data, center = TRUE, scale. = TRUE)
pca_scores <- as.data.frame(pca_result$x)
pca_meta <- cbind(pca_scores, metadata)

ggplot(pca_meta, aes(x = PC1, y = PC2, color = smoker, shape = on_art)) +
  geom_point() +
  labs(
    title = "PCA: PC1 vs PC2",
    x = "Principal Component 1",
    y = "Principal Component 2"
  ) +
  theme_minimal()

ggplot(pca_meta, aes(x = PC3, y = PC4, color = smoker, shape = on_art)) +
  geom_point() +
  labs(
    title = "PCA: PC3 vs PC4 ",
    x = "Principal Component 3",
    y = "Principal Component 4"
  ) +
  theme_minimal()

pca_result

sdev <- pca_result$sdev
variance <- sdev^2
variance_explained <- variance / sum(variance)
print(variance_explained)

screeplot_df <- data.frame(
  PC = seq_along(pca_result$sdev),
  Variance = pca_result$sdev^2,
  Proportion = pca_result$sdev^2 / sum(pca_result$sdev^2)
)

ggplot(screeplot_df, aes(x = PC, y = Proportion)) +
  geom_point() +
  geom_line() +
  labs(
    title = "Scree Plot",
    x = "Principal Component",
    y = "Proportion of Variance Explained"
  ) +
  theme_minimal()

# Smoker yes/no: compare PC1
pca_dta <- dta |>
  select(smoker_new, on_art, all_of(retinal_var))

complete_rows <- complete.cases(pca_dta |>
  select(all_of(retinal_var)))

filtered_pca_dta <- pca_dta[complete_rows, ]

numeric_data <- filtered_pca_dta |>
  select(all_of(retinal_var))

metadata <- filtered_pca_dta |>
  select(smoker_new, on_art)

metadata <- metadata |>
  mutate(on_art = ifelse(is.na(on_art), "HIV negative", on_art))

pca_result <- prcomp(numeric_data, center = TRUE, scale. = TRUE)
pca_scores <- as.data.frame(pca_result$x)
pca_meta <- cbind(pca_scores, metadata)

t_test_result <- t.test(PC1 ~ smoker_new, data = pca_meta, var.equal = TRUE)

print(t_test_result)

wilcox_test_result <- wilcox.test(PC1 ~ smoker_new, data = pca_meta)

print(wilcox_test_result)
```

### CVD risk factor and HIV status: retinal

```{r}
#| label: pca-cvd-hiv-retinal
pca_dta <- dta |>
  select(cvd_risk_factor, on_art, all_of(retinal_var))

complete_rows <- complete.cases(pca_dta |>
  select(all_of(retinal_var)))

filtered_pca_dta <- pca_dta[complete_rows, ]

numeric_data <- filtered_pca_dta |>
  select(all_of(retinal_var))

metadata <- filtered_pca_dta |>
  select(cvd_risk_factor, on_art)

metadata <- metadata |>
  mutate(on_art = ifelse(is.na(on_art), "HIV negative", on_art))

pca_result <- prcomp(numeric_data, center = TRUE, scale. = TRUE)
pca_scores <- as.data.frame(pca_result$x)
pca_meta <- cbind(pca_scores, metadata)

ggplot(
  pca_meta,
  aes(x = PC1, y = PC2, color = cvd_risk_factor, shape = on_art)
) +
  geom_point() +
  labs(
    title = "PCA: PC1 vs PC2",
    x = "Principal Component 1",
    y = "Principal Component 2"
  ) +
  theme_minimal()

ggplot(
  pca_meta,
  aes(x = PC3, y = PC4, color = cvd_risk_factor, shape = on_art)
) +
  geom_point() +
  labs(
    title = "PCA: PC3 vs PC4 ",
    x = "Principal Component 3",
    y = "Principal Component 4"
  ) +
  theme_minimal()

pca_result

sdev <- pca_result$sdev
variance <- sdev^2
variance_explained <- variance / sum(variance)
print(variance_explained)

screeplot_df <- data.frame(
  PC = seq_along(pca_result$sdev),
  Variance = pca_result$sdev^2,
  Proportion = pca_result$sdev^2 / sum(pca_result$sdev^2)
)

ggplot(screeplot_df, aes(x = PC, y = Proportion)) +
  geom_point() +
  geom_line() +
  labs(
    title = "Scree Plot",
    x = "Principal Component",
    y = "Proportion of Variance Explained"
  ) +
  theme_minimal()

t_test_result <- t.test(
  PC1 ~ cvd_risk_factor,
  data = pca_meta,
  var.equal = TRUE
)

print(t_test_result)

wilcox_test_result <- wilcox.test(PC1 ~ cvd_risk_factor, data = pca_meta)

print(wilcox_test_result)
```

### HIV status: significant variables

```{r}
#| label: pca-hiv-sig
pca_dta <- dta |>
  select(hiv_pos, gender, on_art, all_of(t_sig_var_hiv))

complete_rows <- complete.cases(pca_dta |> select(all_of(t_sig_var_hiv)))

filtered_pca_dta <- pca_dta[complete_rows, ]

numeric_data <- filtered_pca_dta |>
  select(all_of(t_sig_var_hiv))

metadata <- filtered_pca_dta |>
  select(hiv_pos, gender, on_art)

metadata <- metadata |>
  mutate(on_art = ifelse(is.na(on_art), "HIV negative", on_art))

pca_result <- prcomp(numeric_data, center = TRUE, scale. = TRUE)
pca_scores <- as.data.frame(pca_result$x)
pca_meta <- cbind(pca_scores, metadata)

pca_plot <- ggplot(
  pca_meta,
  aes(x = PC1, y = PC2, color = hiv_pos, shape = on_art, linetype = gender)
) +
  geom_point() +
  labs(
    title = "PCA: PC1 vs PC2",
    x = "Principal Component 1",
    y = "Principal Component 2"
  ) +
  scale_color_manual(values = c(
    "Yes" = "red",
    "No" = "blue",
    "HIV negative" = "black"
  )) +
  theme_minimal()

ggsave("pca_HIV_sig_var.svg", plot = pca_plot, width = 10, height = 8)

ggplot(
  pca_meta,
  aes(x = PC3, y = PC4, color = hiv_pos, shape = on_art, linetype = gender)
) +
  geom_point() +
  labs(
    title = "PCA: PC3 vs PC4",
    x = "Principal Component 3",
    y = "Principal Component 4"
  ) +
  scale_color_manual(values = c(
    "Yes" = "red",
    "No" = "blue",
    "HIV negative" = "black"
  )) +
  theme_minimal()

ggplot(
  pca_meta,
  aes(x = PC2, y = PC8, color = hiv_pos, shape = on_art, linetype = gender)
) +
  geom_point() +
  labs(
    title = "PCA: PC2 vs PC8",
    x = "Principal Component 1",
    y = "Principal Component 2"
  ) +
  scale_color_manual(values = c(
    "Yes" = "red",
    "No" = "blue",
    "HIV negative" = "black"
  )) +
  theme_minimal()

ggplot(
  pca_meta,
  aes(x = PC12, y = PC13, color = hiv_pos, shape = on_art, linetype = gender)
) +
  geom_point() +
  labs(
    title = "PCA: PC2 vs PC8",
    x = "Principal Component 3",
    y = "Principal Component 4"
  ) +
  scale_color_manual(values = c(
    "Yes" = "red",
    "No" = "blue",
    "HIV negative" = "black"
  )) +
  theme_minimal()

pca_scores <- as.data.frame(pca_result$x)

pca_meta <- cbind(pca_scores, metadata)

datatable(pca_meta) |>
  formatStyle(
    columns = c(
      "hiv_pos", "gender", "on_art", "PC1", "PC2", "PC3", "PC4", "PC5", "PC6",
      "PC7", "PC8", "PC9", "PC10", "PC11", "PC12", "PC13"
    ),
    fontSize = "12px",
    color = "black",
    backgroundColor = "#DFF8EB"
  )
datatable(pca_meta,
  options = list(pageLength = 10, autoWidth = TRUE),
  rownames = FALSE
)

pca_result

screeplot_df <- data.frame(
  PC = seq_along(pca_result$sdev),
  Variance = pca_result$sdev^2,
  Proportion = pca_result$sdev^2 / sum(pca_result$sdev^2)
)

sdev <- pca_result$sdev
variance <- sdev^2
variance_explained <- variance / sum(variance)
print(variance_explained)

screeplot_df <- data.frame(
  PC = seq_along(pca_result$sdev),
  Variance = pca_result$sdev^2,
  Proportion = pca_result$sdev^2 / sum(pca_result$sdev^2)
)

ggplot(screeplot_df, aes(x = PC, y = Proportion)) +
  geom_point() +
  geom_line() +
  labs(
    title = "Scree Plot",
    x = "Principal Component",
    y = "Proportion of Variance Explained"
  ) +
  theme_minimal()

t_test_result <- t.test(PC1 ~ hiv_pos, data = pca_meta, var.equal = TRUE)

print(t_test_result)

wilcox_test_result <- wilcox.test(PC1 ~ hiv_pos, data = pca_meta)

print(wilcox_test_result)
```

### HIV status: significant retinal variables

```{r}
#| label: pca-hiv-sig-retinal
pca_dta <- dta |>
  select(
    hiv_pos, gender, on_art, venule_branch_parent_width,
    skeleton_fractal_analysis, venule_branch_larger_branch, crve
  )

complete_rows <- complete.cases(pca_dta |> select(
  venule_branch_parent_width, skeleton_fractal_analysis,
  venule_branch_larger_branch, crve
))

filtered_pca_dta <- pca_dta[complete_rows, ]

numeric_data <- filtered_pca_dta |>
  select(
    venule_branch_parent_width, skeleton_fractal_analysis,
    venule_branch_larger_branch, crve
  )

metadata <- filtered_pca_dta |>
  select(hiv_pos, gender, on_art)

metadata <- metadata |>
  mutate(on_art = ifelse(is.na(on_art), "HIV negative", on_art))

pca_result <- prcomp(numeric_data, center = TRUE, scale. = TRUE)
pca_scores <- as.data.frame(pca_result$x)
pca_meta <- cbind(pca_scores, metadata)

ggplot(
  pca_meta,
  aes(x = PC1, y = PC2, color = hiv_pos, shape = on_art, linetype = gender)
) +
  geom_point() +
  labs(
    title = "PCA: PC1 vs PC2",
    x = "Principal Component 1",
    y = "Principal Component 2"
  ) +
  scale_color_manual(values = c(
    "Yes" = "red",
    "No" = "blue",
    "HIV negative" = "black"
  )) +
  theme_minimal()

ggplot(
  pca_meta,
  aes(x = PC3, y = PC4, color = hiv_pos, shape = on_art, linetype = gender)
) +
  geom_point() +
  labs(
    title = "PCA: PC3 vs PC4",
    x = "Principal Component 3",
    y = "Principal Component 4"
  ) +
  scale_color_manual(values = c(
    "Yes" = "red",
    "No" = "blue",
    "HIV negative" = "black"
  )) +
  theme_minimal()

ggplot(
  pca_meta,
  aes(x = PC1, y = PC2, color = on_art, shape = hiv_pos, linetype = gender)
) +
  geom_point() +
  labs(
    title = "PCA: PC1 vs PC2",
    x = "Principal Component 1",
    y = "Principal Component 2"
  ) +
  scale_color_manual(values = c(
    "Yes" = "red",
    "No" = "blue",
    "HIV negative" = "black"
  )) +
  theme_minimal()

ggplot(
  pca_meta,
  aes(x = PC3, y = PC4, color = on_art, shape = hiv_pos, linetype = gender)
) +
  geom_point() +
  labs(
    title = "PCA: PC3 vs PC4",
    x = "Principal Component 3",
    y = "Principal Component 4"
  ) +
  scale_color_manual(values = c(
    "Yes" = "red",
    "No" = "blue",
    "HIV negative" = "black"
  )) +
  theme_minimal()

screeplot_df <- data.frame(
  PC = seq_along(pca_result$sdev),
  Variance = pca_result$sdev^2,
  Proportion = pca_result$sdev^2 / sum(pca_result$sdev^2)
)

sdev <- pca_result$sdev
variance <- sdev^2
variance_explained <- variance / sum(variance)
print(variance_explained)

screeplot_df <- data.frame(
  PC = seq_along(pca_result$sdev),
  Variance = pca_result$sdev^2,
  Proportion = pca_result$sdev^2 / sum(pca_result$sdev^2)
)

ggplot(screeplot_df, aes(x = PC, y = Proportion)) +
  geom_point() +
  geom_line() +
  labs(
    title = "Scree Plot",
    x = "Principal Component",
    y = "Proportion of Variance Explained"
  ) +
  theme_minimal()

t_test_result <- t.test(PC1 ~ hiv_pos, data = pca_meta, var.equal = TRUE)

print(t_test_result)

wilcox_test_result <- wilcox.test(PC1 ~ hiv_pos, data = pca_meta)

print(wilcox_test_result)
```

### ART status: significant variables

```{r}
#| label: pca-art-sig
pca_dta <- dta |>
  select(on_art, all_of(t_sig_var_art)) |>
  filter((!is.na(on_art)))

complete_rows <- complete.cases(pca_dta |> select(all_of(t_sig_var_art)))

filtered_pca_dta <- pca_dta[complete_rows, ]

numeric_data <- filtered_pca_dta |>
  select(all_of(t_sig_var_art))

metadata <- filtered_pca_dta |>
  select(on_art)

pca_result <- prcomp(numeric_data, center = TRUE, scale. = TRUE)
pca_scores <- as.data.frame(pca_result$x)
pca_meta <- cbind(pca_scores, metadata)

pca_plot <- ggplot(pca_meta, aes(x = PC1, y = PC2, color = on_art)) +
  geom_point() +
  labs(
    title = "PCA: PC1 vs PC2",
    x = "Principal Component 1",
    y = "Principal Component 2"
  ) +
  scale_color_manual(values = c("Yes" = "red", "No" = "blue")) +
  theme_minimal()

ggsave("pca_ART_sig.svg", plot = pca_plot, width = 10, height = 8)

ggplot(pca_meta, aes(x = PC3, y = PC4, color = on_art)) +
  geom_point() +
  labs(
    title = "PCA: PC3 vs PC4",
    x = "Principal Component 3",
    y = "Principal Component 4"
  ) +
  scale_color_manual(values = c("Yes" = "red", "No" = "blue")) +
  theme_minimal()

ggplot(pca_meta, aes(x = PC3, y = PC6, color = on_art)) +
  geom_point() +
  labs(
    title = "PCA: PC1 vs PC2",
    x = "Principal Component 1",
    y = "Principal Component 2"
  ) +
  scale_color_manual(values = c("Yes" = "red", "No" = "blue")) +
  theme_minimal()

datatable(pca_meta) |>
  formatStyle(
    columns = c(
      "on_art", "PC1", "PC2", "PC3", "PC4", "PC5", "PC6", "PC7", "PC8", "PC9",
      "PC10", "PC11", "PC12"
    ),
    fontSize = "12px",
    color = "black",
    backgroundColor = "#DFF8EB"
  )
datatable(pca_meta,
  options = list(pageLength = 10, autoWidth = TRUE),
  rownames = FALSE
)

screeplot_df <- data.frame(
  PC = seq_along(pca_result$sdev),
  Variance = pca_result$sdev^2,
  Proportion = pca_result$sdev^2 / sum(pca_result$sdev^2)
)

pca_result

sdev <- pca_result$sdev
variance <- sdev^2
variance_explained <- variance / sum(variance)
print(variance_explained)

screeplot_df <- data.frame(
  PC = seq_along(pca_result$sdev),
  Variance = pca_result$sdev^2,
  Proportion = pca_result$sdev^2 / sum(pca_result$sdev^2)
)

ggplot(screeplot_df, aes(x = PC, y = Proportion)) +
  geom_point() +
  geom_line() +
  labs(
    title = "Scree Plot",
    x = "Principal Component",
    y = "Proportion of Variance Explained"
  ) +
  theme_minimal()

t_test_result <- t.test(PC1 ~ on_art, data = pca_meta, var.equal = TRUE)

print(t_test_result)

wilcox_test_result <- wilcox.test(PC1 ~ on_art, data = pca_meta)

print(wilcox_test_result)
```

### ART status: significant retinal variables

```{r}
#| label: pca-art-sig-retinal
pca_dta <- dta |>
  select(
    on_art, fourier_fractal_enhanced, branch_segment_length_kurtosis,
    venule_mean_angular_asymmetry, segmented_fractal_analysis
  ) |>
  filter((!is.na(on_art)))

complete_rows <- complete.cases(pca_dta |>
  select(
    fourier_fractal_enhanced, branch_segment_length_kurtosis,
    venule_mean_angular_asymmetry, segmented_fractal_analysis
  ))

filtered_pca_dta <- pca_dta[complete_rows, ]

numeric_data <- filtered_pca_dta |>
  select(
    fourier_fractal_enhanced, branch_segment_length_kurtosis,
    venule_mean_angular_asymmetry, segmented_fractal_analysis
  )

metadata <- filtered_pca_dta |>
  select(on_art)

pca_result <- prcomp(numeric_data, center = TRUE, scale. = TRUE)
pca_scores <- as.data.frame(pca_result$x)
pca_meta <- cbind(pca_scores, metadata)

pca_plot <- ggplot(pca_meta, aes(x = PC1, y = PC2, color = on_art)) +
  geom_point(size = 4) +
  labs(
    title = "PCA: PC1 vs PC2",
    x = "Principal Component 1",
    y = "Principal Component 2"
  ) +
  scale_color_manual(values = c("Yes" = "red", "No" = "blue")) +
  theme_minimal()

ggsave("pca_ret_sig.svg", plot = pca_plot, width = 10, height = 8)

ggplot(pca_meta, aes(x = PC3, y = PC4, color = on_art)) +
  geom_point() +
  labs(
    title = "PCA: PC3 vs PC4",
    x = "Principal Component 3",
    y = "Principal Component 4"
  ) +
  scale_color_manual(values = c("Yes" = "red", "No" = "blue")) +
  theme_minimal()

pca_result

sdev <- pca_result$sdev
variance <- sdev^2
variance_explained <- variance / sum(variance)
print(variance_explained)

screeplot_df <- data.frame(
  PC = seq_along(pca_result$sdev),
  Variance = pca_result$sdev^2,
  Proportion = pca_result$sdev^2 / sum(pca_result$sdev^2)
)

ggplot(screeplot_df, aes(x = PC, y = Proportion)) +
  geom_point() +
  geom_line() +
  labs(
    title = "Scree Plot",
    x = "Principal Component",
    y = "Proportion of Variance Explained"
  ) +
  theme_minimal()

t_test_result <- t.test(PC1 ~ on_art, data = pca_meta, var.equal = TRUE)

print(t_test_result)

wilcox_test_result <- wilcox.test(PC1 ~ on_art, data = pca_meta)

print(wilcox_test_result)
```

### HIV and ART status: clinical

```{r}
#| label: pca-hiv-art-clinical
pca_dta <- dta |>
  select(hiv_pos, on_art, all_of(clinical_var_pca))

complete_rows <- complete.cases(pca_dta |>
  select(all_of(clinical_var_pca)))

filtered_pca_dta <- pca_dta[complete_rows, ]

numeric_data <- filtered_pca_dta |>
  select(all_of(clinical_var_pca))

metadata <- filtered_pca_dta |>
  select(on_art, hiv_pos)

metadata <- metadata |>
  mutate(on_art = ifelse(is.na(on_art), "HIV negative", on_art))

pca_result <- prcomp(numeric_data, center = TRUE, scale. = TRUE)
pca_scores <- as.data.frame(pca_result$x)
pca_meta <- cbind(pca_scores, metadata)

ggplot(pca_meta, aes(x = PC1, y = PC2, color = on_art, shape = hiv_pos)) +
  geom_point() +
  labs(
    title = "PCA: PC1 vs PC2",
    x = "Principal Component 1",
    y = "Principal Component 2"
  ) +
  scale_color_manual(values = c(
    "Yes" = "red",
    "No" = "blue",
    "HIV negative" = "black"
  )) +
  theme_minimal()

ggplot(pca_meta, aes(x = PC3, y = PC4, color = on_art, shape = hiv_pos)) +
  geom_point() +
  labs(
    title = "PCA: PC3 vs PC4 ",
    x = "Principal Component 3",
    y = "Principal Component 4"
  ) +
  scale_color_manual(values = c(
    "Yes" = "red",
    "No" = "blue",
    "HIV negative" = "black"
  )) +
  theme_minimal()

pca_result
```

### Age group and ART status: clinical

```{r}
#| label: pca-age-art-clinical
pca_dta <- dta |>
  select(age_group, on_art, all_of(clinical_var_pca))

complete_rows <- complete.cases(pca_dta |>
  select(all_of(clinical_var_pca)))

filtered_pca_dta <- pca_dta[complete_rows, ]

numeric_data <- filtered_pca_dta |>
  select(all_of(clinical_var_pca))

metadata <- filtered_pca_dta |>
  select(age_group, on_art)

metadata <- metadata |>
  mutate(on_art = ifelse(is.na(on_art), "HIV negative", on_art))

pca_result <- prcomp(numeric_data, center = TRUE, scale. = TRUE)
pca_scores <- as.data.frame(pca_result$x)
pca_meta <- cbind(pca_scores, metadata)

ggplot(pca_meta, aes(x = PC1, y = PC2, color = age_group, shape = on_art)) +
  geom_point() +
  labs(
    title = "PCA: PC1 vs PC2",
    x = "Principal Component 1",
    y = "Principal Component 2"
  ) +
  theme_minimal()

ggplot(pca_meta, aes(x = PC3, y = PC4, color = age_group, shape = on_art)) +
  geom_point() +
  labs(
    title = "PCA: PC3 vs PC4 ",
    x = "Principal Component 3",
    y = "Principal Component 4"
  ) +
  theme_minimal()

pca_result
```

### Smoking and HIV status: clinical

```{r}
#| label: pca-smoking-hiv-clinical
pca_dta <- dta |>
  select(smoker, on_art, all_of(clinical_var_pca))

complete_rows <- complete.cases(pca_dta |>
  select(all_of(clinical_var_pca)))

filtered_pca_dta <- pca_dta[complete_rows, ]

numeric_data <- filtered_pca_dta |>
  select(all_of(clinical_var_pca))

metadata <- filtered_pca_dta |>
  select(smoker, on_art)

metadata <- metadata |>
  mutate(on_art = ifelse(is.na(on_art), "HIV negative", on_art))

pca_result <- prcomp(numeric_data, center = TRUE, scale. = TRUE)
pca_scores <- as.data.frame(pca_result$x)
pca_meta <- cbind(pca_scores, metadata)

ggplot(pca_meta, aes(x = PC1, y = PC2, color = on_art, shape = smoker)) +
  geom_point() +
  labs(
    title = "PCA: PC1 vs PC2",
    x = "Principal Component 1",
    y = "Principal Component 2"
  ) +
  theme_minimal()

ggplot(pca_meta, aes(x = PC3, y = PC4, color = on_art, shape = smoker)) +
  geom_point() +
  labs(
    title = "PCA: PC3 vs PC4 ",
    x = "Principal Component 3",
    y = "Principal Component 4"
  ) +
  theme_minimal()

pca_result
```

### CVD risk factor and HIV status: clinical

```{r}
#| label: pca-cvd-hiv-clinical
pca_dta <- dta |>
  select(cvd_risk_factor, on_art, all_of(clinical_var_pca))

complete_rows <- complete.cases(pca_dta |>
  select(all_of(clinical_var_pca)))

filtered_pca_dta <- pca_dta[complete_rows, ]

numeric_data <- filtered_pca_dta |>
  select(all_of(clinical_var_pca))

metadata <- filtered_pca_dta |>
  select(cvd_risk_factor, on_art)

metadata <- metadata |>
  mutate(on_art = ifelse(is.na(on_art), "HIV negative", on_art))

pca_result <- prcomp(numeric_data, center = TRUE, scale. = TRUE)
pca_scores <- as.data.frame(pca_result$x)
pca_meta <- cbind(pca_scores, metadata)

ggplot(
  pca_meta,
  aes(x = PC1, y = PC2, color = on_art, shape = cvd_risk_factor)
) +
  geom_point() +
  labs(
    title = "PCA: PC1 vs PC2",
    x = "Principal Component 1",
    y = "Principal Component 2"
  ) +
  theme_minimal()

ggplot(
  pca_meta,
  aes(x = PC3, y = PC4, color = on_art, shape = cvd_risk_factor)
) +
  geom_point() +
  labs(
    title = "PCA: PC3 vs PC4 ",
    x = "Principal Component 3",
    y = "Principal Component 4"
  ) +
  theme_minimal()

pca_result
```

### HIV and ART status: all variables

```{r}
#| label: pca-hiv-art-all
pca_dta <- dta |>
  select(hiv_pos, on_art, all_of(all_var))

complete_rows <- complete.cases(pca_dta |>
  select(all_of(all_var)))

filtered_pca_dta <- pca_dta[complete_rows, ]

numeric_data <- filtered_pca_dta |>
  select(all_of(all_var))

metadata <- filtered_pca_dta |>
  select(on_art, hiv_pos)

metadata <- metadata |>
  mutate(on_art = ifelse(is.na(on_art), "HIV negative", on_art))

pca_result <- prcomp(numeric_data, center = TRUE, scale. = TRUE)
pca_scores <- as.data.frame(pca_result$x)
pca_meta <- cbind(pca_scores, metadata)

ggplot(pca_meta, aes(x = PC1, y = PC2, color = on_art, shape = hiv_pos)) +
  geom_point() +
  labs(
    title = "PCA: PC1 vs PC2",
    x = "Principal Component 1",
    y = "Principal Component 2"
  ) +
  scale_color_manual(values = c(
    "Yes" = "red",
    "No" = "blue",
    "HIV negative" = "black"
  )) +
  theme_minimal()

ggplot(pca_meta, aes(x = PC3, y = PC4, color = on_art, shape = hiv_pos)) +
  geom_point() +
  labs(
    title = "PCA: PC3 vs PC4 ",
    x = "Principal Component 3",
    y = "Principal Component 4"
  ) +
  scale_color_manual(values = c(
    "Yes" = "red",
    "No" = "blue",
    "HIV negative" = "black"
  )) +
  theme_minimal()

pca_result
sdev <- pca_result$sdev
variance <- sdev^2
variance_explained <- variance / sum(variance)
print(variance_explained)
```

### Age group and HIV status: all variables

```{r}
#| label: pca-age-hiv-all
pca_dta <- dta |>
  select(age_group, on_art, all_of(all_var))

complete_rows <- complete.cases(pca_dta |>
  select(all_of(all_var)))

filtered_pca_dta <- pca_dta[complete_rows, ]

numeric_data <- filtered_pca_dta |>
  select(all_of(all_var))

metadata <- filtered_pca_dta |>
  select(age_group, on_art)

metadata <- metadata |>
  mutate(on_art = ifelse(is.na(on_art), "HIV negative", on_art))

pca_result <- prcomp(numeric_data, center = TRUE, scale. = TRUE)
pca_scores <- as.data.frame(pca_result$x)
pca_meta <- cbind(pca_scores, metadata)

ggplot(pca_meta, aes(x = PC1, y = PC2, color = age_group, shape = on_art)) +
  geom_point() +
  labs(
    title = "PCA: PC1 vs PC2",
    x = "Principal Component 1",
    y = "Principal Component 2"
  ) +
  theme_minimal()

ggplot(pca_meta, aes(x = PC3, y = PC4, color = age_group, shape = on_art)) +
  geom_point() +
  labs(
    title = "PCA: PC3 vs PC4 ",
    x = "Principal Component 3",
    y = "Principal Component 4"
  ) +
  theme_minimal()

pca_result
```

### Smoking and HIV status: all variables

```{r}
#| label: pca-smoking-hiv-all
pca_dta <- dta |>
  select(smoker, on_art, all_of(all_var))

complete_rows <- complete.cases(pca_dta |>
  select(all_of(all_var)))

filtered_pca_dta <- pca_dta[complete_rows, ]

numeric_data <- filtered_pca_dta |>
  select(all_of(all_var))

metadata <- filtered_pca_dta |>
  select(smoker, on_art)

metadata <- metadata |>
  mutate(on_art = ifelse(is.na(on_art), "HIV negative", on_art))

pca_result <- prcomp(numeric_data, center = TRUE, scale. = TRUE)
pca_scores <- as.data.frame(pca_result$x)
pca_meta <- cbind(pca_scores, metadata)

ggplot(pca_meta, aes(x = PC1, y = PC2, color = on_art, shape = smoker)) +
  geom_point() +
  labs(
    title = "PCA: PC1 vs PC2",
    x = "Principal Component 1",
    y = "Principal Component 2"
  ) +
  theme_minimal()

ggplot(pca_meta, aes(x = PC3, y = PC4, color = on_art, shape = smoker)) +
  geom_point() +
  labs(
    title = "PCA: PC3 vs PC4 ",
    x = "Principal Component 3",
    y = "Principal Component 4"
  ) +
  theme_minimal()

pca_result
```

### CVD risk factor and HIV status: all variables

```{r}
#| label: pca-cvd-hiv-all
pca_dta <- dta |>
  select(cvd_risk_factor, on_art, all_of(all_var))

complete_rows <- complete.cases(pca_dta |>
  select(all_of(all_var)))

filtered_pca_dta <- pca_dta[complete_rows, ]

numeric_data <- filtered_pca_dta |>
  select(all_of(all_var))

metadata <- filtered_pca_dta |>
  select(cvd_risk_factor, on_art)

metadata <- metadata |>
  mutate(on_art = ifelse(is.na(on_art), "HIV negative", on_art))

pca_result <- prcomp(numeric_data, center = TRUE, scale. = TRUE)
pca_scores <- as.data.frame(pca_result$x)
pca_meta <- cbind(pca_scores, metadata)

ggplot(
  pca_meta,
  aes(x = PC1, y = PC2, color = on_art, shape = cvd_risk_factor)
) +
  geom_point() +
  labs(
    title = "PCA: PC1 vs PC2",
    x = "Principal Component 1",
    y = "Principal Component 2"
  ) +
  theme_minimal()

ggplot(
  pca_meta,
  aes(x = PC3, y = PC4, color = on_art, shape = cvd_risk_factor)
) +
  geom_point() +
  labs(
    title = "PCA: PC3 vs PC4 ",
    x = "Principal Component 3",
    y = "Principal Component 4"
  ) +
  theme_minimal()

pca_result
```

### ART line and viral load: all variables

```{r}
#| label: pca-art-line-viral-load-all
pca_dta <- dta |>
  select(art_line, viral_load_group, all_of(all_var))

complete_rows <- complete.cases(pca_dta |>
  select(all_of(all_var)))

filtered_pca_dta <- pca_dta[complete_rows, ]

numeric_data <- filtered_pca_dta |>
  select(all_of(all_var))

metadata <- filtered_pca_dta |>
  select(art_line, viral_load_group)

metadata <- metadata |>
  mutate(
    art_line = ifelse(is.na(art_line), "HIV negative", art_line),
    viral_load_group = ifelse(
      is.na(viral_load_group), "HIV negative", viral_load_group
    )
  )

pca_result <- prcomp(numeric_data, center = TRUE, scale. = TRUE)
pca_scores <- as.data.frame(pca_result$x)
pca_meta <- cbind(pca_scores, metadata)

ggplot(
  pca_meta,
  aes(x = PC1, y = PC2, color = viral_load_group, shape = art_line)
) +
  geom_point() +
  labs(
    title = "PCA: PC1 vs PC2",
    x = "Principal Component 1",
    y = "Principal Component 2"
  ) +
  theme_minimal()

ggplot(
  pca_meta,
  aes(x = PC3, y = PC4, color = viral_load_group, shape = art_line)
) +
  geom_point() +
  labs(
    title = "PCA: PC3 vs PC4 ",
    x = "Principal Component 3",
    y = "Principal Component 4"
  ) +
  theme_minimal()

pca_result
```

### ART duration: significant retinal variables

```{r}
#| label: pca-art-duration-sig-retinal
pca_dta <- dta |>
  select(art_duration_group, on_art, all_of(sig_art_duration_ret))

complete_rows <- complete.cases(pca_dta |>
  select(all_of(sig_art_duration_ret)))

filtered_pca_dta <- pca_dta[complete_rows, ]

numeric_data <- filtered_pca_dta |>
  select(all_of(sig_art_duration_ret))

metadata <- filtered_pca_dta |>
  select(on_art, art_duration_group)

pca_result <- prcomp(numeric_data, center = TRUE, scale. = TRUE)
pca_scores <- as.data.frame(pca_result$x)
pca_meta <- cbind(pca_scores, metadata)

ggplot(
  pca_meta,
  aes(x = PC1, y = PC2, color = art_duration_group, shape = art_duration_group)
) +
  geom_point() +
  labs(
    title = "PCA: PC1 vs PC2",
    x = "Principal Component 1",
    y = "Principal Component 2"
  ) +
  scale_color_manual(values = c(
    "1-5 years" = "red",
    "More than 5 years" = "blue",
    "Less than a year" = "black"
  )) +
  theme_minimal()

ggplot(
  pca_meta,
  aes(x = PC3, y = PC4, color = art_duration_group, shape = art_duration_group)
) +
  geom_point() +
  labs(
    title = "PCA: PC3 vs PC4 ",
    x = "Principal Component 3",
    y = "Principal Component 4"
  ) +
  scale_color_manual(values = c(
    "1-5 years" = "red",
    "More than 5 years" = "blue",
    "Less than a year" = "black"
  )) +
  theme_minimal()

pca_result

sdev <- pca_result$sdev
variance <- sdev^2
variance_explained <- variance / sum(variance)
print(variance_explained)

screeplot_df <- data.frame(
  PC = seq_along(pca_result$sdev),
  Variance = pca_result$sdev^2,
  Proportion = pca_result$sdev^2 / sum(pca_result$sdev^2)
)

ggplot(screeplot_df, aes(x = PC, y = Proportion)) +
  geom_point() +
  geom_line() +
  labs(
    title = "Scree Plot",
    x = "Principal Component",
    y = "Proportion of Variance Explained"
  ) +
  theme_minimal()

# Two ART duration groups (0-5 vs more than 5 years): compare PC1

pca_dta <- dta |>
  select(art_duration_group_2, on_art, all_of(sig_art_duration_ret))

complete_rows <- complete.cases(pca_dta |>
  select(all_of(sig_art_duration_ret)))

filtered_pca_dta <- pca_dta[complete_rows, ]

numeric_data <- filtered_pca_dta |>
  select(all_of(sig_art_duration_ret))

metadata <- filtered_pca_dta |>
  select(on_art, art_duration_group_2)

pca_result <- prcomp(numeric_data, center = TRUE, scale. = TRUE)
pca_scores <- as.data.frame(pca_result$x)
pca_meta <- cbind(pca_scores, metadata)

t_test_result <- t.test(
  PC1 ~ art_duration_group_2,
  data = pca_meta,
  var.equal = TRUE
)

print(t_test_result)

wilcox_test_result <- wilcox.test(PC1 ~ art_duration_group_2, data = pca_meta)

print(wilcox_test_result)
```

### CD4 count (clinic): significant variables

```{r}
#| label: pca-cd4-clinic-sig
pca_dta <- dta |>
  select(cd_count_clinic_group, on_art, all_of(sig_clinic_cd_count))

complete_rows <- complete.cases(pca_dta |>
  select(all_of(sig_clinic_cd_count)))

filtered_pca_dta <- pca_dta[complete_rows, ]

numeric_data <- filtered_pca_dta |>
  select(all_of(sig_clinic_cd_count))

metadata <- filtered_pca_dta |>
  select(on_art, (cd_count_clinic_group))

pca_result <- prcomp(numeric_data, center = TRUE, scale. = TRUE)
pca_scores <- as.data.frame(pca_result$x)
pca_meta <- cbind(pca_scores, metadata)

ggplot(
  pca_meta,
  aes(x = PC1, y = PC2, color = cd_count_clinic_group, shape = on_art)
) +
  geom_point() +
  labs(
    title = "PCA: PC1 vs PC2",
    x = "Principal Component 1",
    y = "Principal Component 2"
  ) +
  scale_color_manual(values = c(
    "0-199" = "red",
    "200-499" = "blue",
    "500+" = "black"
  )) +
  theme_minimal()

ggplot(
  pca_meta,
  aes(x = PC3, y = PC4, color = cd_count_clinic_group, shape = on_art)
) +
  geom_point() +
  labs(
    title = "PCA: PC3 vs PC4 ",
    x = "Principal Component 3",
    y = "Principal Component 4"
  ) +
  scale_color_manual(values = c(
    "0-199" = "red",
    "200-499" = "blue",
    "500+" = "black"
  )) +
  theme_minimal()

pca_result

sdev <- pca_result$sdev
variance <- sdev^2
variance_explained <- variance / sum(variance)
print(variance_explained)

screeplot_df <- data.frame(
  PC = seq_along(pca_result$sdev),
  Variance = pca_result$sdev^2,
  Proportion = pca_result$sdev^2 / sum(pca_result$sdev^2)
)

ggplot(screeplot_df, aes(x = PC, y = Proportion)) +
  geom_point() +
  geom_line() +
  labs(
    title = "Scree Plot",
    x = "Principal Component",
    y = "Proportion of Variance Explained"
  ) +
  theme_minimal()

# Two CD4 count groups (below 500 vs 500+): compare PC1
pca_dta <- dta |>
  select(cd_count_clinic_group_2, on_art, all_of(sig_clinic_cd_count))

complete_rows <- complete.cases(pca_dta |>
  select(all_of(sig_clinic_cd_count)))

filtered_pca_dta <- pca_dta[complete_rows, ]

numeric_data <- filtered_pca_dta |>
  select(all_of(sig_clinic_cd_count))

metadata <- filtered_pca_dta |>
  select(on_art, (cd_count_clinic_group_2))

pca_result <- prcomp(numeric_data, center = TRUE, scale. = TRUE)
pca_scores <- as.data.frame(pca_result$x)
pca_meta <- cbind(pca_scores, metadata)

t_test_result <- t.test(
  PC1 ~ cd_count_clinic_group_2,
  data = pca_meta,
  var.equal = TRUE
)

print(t_test_result)

wilcox_test_result <- wilcox.test(
  PC1 ~ cd_count_clinic_group_2,
  data = pca_meta
)

print(wilcox_test_result)
```

### CD4 count (clinic): significant retinal variables

```{r}
#| label: pca-cd4-clinic-sig-retinal
pca_dta <- dta |>
  select(cd_count_clinic_group, on_art, all_of(sig_clinic_cd_count_ret))

complete_rows <- complete.cases(pca_dta |>
  select(all_of(sig_clinic_cd_count_ret)))

filtered_pca_dta <- pca_dta[complete_rows, ]

numeric_data <- filtered_pca_dta |>
  select(all_of(sig_clinic_cd_count_ret))

metadata <- filtered_pca_dta |>
  select(on_art, (cd_count_clinic_group))

pca_result <- prcomp(numeric_data, center = TRUE, scale. = TRUE)
pca_scores <- as.data.frame(pca_result$x)
pca_meta <- cbind(pca_scores, metadata)

ggplot(
  pca_meta,
  aes(x = PC1, y = PC2, color = cd_count_clinic_group, shape = on_art)
) +
  geom_point() +
  labs(
    title = "PCA: PC1 vs PC2",
    x = "Principal Component 1",
    y = "Principal Component 2"
  ) +
  scale_color_manual(values = c(
    "0-199" = "red",
    "200-499" = "blue",
    "500+" = "black"
  )) +
  theme_minimal()

ggplot(
  pca_meta,
  aes(x = PC3, y = PC4, color = cd_count_clinic_group, shape = on_art)
) +
  geom_point() +
  labs(
    title = "PCA: PC3 vs PC4 ",
    x = "Principal Component 3",
    y = "Principal Component 4"
  ) +
  scale_color_manual(values = c(
    "0-199" = "red",
    "200-499" = "blue",
    "500+" = "black"
  )) +
  theme_minimal()

pca_result

sdev <- pca_result$sdev
variance <- sdev^2
variance_explained <- variance / sum(variance)
print(variance_explained)

screeplot_df <- data.frame(
  PC = seq_along(pca_result$sdev),
  Variance = pca_result$sdev^2,
  Proportion = pca_result$sdev^2 / sum(pca_result$sdev^2)
)

ggplot(screeplot_df, aes(x = PC, y = Proportion)) +
  geom_point() +
  geom_line() +
  labs(
    title = "Scree Plot",
    x = "Principal Component",
    y = "Proportion of Variance Explained"
  ) +
  theme_minimal()
```

### CD4 count (EndoAfrica): significant variables

```{r}
#| label: pca-cd4-endo-sig
pca_dta <- dta |>
  select(cd4_count_endo_group, on_art, all_of(sig_endo_result))

complete_rows <- complete.cases(pca_dta |>
  select(all_of(sig_endo_result)))

filtered_pca_dta <- pca_dta[complete_rows, ]

numeric_data <- filtered_pca_dta |>
  select(all_of(sig_endo_result))

metadata <- filtered_pca_dta |>
  select(on_art, cd4_count_endo_group)

pca_result <- prcomp(numeric_data, center = TRUE, scale. = TRUE)
pca_scores <- as.data.frame(pca_result$x)
pca_meta <- cbind(pca_scores, metadata)

ggplot(
  pca_meta,
  aes(x = PC1, y = PC2, color = cd4_count_endo_group, shape = on_art)
) +
  geom_point() +
  labs(
    title = "PCA: PC3 vs PC4 ",
    x = "Principal Component 3",
    y = "Principal Component 4"
  ) +
  scale_color_manual(values = c(
    "0-199" = "red",
    "200-499" = "blue",
    "500+" = "black"
  )) +
  theme_minimal()

ggplot(
  pca_meta,
  aes(x = PC3, y = PC4, color = cd4_count_endo_group, shape = on_art)
) +
  geom_point() +
  labs(
    title = "PCA: PC3 vs PC4 ",
    x = "Principal Component 3",
    y = "Principal Component 4"
  ) +
  scale_color_manual(values = c(
    "0-199" = "red",
    "200-499" = "blue",
    "500+" = "black"
  )) +
  theme_minimal()

pca_result

sdev <- pca_result$sdev
variance <- sdev^2
variance_explained <- variance / sum(variance)
print(variance_explained)

screeplot_df <- data.frame(
  PC = seq_along(pca_result$sdev),
  Variance = pca_result$sdev^2,
  Proportion = pca_result$sdev^2 / sum(pca_result$sdev^2)
)

ggplot(screeplot_df, aes(x = PC, y = Proportion)) +
  geom_point() +
  geom_line() +
  labs(
    title = "Scree Plot",
    x = "Principal Component",
    y = "Proportion of Variance Explained"
  ) +
  theme_minimal()
```

### CD4 count (EndoAfrica): significant retinal variables

```{r}
#| label: pca-cd4-endo-sig-retinal
pca_dta <- dta |>
  select(cd4_count_endo_group, on_art, all_of(sig_endo_result_ret))

complete_rows <- complete.cases(pca_dta |>
  select(all_of(sig_endo_result_ret)))

filtered_pca_dta <- pca_dta[complete_rows, ]

numeric_data <- filtered_pca_dta |>
  select(all_of(sig_endo_result_ret))

metadata <- filtered_pca_dta |>
  select(on_art, cd4_count_endo_group)

pca_result <- prcomp(numeric_data, center = TRUE, scale. = TRUE)
pca_scores <- as.data.frame(pca_result$x)
pca_meta <- cbind(pca_scores, metadata)

ggplot(
  pca_meta,
  aes(x = PC1, y = PC2, color = cd4_count_endo_group, shape = on_art)
) +
  geom_point() +
  labs(
    title = "PCA: PC3 vs PC4 ",
    x = "Principal Component 3",
    y = "Principal Component 4"
  ) +
  scale_color_manual(values = c(
    "0-199" = "red",
    "200-499" = "blue",
    "500+" = "black"
  )) +
  theme_minimal()

ggplot(
  pca_meta,
  aes(x = PC3, y = PC4, color = cd4_count_endo_group, shape = on_art)
) +
  geom_point() +
  labs(
    title = "PCA: PC3 vs PC4 ",
    x = "Principal Component 3",
    y = "Principal Component 4"
  ) +
  scale_color_manual(values = c(
    "0-199" = "red",
    "200-499" = "blue",
    "500+" = "black"
  )) +
  theme_minimal()

pca_result

sdev <- pca_result$sdev
variance <- sdev^2
variance_explained <- variance / sum(variance)
print(variance_explained)

screeplot_df <- data.frame(
  PC = seq_along(pca_result$sdev),
  Variance = pca_result$sdev^2,
  Proportion = pca_result$sdev^2 / sum(pca_result$sdev^2)
)

ggplot(screeplot_df, aes(x = PC, y = Proportion)) +
  geom_point() +
  geom_line() +
  labs(
    title = "Scree Plot",
    x = "Principal Component",
    y = "Proportion of Variance Explained"
  ) +
  theme_minimal()
```

### Hypertension: significant variables

```{r}
#| label: pca-hypertension-sig
pca_dta <- dta |>
  select(hypertension, on_art, all_of(sig_hypertension))

complete_rows <- complete.cases(pca_dta |>
  select(all_of(sig_hypertension)))

filtered_pca_dta <- pca_dta[complete_rows, ]

numeric_data <- filtered_pca_dta |>
  select(all_of(sig_hypertension))

metadata <- filtered_pca_dta |>
  select(on_art, hypertension)

pca_result <- prcomp(numeric_data, center = TRUE, scale. = TRUE)
pca_scores <- as.data.frame(pca_result$x)
pca_meta <- cbind(pca_scores, metadata)

ggplot(pca_meta, aes(x = PC1, y = PC2, color = hypertension, shape = on_art)) +
  geom_point() +
  labs(
    title = "PCA: PC3 vs PC4 ",
    x = "Principal Component 3",
    y = "Principal Component 4"
  ) +
  scale_color_manual(values = c(
    "Yes" = "red",
    "No" = "blue",
    "500+" = "black"
  )) +
  theme_minimal()

ggplot(pca_meta, aes(x = PC3, y = PC4, color = hypertension, shape = on_art)) +
  geom_point() +
  labs(
    title = "PCA: PC3 vs PC4 ",
    x = "Principal Component 3",
    y = "Principal Component 4"
  ) +
  scale_color_manual(values = c(
    "Yes" = "red",
    "No" = "blue",
    "500+" = "black"
  )) +
  theme_minimal()

pca_result

sdev <- pca_result$sdev
variance <- sdev^2
variance_explained <- variance / sum(variance)
print(variance_explained)

screeplot_df <- data.frame(
  PC = seq_along(pca_result$sdev),
  Variance = pca_result$sdev^2,
  Proportion = pca_result$sdev^2 / sum(pca_result$sdev^2)
)

ggplot(screeplot_df, aes(x = PC, y = Proportion)) +
  geom_point() +
  geom_line() +
  labs(
    title = "Scree Plot",
    x = "Principal Component",
    y = "Proportion of Variance Explained"
  ) +
  theme_minimal()
```

### Smoking: significant variables

```{r}
#| label: pca-smoking-sig
pca_dta <- dta |>
  select(smoker, on_art, all_of(sig_smoker))

complete_rows <- complete.cases(pca_dta |>
  select(all_of(sig_smoker)))

filtered_pca_dta <- pca_dta[complete_rows, ]

numeric_data <- filtered_pca_dta |>
  select(all_of(sig_smoker))

metadata <- filtered_pca_dta |>
  select(on_art, smoker)

pca_result <- prcomp(numeric_data, center = TRUE, scale. = TRUE)
pca_scores <- as.data.frame(pca_result$x)
pca_meta <- cbind(pca_scores, metadata)

ggplot(pca_meta, aes(x = PC1, y = PC2, color = smoker, shape = on_art)) +
  geom_point() +
  labs(
    title = "PCA: PC3 vs PC4 ",
    x = "Principal Component 3",
    y = "Principal Component 4"
  ) +
  scale_color_manual(values = c(
    "Yes currently" = "red",
    "In the past" = "blue",
    "Never smoked" = "black"
  )) +
  theme_minimal()

ggplot(pca_meta, aes(x = PC3, y = PC4, color = smoker, shape = on_art)) +
  geom_point() +
  labs(
    title = "PCA: PC3 vs PC4 ",
    x = "Principal Component 3",
    y = "Principal Component 4"
  ) +
  scale_color_manual(values = c(
    "Yes currently" = "red",
    "In the past" = "blue",
    "Never smoked" = "black"
  )) +
  theme_minimal()

pca_result

sdev <- pca_result$sdev
variance <- sdev^2
variance_explained <- variance / sum(variance)
print(variance_explained)

screeplot_df <- data.frame(
  PC = seq_along(pca_result$sdev),
  Variance = pca_result$sdev^2,
  Proportion = pca_result$sdev^2 / sum(pca_result$sdev^2)
)

ggplot(screeplot_df, aes(x = PC, y = Proportion)) +
  geom_point() +
  geom_line() +
  labs(
    title = "Scree Plot",
    x = "Principal Component",
    y = "Proportion of Variance Explained"
  ) +
  theme_minimal()
```

### Smoking: significant retinal variables

```{r}
#| label: pca-smoking-sig-retinal
pca_dta <- dta |>
  select(smoker, on_art, all_of(sig_smoker_ret))

complete_rows <- complete.cases(pca_dta |>
  select(all_of(sig_smoker_ret)))

filtered_pca_dta <- pca_dta[complete_rows, ]

numeric_data <- filtered_pca_dta |>
  select(all_of(sig_smoker_ret))

metadata <- filtered_pca_dta |>
  select(on_art, smoker)

pca_result <- prcomp(numeric_data, center = TRUE, scale. = TRUE)
pca_scores <- as.data.frame(pca_result$x)
pca_meta <- cbind(pca_scores, metadata)

ggplot(pca_meta, aes(x = PC1, y = PC2, color = smoker, shape = on_art)) +
  geom_point() +
  labs(
    title = "PCA: PC3 vs PC4 ",
    x = "Principal Component 3",
    y = "Principal Component 4"
  ) +
  scale_color_manual(values = c(
    "Yes currently" = "red",
    "In the past" = "blue",
    "Never smoked" = "black"
  )) +
  theme_minimal()

ggplot(pca_meta, aes(x = PC3, y = PC4, color = smoker, shape = on_art)) +
  geom_point() +
  labs(
    title = "PCA: PC3 vs PC4 ",
    x = "Principal Component 3",
    y = "Principal Component 4"
  ) +
  scale_color_manual(values = c(
    "Yes currently" = "red",
    "In the past" = "blue",
    "Never smoked" = "black"
  )) +
  theme_minimal()

pca_result

sdev <- pca_result$sdev
variance <- sdev^2
variance_explained <- variance / sum(variance)
print(variance_explained)

screeplot_df <- data.frame(
  PC = seq_along(pca_result$sdev),
  Variance = pca_result$sdev^2,
  Proportion = pca_result$sdev^2 / sum(pca_result$sdev^2)
)

ggplot(screeplot_df, aes(x = PC, y = Proportion)) +
  geom_point() +
  geom_line() +
  labs(
    title = "Scree Plot",
    x = "Principal Component",
    y = "Proportion of Variance Explained"
  ) +
  theme_minimal()
```

### Viral load: significant variables

```{r}
#| label: pca-viral-load-sig
pca_dta <- dta |>
  select(viral_load_group, on_art, all_of(sig_viral_load))

complete_rows <- complete.cases(pca_dta |>
  select(all_of(sig_viral_load)))

filtered_pca_dta <- pca_dta[complete_rows, ]

numeric_data <- filtered_pca_dta |>
  select(all_of(sig_viral_load))

metadata <- filtered_pca_dta |>
  select(on_art, viral_load_group)

pca_result <- prcomp(numeric_data, center = TRUE, scale. = TRUE)
pca_scores <- as.data.frame(pca_result$x)
pca_meta <- cbind(pca_scores, metadata)

ggplot(
  pca_meta,
  aes(x = PC1, y = PC2, color = viral_load_group, shape = on_art)
) +
  geom_point() +
  labs(
    title = "PCA: PC3 vs PC4 ",
    x = "Principal Component 1",
    y = "Principal Component 2"
  ) +
  scale_color_manual(values = c(
    "Viral load > 1000" = "red",
    "Viral load < 1000" = "blue",
    "Never smoked" = "black"
  )) +
  theme_minimal()

ggplot(
  pca_meta,
  aes(x = PC3, y = PC4, color = viral_load_group, shape = on_art)
) +
  geom_point() +
  labs(
    title = "PCA: PC3 vs PC4 ",
    x = "Principal Component 3",
    y = "Principal Component 4"
  ) +
  scale_color_manual(values = c(
    "Viral load > 1000" = "red",
    "Viral load < 1000" = "blue",
    "Never smoked" = "black"
  )) +
  theme_minimal()

pca_result

sdev <- pca_result$sdev
variance <- sdev^2
variance_explained <- variance / sum(variance)
print(variance_explained)

screeplot_df <- data.frame(
  PC = seq_along(pca_result$sdev),
  Variance = pca_result$sdev^2,
  Proportion = pca_result$sdev^2 / sum(pca_result$sdev^2)
)

ggplot(screeplot_df, aes(x = PC, y = Proportion)) +
  geom_point() +
  geom_line() +
  labs(
    title = "Scree Plot",
    x = "Principal Component",
    y = "Proportion of Variance Explained"
  ) +
  theme_minimal()
```

### Viral load: significant retinal variables

```{r}
#| label: pca-viral-load-sig-retinal
pca_dta <- dta |>
  select(viral_load_group, on_art, all_of(sig_viral_load_ret))

complete_rows <- complete.cases(pca_dta |>
  select(all_of(sig_viral_load_ret)))

filtered_pca_dta <- pca_dta[complete_rows, ]

numeric_data <- filtered_pca_dta |>
  select(all_of(sig_viral_load_ret))

metadata <- filtered_pca_dta |>
  select(on_art, viral_load_group)

pca_result <- prcomp(numeric_data, center = TRUE, scale. = TRUE)
pca_scores <- as.data.frame(pca_result$x)
pca_meta <- cbind(pca_scores, metadata)

ggplot(
  pca_meta,
  aes(x = PC1, y = PC2, color = viral_load_group, shape = on_art)
) +
  geom_point() +
  labs(
    title = "PCA: PC3 vs PC4 ",
    x = "Principal Component 1",
    y = "Principal Component 2"
  ) +
  scale_color_manual(values = c(
    "Viral load > 1000" = "red",
    "Viral load < 1000" = "blue",
    "Never smoked" = "black"
  )) +
  theme_minimal()

ggplot(
  pca_meta,
  aes(x = PC3, y = PC4, color = viral_load_group, shape = on_art)
) +
  geom_point() +
  labs(
    title = "PCA: PC3 vs PC4 ",
    x = "Principal Component 3",
    y = "Principal Component 4"
  ) +
  scale_color_manual(values = c(
    "Viral load > 1000" = "red",
    "Viral load < 1000" = "blue",
    "Never smoked" = "black"
  )) +
  theme_minimal()

pca_result

sdev <- pca_result$sdev
variance <- sdev^2
variance_explained <- variance / sum(variance)
print(variance_explained)

screeplot_df <- data.frame(
  PC = seq_along(pca_result$sdev),
  Variance = pca_result$sdev^2,
  Proportion = pca_result$sdev^2 / sum(pca_result$sdev^2)
)

ggplot(screeplot_df, aes(x = PC, y = Proportion)) +
  geom_point() +
  geom_line() +
  labs(
    title = "Scree Plot",
    x = "Principal Component",
    y = "Proportion of Variance Explained"
  ) +
  theme_minimal()

t_test_result <- t.test(
  PC1 ~ viral_load_group,
  data = pca_meta,
  var.equal = TRUE
)

print(t_test_result)

wilcox_test_result <- wilcox.test(PC1 ~ viral_load_group, data = pca_meta)

print(wilcox_test_result)
```

:::

## Linear regression

One linear model per retinal measure, with HIV status, ART status or another HIV-related variable as the predictor, adjusted for age, sex, BMI, smoking and (in most models) CVD risk factors.

::: panel-tabset

### HIV status

```{r}
#| label: lm-hiv
# One linear model per retinal measure: HIV status plus covariates
fit_model <- function(var) {
  formula <- as.formula(paste(
    var, "~ hiv_pos + age + gender + cvd_risk_factor + bmi + smoker"
  ))
  model <- lm(formula, data = dta_nums)

  tidy(model) |>
    mutate(
      Adj_R_squared = summary(model)$adj.r.squared,
      Model = var
    )
}

dta_nums <- dta |>
  select(
    study_id, all_of(retinal_var), hiv_pos, age, gender, cvd_risk_factor, bmi,
    smoker
  )

linear_results_table <- retinal_var |>
  map_dfr(fit_model)

results_df <- linear_results_table |>
  filter(term != "(Intercept)") |>
  select(Model, term, estimate, Adj_R_squared, std.error, statistic, p.value) |>
  arrange(p.value)

datatable(results_df) |>
  formatStyle(
    columns = c(
      "Model", "term", "Adj_R_squared", "estimate", "std.error", "statistic",
      "p.value"
    ),
    fontSize = "11px",
    color = "black",
    backgroundColor = "#FFFFFF"
  ) |>
  formatRound(
    columns = c(
      "Adj_R_squared", "estimate", "std.error", "statistic", "p.value"
    ),
    digits = 4
  )
```

### HIV status: plots

```{r}
#| label: lm-hiv-plots
# Observed vs predicted values from the adjusted model
plot_regression_line <- function(data, vars, angle = 45) {
  data <- dta |>
    select(all_of(c(
      vars, "hiv_pos", "age", "gender", "cvd_risk_factor", "bmi", "smoker"
    ))) |>
    drop_na()

  formula <- as.formula(paste(
    vars, "~ hiv_pos + age + gender + cvd_risk_factor + bmi + smoker"
  ))

  model <- lm(formula, data = dta)

  data$predicted <- predict(model, newdata = data)

  p <- ggplot(data, aes_string(x = vars, y = "predicted")) +
    geom_point(aes(color = "Data Points")) +
    geom_smooth(method = "lm", se = FALSE, color = "blue") +
    labs(
      title = paste("Regression Line for", vars),
      x = vars,
      y = "Predicted Values"
    ) +
    theme(axis.text.x = element_text(angle = angle, hjust = 1))

  return(p)
}

for (var in retinal_var) {
  plot_regression_line(data = dta_nums, vars = var, angle = 45) |>
    print()
}
```

### ART status

```{r}
#| label: lm-art
# One linear model per retinal measure: ART status plus covariates
# (rows with missing on_art, i.e. HIV-negative participants, are dropped)
dta_nums <- dta |>
  select(
    study_id, all_of(retinal_var), on_art, age, gender, cvd_risk_factor, bmi,
    smoker
  )

fit_model <- function(var) {
  formula <- as.formula(paste(
    var, "~ on_art + age + gender + cvd_risk_factor + bmi + smoker"
  ))
  model <- lm(formula, data = dta_nums)

  tidy(model) |>
    mutate(
      Adj_R_squared = summary(model)$adj.r.squared,
      Model = var
    )
}

dta_nums <- dta |>
  select(
    study_id, all_of(retinal_var), on_art, age, gender, cvd_risk_factor, bmi,
    smoker
  ) |>
  drop_na()

linear_results_table <- retinal_var |>
  map_dfr(fit_model)

results_df <- linear_results_table |>
  filter(term != "(Intercept)") |>
  select(Model, term, estimate, Adj_R_squared, std.error, statistic, p.value) |>
  arrange(Model)

datatable(results_df) |>
  formatStyle(
    columns = c(
      "Model", "term", "Adj_R_squared", "estimate", "std.error", "statistic",
      "p.value"
    ),
    fontSize = "11px",
    color = "black",
    backgroundColor = "#FFFFFF"
  ) |>
  formatRound(
    columns = c(
      "Adj_R_squared", "estimate", "std.error", "statistic", "p.value"
    ),
    digits = 4
  )
```

### ART status and viral load group

```{r}
#| label: lm-art-viral-load
dta_nums <- dta |>
  select(
    study_id, all_of(retinal_var), on_art, age, gender, bmi, smoker,
    viral_load_group
  )

fit_model <- function(var) {
  formula <- as.formula(paste(
    var, "~ on_art + age + gender +  bmi + smoker + viral_load_group"
  ))
  model <- lm(formula, data = dta_nums)

  tidy(model) |>
    mutate(
      Adj_R_squared = summary(model)$adj.r.squared,
      Model = var
    )
}

dta_nums <- dta |>
  select(
    study_id, all_of(retinal_var), on_art, age, gender, bmi, smoker,
    viral_load_group
  ) |>
  drop_na()

linear_results_table <- retinal_var |>
  map_dfr(fit_model)

results_df <- linear_results_table |>
  filter(term != "(Intercept)") |>
  select(Model, term, estimate, Adj_R_squared, std.error, statistic, p.value) |>
  arrange(Model)

datatable(results_df) |>
  formatStyle(
    columns = c(
      "Model", "term", "Adj_R_squared", "estimate", "std.error", "statistic",
      "p.value"
    ),
    fontSize = "11px",
    color = "black",
    backgroundColor = "#FFFFFF"
  ) |>
  formatRound(
    columns = c(
      "Adj_R_squared", "estimate", "std.error", "statistic", "p.value"
    ),
    digits = 4
  )
```

### ART status and CD4 count group

```{r}
#| label: lm-art-cd4
dta_nums <- dta |>
  select(
    study_id, all_of(retinal_var), on_art, age, gender, bmi, smoker,
    cd4_count_endo_group
  )

fit_model <- function(var) {
  formula <- as.formula(paste(
    var, "~ on_art + age + gender +  bmi + smoker + cd4_count_endo_group_2"
  ))
  model <- lm(formula, data = dta_nums)

  tidy(model) |>
    mutate(
      Adj_R_squared = summary(model)$adj.r.squared,
      Model = var
    )
}

dta_nums <- dta |>
  select(
    study_id, all_of(retinal_var), on_art, age, gender, bmi, smoker,
    cd4_count_endo_group
  ) |>
  drop_na()

linear_results_table <- retinal_var |>
  map_dfr(fit_model)

results_df <- linear_results_table |>
  filter(term != "(Intercept)") |>
  select(Model, term, estimate, Adj_R_squared, std.error, statistic, p.value) |>
  arrange(Model)

datatable(results_df) |>
  formatStyle(
    columns = c(
      "Model", "term", "Adj_R_squared", "estimate", "std.error", "statistic",
      "p.value"
    ),
    fontSize = "11px",
    color = "black",
    backgroundColor = "#FFFFFF"
  ) |>
  formatRound(
    columns = c(
      "Adj_R_squared", "estimate", "std.error", "statistic", "p.value"
    ),
    digits = 4
  )
```

### ART status: plots

```{r}
#| label: lm-art-plots
# Observed vs predicted values from the adjusted model
dta_nums <- dta |>
  select(
    study_id, all_of(retinal_var), on_art, age, gender, cvd_risk_factor, bmi,
    smoker
  )

plot_regression_line <- function(data, vars, angle = 45) {
  data <- dta |>
    select(all_of(c(
      vars, "on_art", "age", "gender", "cvd_risk_factor", "bmi", "smoker"
    ))) |>
    drop_na()

  formula <- as.formula(paste(
    vars, "~ on_art + age + gender + cvd_risk_factor + bmi + smoker"
  ))

  model <- lm(formula, data = dta)

  data$predicted <- predict(model, newdata = data)

  p <- ggplot(data, aes_string(x = vars, y = "predicted")) +
    geom_point(aes(color = "Data Points")) +
    geom_smooth(method = "lm", se = FALSE, color = "blue") +
    labs(
      title = paste("Regression Line for", vars),
      x = vars,
      y = "Predicted Values"
    ) +
    theme(axis.text.x = element_text(angle = angle, hjust = 1))

  return(p)
}

for (var in retinal_var) {
  plot_regression_line(data = dta_nums, vars = var, angle = 45) |>
    print()
}
```

### CD4 count (clinic)

```{r}
#| label: lm-cd4-clinic
fit_model <- function(var) {
  formula <- as.formula(paste(
    var, "~ cd4_count_clinic +  age + gender + cvd_risk_factor + bmi + smoker"
  ))
  model <- lm(formula, data = dta_nums)

  tidy(model) |>
    mutate(
      Adj_R_squared = summary(model)$adj.r.squared,
      Model = var
    )
}

dta_nums <- dta |>
  select(
    study_id, all_of(retinal_var), cd4_count_clinic, age, gender,
    cvd_risk_factor, bmi, smoker
  ) |>
  drop_na()

linear_results_table <- retinal_var |>
  map_dfr(fit_model)

results_df <- linear_results_table |>
  filter(term != "(Intercept)") |>
  select(Model, term, estimate, Adj_R_squared, std.error, statistic, p.value) |>
  arrange(Model)

datatable(results_df) |>
  formatStyle(
    columns = c(
      "Model", "term", "Adj_R_squared", "estimate", "std.error", "statistic",
      "p.value"
    ),
    fontSize = "11px",
    color = "black",
    backgroundColor = "#FFFFFF"
  ) |>
  formatRound(
    columns = c(
      "Adj_R_squared", "estimate", "std.error", "statistic", "p.value"
    ),
    digits = 4
  )
```

### CD4 count (EndoAfrica)

```{r}
#| label: lm-cd4-endo
fit_model <- function(var) {
  formula <- as.formula(paste(
    var, "~ endo_result +  age + gender + cvd_risk_factor + bmi + smoker"
  ))
  model <- lm(formula, data = dta_nums)

  tidy(model) |>
    mutate(
      Adj_R_squared = summary(model)$adj.r.squared,
      Model = var
    )
}

dta_nums <- dta |>
  select(
    study_id, all_of(retinal_var), endo_result, age, gender, cvd_risk_factor,
    bmi, smoker
  ) |>
  drop_na()

linear_results_table <- retinal_var |>
  map_dfr(fit_model)

results_df <- linear_results_table |>
  filter(term != "(Intercept)") |>
  select(Model, term, estimate, Adj_R_squared, std.error, statistic, p.value) |>
  arrange(Model)

datatable(results_df) |>
  formatStyle(
    columns = c(
      "Model", "term", "Adj_R_squared", "estimate", "std.error", "statistic",
      "p.value"
    ),
    fontSize = "11px",
    color = "black",
    backgroundColor = "#FFFFFF"
  ) |>
  formatRound(
    columns = c(
      "Adj_R_squared", "estimate", "std.error", "statistic", "p.value"
    ),
    digits = 4
  )
```

### ART duration group

```{r}
#| label: lm-art-duration
fit_model <- function(var) {
  formula <- as.formula(paste(
    var, "~ art_duration_group + age + gender + cvd_risk_factor + bmi + smoker"
  ))
  model <- lm(formula, data = dta_nums)

  tidy(model) |>
    mutate(
      Adj_R_squared = summary(model)$adj.r.squared,
      Model = var
    )
}

dta_nums <- dta |>
  select(
    study_id, all_of(retinal_var), art_duration_group, age, gender,
    cvd_risk_factor, bmi, smoker
  ) |>
  drop_na()

linear_results_table <- retinal_var |>
  map_dfr(fit_model)

results_df <- linear_results_table |>
  filter(term != "(Intercept)") |>
  select(Model, term, estimate, Adj_R_squared, std.error, statistic, p.value) |>
  arrange(Model)

datatable(results_df) |>
  formatStyle(
    columns = c(
      "Model", "term", "Adj_R_squared", "estimate", "std.error", "statistic",
      "p.value"
    ),
    fontSize = "11px",
    color = "black",
    backgroundColor = "#FFFFFF"
  ) |>
  formatRound(
    columns = c(
      "Adj_R_squared", "estimate", "std.error", "statistic", "p.value"
    ),
    digits = 4
  )
```

### ART duration: plots

```{r}
#| label: lm-art-duration-plots
# Observed vs predicted values from the adjusted model
plot_regression_line <- function(data, vars, angle = 45) {
  data <- dta |>
    select(all_of(c(
      vars, "art_duration_weeks", "age", "gender", "cvd_risk_factor", "bmi",
      "smoker"
    ))) |>
    drop_na()

  formula <- as.formula(paste(
    vars,
    "~ art_duration_weeks + age + gender + cvd_risk_factor + bmi + smoker"
  ))

  model <- lm(formula, data = dta)

  data$predicted <- predict(model, newdata = data)

  p <- ggplot(data, aes_string(x = vars, y = "predicted")) +
    geom_point(aes(color = "Data Points")) +
    geom_smooth(method = "lm", se = FALSE, color = "blue") +
    labs(
      title = paste("Regression Line for", vars),
      x = vars,
      y = "Predicted Values"
    ) +
    theme(axis.text.x = element_text(angle = angle, hjust = 1))

  return(p)
}

for (var in retinal_var) {
  plot_regression_line(data = dta_nums, vars = var, angle = 45) |>
    print()
}
```

### Viral load group

```{r}
#| label: lm-viral-load
fit_model <- function(var) {
  formula <- as.formula(paste(
    var, "~ viral_load + age + gender + cvd_risk_factor + bmi + smoker"
  ))
  model <- lm(formula, data = dta_nums)

  tidy(model) |>
    mutate(
      Adj_R_squared = summary(model)$adj.r.squared,
      Model = var
    )
}

dta_nums <- dta |>
  select(
    study_id, all_of(retinal_var), viral_load, age, gender, cvd_risk_factor,
    bmi, smoker
  ) |>
  drop_na()

linear_results_table <- retinal_var |>
  map_dfr(fit_model)

results_df <- linear_results_table |>
  filter(term != "(Intercept)") |>
  select(Model, term, estimate, Adj_R_squared, std.error, statistic, p.value) |>
  arrange(Model)

datatable(results_df) |>
  formatStyle(
    columns = c(
      "Model", "term", "Adj_R_squared", "estimate", "std.error", "statistic",
      "p.value"
    ),
    fontSize = "11px",
    color = "black",
    backgroundColor = "#FFFFFF"
  ) |>
  formatRound(
    columns = c(
      "Adj_R_squared", "estimate", "std.error", "statistic", "p.value"
    ),
    digits = 4
  )
```

:::

## Random forest and decision trees (exploratory)

Exploratory classification of HIV and ART status from the retinal measures. These models were not part of the final report.

::: panel-tabset

### Initial model

```{r}
#| label: rf-initial
dta_nums <- dta |>
  select(study_id, hiv_pos, all_of(retinal_var)) |>
  drop_na() |>
  mutate(
    hiv_pos = as.factor(hiv_pos)
  )

rf_model <- randomForest(
  hiv_pos ~ crae + crve + crae_crve_ratio + segmented_fractal_analysis +
    skeleton_fractal_analysis + average_arterial_tree_diameter +
    arterial_tree_segments + branch_segment_tortuosity_skewness +
    branch_segment_tortuosity_kurtosis + branch_segment_length_average +
    arteriole_branch_parent_width + arteriole_branch_smaller_branch +
    arteriole_branch_larger_branch + arteriole_branch_angle +
    arteriole_branch_optimality + arteriole_mean_asymmetry_ratio +
    arteriole_mean_angular_asymmetry + venule_branch_parent_width +
    venule_branch_smaller_branch + venule_branch_larger_branch +
    venule_branch_angle + venule_branch_optimality +
    venule_mean_branching_coefficient + venule_mean_asymmetry_ratio +
    venule_mean_angular_asymmetry,
  data = dta_nums,
  proximity = TRUE
)

rf_model

# Print confusion matrix
print(rf_model$confusion)

# Print variable importance
print(importance(rf_model))

# Results from one run (no seed is set, so these numbers vary between runs):
# OOB error rate 44.84%, i.e. about 45% of cases misclassified
# true negatives = 24, false positives = 89
# false negatives = 37, true positives = 131
# 79% of actual "No" and 22% of actual "Yes" cases were misclassified
# Variables with higher MeanDecreaseGini are more influential in predicting
# hiv_pos
```

### Error rate by number of trees

```{r}
#| label: rf-error-by-trees
num_trees <- 1:nrow(rf_model$err.rate)

oob_error_1 <- data.frame(
  trees = rep(num_trees, times = 3),
  type = rep(c("OOB", "Yes", "No"), each = length(num_trees)),
  Error = c(
    rf_model$err.rate[, "OOB"], # Out-Of-Bag error
    rf_model$err.rate[, "Yes"], # Error rate for "Yes" class
    rf_model$err.rate[, "No"] # Error rate for "No" class
  )
)

print(oob_error_1)

# Error rates as trees are added to the forest
ggplot(data = oob_error_1, aes(x = trees, y = Error)) +
  geom_line(aes(color = type))
```

### Change number of trees (500 to 1000)

```{r}
#| label: rf-more-trees
rf_model_2 <- randomForest(
  hiv_pos ~ crae + crve + crae_crve_ratio + segmented_fractal_analysis +
    skeleton_fractal_analysis + average_arterial_tree_diameter +
    arterial_tree_segments + branch_segment_tortuosity_skewness +
    branch_segment_tortuosity_kurtosis + branch_segment_length_average +
    arteriole_branch_parent_width + arteriole_branch_smaller_branch +
    arteriole_branch_larger_branch + arteriole_branch_angle +
    arteriole_branch_optimality + arteriole_mean_asymmetry_ratio +
    arteriole_mean_angular_asymmetry + venule_branch_parent_width +
    venule_branch_smaller_branch + venule_branch_larger_branch +
    venule_branch_angle + venule_branch_optimality +
    venule_mean_branching_coefficient + venule_mean_asymmetry_ratio +
    venule_mean_angular_asymmetry,
  data = dta_nums,
  ntrees = 1000,
  proximity = TRUE
)

num_trees <- 1:nrow(rf_model_2$err.rate)

oob_error_1 <- data.frame(
  trees = rep(num_trees, times = 3),
  type = rep(c("OOB", "Yes", "No"), each = length(num_trees)),
  Error = c(
    rf_model$err.rate[, "OOB"], # Out-Of-Bag error
    rf_model$err.rate[, "Yes"], # Error rate for "Yes" class
    rf_model$err.rate[, "No"] # Error rate for "No" class
  )
)
ggplot(data = oob_error_1, aes(x = trees, y = Error)) +
  geom_line(aes(color = type))
```

### Check number of variables used

```{r}
#| label: rf-check-variables
oob_values <- vector(length = 10)
for (i in 1:10) {
  tmp <- randomForest(
    hiv_pos ~ crae + crve + crae_crve_ratio + segmented_fractal_analysis +
      skeleton_fractal_analysis + average_arterial_tree_diameter +
      arterial_tree_segments + branch_segment_tortuosity_skewness +
      branch_segment_tortuosity_kurtosis + branch_segment_length_average +
      arteriole_branch_parent_width + arteriole_branch_smaller_branch +
      arteriole_branch_larger_branch + arteriole_branch_angle +
      arteriole_branch_optimality + arteriole_mean_asymmetry_ratio +
      arteriole_mean_angular_asymmetry + venule_branch_parent_width +
      venule_branch_smaller_branch + venule_branch_larger_branch +
      venule_branch_angle + venule_branch_optimality +
      venule_mean_branching_coefficient + venule_mean_asymmetry_ratio +
      venule_mean_angular_asymmetry,
    data = dta_nums,
    proximity = TRUE
  )
  oob_values[i] <- tmp$err.rate[nrow(tmp$err.rate), 1]
}

# err.rate has one row per tree; the last row of the first column is the
# final OOB error rate for each of the 10 fits
oob_values
```

### HIV status: forest and decision tree

```{r}
#| label: rf-hiv
dta_nums <- dta |>
  select(study_id, hiv_pos, all_of(retinal_var)) |>
  drop_na() |>
  mutate(
    hiv_pos = as.factor(hiv_pos)
  )

importance_list <- list()

rf_model <- randomForest(
  hiv_pos ~ crae + crve + crae_crve_ratio + segmented_fractal_analysis +
    skeleton_fractal_analysis + average_arterial_tree_diameter +
    arterial_tree_segments + branch_segment_tortuosity_skewness +
    branch_segment_tortuosity_kurtosis + branch_segment_length_average +
    arteriole_branch_parent_width + arteriole_branch_smaller_branch +
    arteriole_branch_larger_branch + arteriole_branch_angle +
    arteriole_branch_optimality + arteriole_mean_asymmetry_ratio +
    arteriole_mean_angular_asymmetry + venule_branch_parent_width +
    venule_branch_smaller_branch + venule_branch_larger_branch +
    venule_branch_angle + venule_branch_optimality +
    venule_mean_branching_coefficient + venule_mean_asymmetry_ratio +
    venule_mean_angular_asymmetry,
  data = dta_nums,
  proximity = TRUE
)

importance_values <- importance(rf_model)
importance_df <- as.data.frame(importance_values)
importance_df$Feature <- rownames(importance_df)
importance_df <- importance_df |>
  rename(MeanDecreaseGini = MeanDecreaseGini)
importance_df <- importance_df |>
  arrange(desc(MeanDecreaseGini))
print(importance_df)
p <- ggplot(
  importance_df,
  aes(x = reorder(Feature, MeanDecreaseGini), y = MeanDecreaseGini)
) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(
    title = "Feature Importance",
    x = "Features",
    y = "Importance (MeanDecreaseGini)"
  ) +
  theme_minimal()

print(p)

tree_model <- rpart(
  hiv_pos ~ crae + crve + crae_crve_ratio + segmented_fractal_analysis +
    skeleton_fractal_analysis + average_arterial_tree_diameter +
    arterial_tree_segments + branch_segment_tortuosity_skewness +
    branch_segment_tortuosity_kurtosis + branch_segment_length_average +
    arteriole_branch_parent_width + arteriole_branch_smaller_branch +
    arteriole_branch_larger_branch + arteriole_branch_angle +
    arteriole_branch_optimality + arteriole_mean_asymmetry_ratio +
    arteriole_mean_angular_asymmetry + venule_branch_parent_width +
    venule_branch_smaller_branch + venule_branch_larger_branch +
    venule_branch_angle + venule_branch_optimality +
    venule_mean_branching_coefficient + venule_mean_asymmetry_ratio +
    venule_mean_angular_asymmetry,
  data = dta_nums,
  method = "class"
)

rpart.plot(tree_model)
```

### ART status: forest and decision tree

```{r}
#| label: rf-art
dta_nums <- dta |>
  select(study_id, on_art, all_of(retinal_var)) |>
  drop_na() |>
  mutate(
    on_art = as.factor(on_art)
  )

importance_list <- list()

rf_model <- randomForest(
  on_art ~ crae + crve + crae_crve_ratio + segmented_fractal_analysis +
    skeleton_fractal_analysis + average_arterial_tree_diameter +
    arterial_tree_segments + branch_segment_tortuosity_skewness +
    branch_segment_tortuosity_kurtosis + branch_segment_length_average +
    arteriole_branch_parent_width + arteriole_branch_smaller_branch +
    arteriole_branch_larger_branch + arteriole_branch_angle +
    arteriole_branch_optimality + arteriole_mean_asymmetry_ratio +
    arteriole_mean_angular_asymmetry + venule_branch_parent_width +
    venule_branch_smaller_branch + venule_branch_larger_branch +
    venule_branch_angle + venule_branch_optimality +
    venule_mean_branching_coefficient + venule_mean_asymmetry_ratio +
    venule_mean_angular_asymmetry,
  data = dta_nums,
  proximity = TRUE
)

importance_values <- importance(rf_model)
importance_df <- as.data.frame(importance_values)
importance_df$Feature <- rownames(importance_df)
importance_df <- importance_df |>
  rename(MeanDecreaseGini = MeanDecreaseGini)
importance_df <- importance_df |>
  arrange(desc(MeanDecreaseGini))
print(importance_df)
p <- ggplot(
  importance_df,
  aes(x = reorder(Feature, MeanDecreaseGini), y = MeanDecreaseGini)
) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(
    title = "Feature Importance",
    x = "Features",
    y = "Importance (MeanDecreaseGini)"
  ) +
  theme_minimal()

print(p)

tree_model_2 <- rpart(
  on_art ~ crae + crve + crae_crve_ratio + segmented_fractal_analysis +
    skeleton_fractal_analysis + average_arterial_tree_diameter +
    arterial_tree_segments + branch_segment_tortuosity_skewness +
    branch_segment_tortuosity_kurtosis + branch_segment_length_average +
    arteriole_branch_parent_width + arteriole_branch_smaller_branch +
    arteriole_branch_larger_branch + arteriole_branch_angle +
    arteriole_branch_optimality + arteriole_mean_asymmetry_ratio +
    arteriole_mean_angular_asymmetry + venule_branch_parent_width +
    venule_branch_smaller_branch + venule_branch_larger_branch +
    venule_branch_angle + venule_branch_optimality +
    venule_mean_branching_coefficient + venule_mean_asymmetry_ratio +
    venule_mean_angular_asymmetry,
  data = dta_nums,
  method = "class"
)

rpart.plot(tree_model_2)
```

:::

## Random forest with a train/test split (exploratory)

Exploratory check of prediction accuracy on a held-out 40% of participants. Not part of the final report.

::: panel-tabset

### HIV status: retinal variables

```{r}
#| label: rf-split-hiv-retinal
dta_nums <- dta |>
  select(study_id, hiv_pos, cvd_risk_factor, all_of(retinal_var)) |>
  drop_na() |>
  mutate(
    hiv_pos = as.factor(hiv_pos)
  )

set.seed(123)

trainIndex <- createDataPartition(dta_nums$hiv_pos, p = 0.6, list = FALSE)
trainData <- dta_nums[trainIndex, ]
testData <- dta_nums[-trainIndex, ]

rf_model <- randomForest(
  hiv_pos ~ crae + crve + crae_crve_ratio + segmented_fractal_analysis +
    skeleton_fractal_analysis + average_arterial_tree_diameter +
    arterial_tree_segments + branch_segment_tortuosity_skewness +
    branch_segment_tortuosity_kurtosis + branch_segment_length_average +
    arteriole_branch_parent_width + arteriole_branch_smaller_branch +
    arteriole_branch_larger_branch + arteriole_branch_angle +
    arteriole_branch_optimality + arteriole_mean_asymmetry_ratio +
    arteriole_mean_angular_asymmetry + venule_branch_parent_width +
    venule_branch_smaller_branch + venule_branch_larger_branch +
    venule_branch_angle + venule_branch_optimality +
    venule_mean_branching_coefficient + venule_mean_asymmetry_ratio +
    venule_mean_angular_asymmetry,
  data = trainData,
  proximity = TRUE
)

predictions <- predict(rf_model, testData)

postResample(predictions, testData$hiv_pos)
```

### HIV status: retinal variables and CVD risk factor

```{r}
#| label: rf-split-hiv-retinal-cvd
dta_nums <- dta |>
  select(study_id, hiv_pos, cvd_risk_factor, all_of(retinal_var)) |>
  drop_na() |>
  mutate(
    hiv_pos = as.factor(hiv_pos)
  )

set.seed(123)

trainIndex <- createDataPartition(dta_nums$hiv_pos, p = 0.6, list = FALSE)
trainData <- dta_nums[trainIndex, ]
testData <- dta_nums[-trainIndex, ]

rf_model <- randomForest(
  hiv_pos ~ cvd_risk_factor + crae + crve + crae_crve_ratio +
    segmented_fractal_analysis + skeleton_fractal_analysis +
    average_arterial_tree_diameter + arterial_tree_segments +
    branch_segment_tortuosity_skewness + branch_segment_tortuosity_kurtosis +
    branch_segment_length_average + arteriole_branch_parent_width +
    arteriole_branch_smaller_branch + arteriole_branch_larger_branch +
    arteriole_branch_angle + arteriole_branch_optimality +
    arteriole_mean_asymmetry_ratio + arteriole_mean_angular_asymmetry +
    venule_branch_parent_width + venule_branch_smaller_branch +
    venule_branch_larger_branch + venule_branch_angle +
    venule_branch_optimality + venule_mean_branching_coefficient +
    venule_mean_asymmetry_ratio + venule_mean_angular_asymmetry,
  data = trainData,
  proximity = TRUE
)

predictions <- predict(rf_model, testData)

postResample(predictions, testData$hiv_pos)
```

### ART status: retinal variables

```{r}
#| label: rf-split-art-retinal
dta_nums <- dta |>
  select(study_id, on_art, all_of(retinal_var)) |>
  drop_na() |>
  mutate(
    on_art = as.factor(on_art)
  )

set.seed(123)

trainIndex <- createDataPartition(dta_nums$on_art, p = 0.6, list = FALSE)
trainData <- dta_nums[trainIndex, ]
testData <- dta_nums[-trainIndex, ]

rf_model <- randomForest(
  on_art ~ crae + crve + crae_crve_ratio + segmented_fractal_analysis +
    skeleton_fractal_analysis + average_arterial_tree_diameter +
    arterial_tree_segments + branch_segment_tortuosity_skewness +
    branch_segment_tortuosity_kurtosis + branch_segment_length_average +
    arteriole_branch_parent_width + arteriole_branch_smaller_branch +
    arteriole_branch_larger_branch + arteriole_branch_angle +
    arteriole_branch_optimality + arteriole_mean_asymmetry_ratio +
    arteriole_mean_angular_asymmetry + venule_branch_parent_width +
    venule_branch_smaller_branch + venule_branch_larger_branch +
    venule_branch_angle + venule_branch_optimality +
    venule_mean_branching_coefficient + venule_mean_asymmetry_ratio +
    venule_mean_angular_asymmetry,
  data = trainData,
  proximity = TRUE
)

predictions <- predict(rf_model, testData)

postResample(predictions, testData$on_art)

# Note: accuracy here mostly reflects class imbalance (far more participants
# are on ART than ART-naive). The imbalance is not corrected for.
```

### HIV status: significant variables and covariates

```{r}
#| label: rf-split-hiv-sig
dta_nums <- dta |>
  select(
    study_id, hiv_pos, age, gender, bmi, cvd_risk_factor, crve,
    venule_branch_larger_branch, venule_branch_parent_width,
    skeleton_fractal_analysis
  ) |>
  drop_na() |>
  mutate(
    hiv_pos = as.factor(hiv_pos)
  )

set.seed(123)

trainIndex <- createDataPartition(dta_nums$hiv_pos, p = 0.6, list = FALSE)
trainData <- dta_nums[trainIndex, ]
testData <- dta_nums[-trainIndex, ]

rf_model <- randomForest(
  hiv_pos ~ age + gender + bmi + cvd_risk_factor + crve +
    venule_branch_larger_branch + venule_branch_parent_width +
    skeleton_fractal_analysis,
  data = trainData,
  proximity = TRUE
)

predictions <- predict(rf_model, testData)

postResample(predictions, testData$hiv_pos)
```

:::

## Session information

R and package versions from the last render.

```{r}
#| label: session-info
sessionInfo()
```
